### 2017-08-31 paper.tex
Initial rough draft commit

\documentclass[sigconf]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage{multicol}
\usepackage{graphicx}

% Dejavu fonts have unicode character support
% \usepackage{DejaVuSansMono}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{listings}

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\settopmatter{printacmref=false}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[LLVM in HPC'17]{LLVM in HPC: Fourth Workshop on the LLVM Compiler Infrastructure in HPC}{November 2017}{Denver Colorado, USA} 
\acmYear{2017}
\copyrightyear{2017}

%\acmPrice{15.00}

\begin{document}

\title{OpenMPIR}
\subtitle{Implementing OpenMP tasks with Tapir}
\author{George Stelle}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{stelleg@lanl.gov}

\author{Stephen Olivier}
\affiliation{\institution{Center for Computing Research \\ Sandia National Laboratories}}
\email{slolivi@sandia.gov}

\author{Pat McCormick}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{pat@lanl.gov}

\begin{abstract}
Optimizing compilers for task-level parallelism are still in their infancy.
This work implements a frontend which compiles a subset of OpenMP tasks to
Tapir IR. This enables analyses and optimizations previously inaccessible to
OpenMP codes. Initial performance results for the Barcelona OpenMP tasking
suite and show performance improvements over existing OpenMP compiler
implementations. 
\end{abstract}

\maketitle

\section{Introduction}

When writing task-parallel programs today, one has a large selection of
programming models. Unfortunately, despite having significant overlap in
semantics, parallel programming models like OpenMP, Cilk, Kokkos, HPX, Charm++,
Qthreads, pthreads, MPI, Chapel, UPC, etc. have virtually no ability to
interoperate, both at compile time and run time. In addition, many of these
programming models are implemented as libraries, and therefore have no compiler
support. There has been some recent work on specializing compilers to reason
about libraries \cite{kokkos}, but the fragmentation among the models that
presents another challenge for compiler writers, which must choose a single
model to analyze and optimize. An ideal solution would be some some form of
common representation of parallelism that all IRs can share.

Internal representations (IRs) are the compiler objects on which almost all
analyses and optimizations are performed. Historically, a major drawback for
parallel codes is that IRs don't have the ability to represent, and therefore
reason about, parallelism. This prevents the application of common
optimizations to concurrent code. The recent work of Schardl et al.
\cite{shardl2017} has shown that standard compiler IRs can be extended to
represent fork-join parallelism. In that work, by extending the LLVM
instruction set with three instructions, they are able to capture all of the
semantics to implement the semantics of Cilk, an extension to C. They suggested
that a similar approach could be taken with OpenMP tasks.

In this work, we put that suggestion to the test, implementing OpenMP tasks by
compiling them to Tapir instructions. This enables us to target non-openmp
runtimes and compare performance to existing OpenMP implementations. Specifically,
we run our prototype implementation on the Barcelona OpenMP task suite
\cite{barcelona}. We discuss the kinds of optimizations this enables, and how
the work could be extended to include other parallelism constructs and semantics
in OpenMP. We also discuss the possibility of extending this work to other
programming models, which would help to alleviate some of the fragmentation
issues discussed above.

The contributions of this work include:

\begin{itemize}
\item A prototype frontend from OpenMP task constructs to Tapir IR
\item Benchmarks of the implementation, showing performance improvements over
existing OpenMP implementations
\item An initial investigation into the reasons for the performance improvement
\end{itemize}

Our hope for this paper is that it moves the community towards a shared IR that
can be used for many programming models. This work presented in this paper only
represents a small step in that direction, and there is still a herculean amount
of work required. We argue that the benefits vastly outweigh the costs, and
hope to see more work follow suite.

\subsection{Outline}

The remainder of the paper is organized as follows: In Section~\ref{Sec:background} 
we give background and motivation for the paper. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including what implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. Finally, we discuss how the many ways the work can
be extended in Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

\section{Background} \label{Sec:Background}

Internal representations are a crucial part of modern compilers. By representing
code in a generic, hardware-agnostic format, they allow both multiple frontends for
different source languages, and multiple backends for different hardware, to all
take advantage of a series of compiler optimizations and analyses.

LLVM is arguably the most successful IR to date \cite{llvm}. By implementing a simple
single static assignment (SSA) representation, it enables powerful optimizations 
that are effective for a wide range of programming languages and hardware
\cite{llvm imps}.

Historically, LLVM has had a sequential semantics: there is no notion of
concurrent or parallel processes. This resulted in existing compilers treating 
parallelism constructs as thin wrappers for calls into runtime systems. These
calls are essentially impossible for the compiler to reason about, due to the
complexity of runtimes being called.

This unfortunate situation was recently remedied by Schardl et al.
\cite{tapir}. By adding three instructions to LLVM, Tapir enables analyses and
optimizations of parallel code. A full description

\subsection{Fragmentation}

In this section we discuss a significant challenge in HPC: fragmentation of
parallel programming models. There are many, many, many ways to write
parallel programs \cite{....}. From libraries, to language extensions, to
standalone languages, to combinations of the above, it can be hard to choose
the right one. This means that when writing tooling, optimizations, or
analyses, one has to choose which model to target. One common difficulty among
all of these is that reasoning about parallel programs is hard \cite{}, and
compilers can help \cite{}. Therefore any duplicated work is extremely costly.
This is exactly analogous the problem LLVM helps to solve. Many optimizations
and analyses can be applied across a wide range of programming languages and 
hardware, but until there was an effective IR, those were often duplicated, 
resulting in a lot of lost opportunity.

Addressing these concerns this is a major motivation for our turning to Tapir.
By standardising a powerful but simple internal representation and building
frontends and backends for it, one both make it easier to build optimizations
and analyses for a given programming model, and avoid duplicating that work
across different implementations. We'll return to this issue later in the paper
when discussing potential limitations and challenges for Tapir for other
hardware and

\section{Tapir} \label{Sec:Tapir}

\section{OpenMP Tasks} \label{Sec:OpenMP}

Initially, the cross-vendor OpenMP~\cite{spec} shared memory programming model 
focused on the execution of data parallelism by a cooperating team of threads, 
e.g., dividing the iterations of a loop among the threads. Version 3.0 of the 
OpenMP API specification introduced support for lightweight asynchronous tasks, 
designated by the application developer and scheduled onto the team of threads 
by the OpenMP run time implementation.  The \texttt{task} construct applied to 
a structured block of code creates an explicit task, and the \texttt{taskwait} 
construct waits for completion of all tasks generated by the current task.

Recursive task creations and synchronizations using the constructs result in 
an implicit directed acyclic graph (DAG) that allows both reasoning about and 
visualization of the program execution.  Figure~\ref{Fig:fib-graph-to-schedule}
shows some example code, a view of the task DAG, and a simplified execution 
schedule mapping the tasks to a team of two threads.  In the example, the Nth 
Fibonacci number is calculated by recursively generating tasks to calculate 
the (N-1)th and (N-2)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
the child tasks have completed before their answers are combined to yield the 
final result.

\lstset{basicstyle=\tt\small, language=C, 
morekeywords={omp, for, pragma, omp, task, taskwait}}
\begin{figure*}
\begin{center}
\begin{tabular}{| c c  c | c  c  c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
    \centering{
   \raisebox{-0.6in}{\includegraphics[scale=0.35]{fib-task-graph.pdf}}
    }
& & &
    \centering{
     \raisebox{-0.25in}{\includegraphics[scale=0.35]{fib-schedule.pdf}}
    }
 &  \\
\hline
\end{tabular}
\end{center}
\caption{Code, task graph, and schedule of a simple brute force recursive 
Fibonacci number calculation on two threads.}
\label{Fig:fib-graph-to-schedule}
\end{figure*}

The initial design of the OpenMP task model~\cite{ayguade09design} established 
a basic framework for asynchronous task parallel execution in OpenMP programs.
Subsequent versions of the OpenMP specification up to the current version 4.5 
have added additional new features to the tasking model.  The \texttt{depend} 
clause codifies data dependences among tasks, indicating that a data location 
is an input or output of a task.  The run time system ensures that a task is 
not scheduled until its input dependences are fulfilled.  The \texttt{taskloop} 
construct combines groups of independent loop iterations into explicit tasks, 
enabling composition of concurrent loop execution and independent explicit 
tasks within the same OpenMP parallel region. The \texttt{taskgroup} construct 
waits on not only all child tasks, but all descendent tasks, providing a deep 
synchronization. The \texttt{taskwait} construct allows the application to 
indicate a point at which the implementation may suspend the current task to 
work on other tasks, as may be desired for long-running tasks that generate 
many others.  The task concept was also leveraged to provide for asynchronous 
offload of data and computation to accelerators by applying the \texttt{nowait} 
clause to device constructs like the \texttt{target} construct to generate an 
asynchronous \textit{target task}.

Several clauses for the \texttt{task} construct aim to optimize execution of 
tasks but can be safely ignored by an implementation that chooses to do so. 
The \texttt{mergeable} clause allows the implementation to omit creation of a 
new data environment for a descendent of a task marked with the \texttt{final}
clause.  The \texttt{priority} clause assigns an integer priority to the task 
and recommends the prioritization of tasks with higher priority values.
The \texttt{untied} clause allows a task to be migrated between threads after 
suspension, which enables practical use of \textit{work-first scheduling} 
(suspending the parent task in favor of executing each child task immediately 
on the thread where it is generated).

\section{Implementation}

Tapir is implemented as an extension to the LLVM instruction set. Clang is a
C family compiler that has support for OpenMP extensions \cite{clang} and targets
LLVM. The existing Clang OpenMP implementation mapps OpenMP constructs directly
to OpenMP runtime library calls, by wrapping a C statement in a
\texttt{CapturedStmt}.  This replaces a C statement with a function call into
the OpenMP runtime library. along with a machine generated function who's body
contains that statement.

For this work, we replaced a subset of the OpenMP implementation to generate
Tapir IR instead of vanilla LLVM IR with OpenMP runtime calls. Specifically, we
replace the two primary pragmas for task parallelism: \texttt{task\_spawn} and
\texttt{task\_wait}. By re-using code from Schardl et al. for code generation
of Cilk constructs, we were able to easily generate Tapir code for these OpenMP
pragmas. The ease with which this was completed is a testament to the quality of
the Tapir implementation.

While we did implement the codegen for the \texttt{task\_spawn} and
\texttt{task\_wait} constructs, it's worth noting here that even for these
pragmas, the implementation is incomplete. Currently, any clauses modifying the 
behaviour of the pragmas is ignored. Probably the most common semantics this
will change of variable semantics, e.g. \texttt{shared} vs. \texttt{private}.
Surprisingly, this had very effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this issue would increase
the performance of this work marginally, due to reducing the number of memory
copies for variables declared \texttt{private} in OpenMP code.

The surprising fact that program behaviour wasn't changed significantly is
worth visiting here. We surmise that it is likely due to the variable semantics
borrowed from the Cilk codegen and their relation to the OpenMP specification
of semantics.  In particular, \emph{OpenMP doesn't specify any memory model for
shared variables}. This results in the Cilk style of only writing to a shared
variable at the end of a parallel section as a perfectly valid implementation
of OpenMP's specification. In contrast, existing OpenMP implementations often
write to shared variables on every write. This is a potential for performance
discrepancy due to differing implementation-defined behaviour, and we will 
return to it when discussing performance on the Barcelona OpenMP task suite.
There is of course the fact that technically, by copying the value of variables
back to the surrounding context even for \texttt{private} variables, one isn't
following the specification. We surmise that in practice this hasn't had an
effect due to the fact that \texttt{private} variables are generally used for
performance, rather than their semantic properties.

A very important property to discuss of the current implementation is the
backend. This work has \emph{only been on the frontend}, which means we use
the only existing backend for Tapir. The existing backend is to generate 
calls into the Cilk runtime, \texttt{libcilkrts}. This obviously has many 
implications for adhering to the OpenMP specification. For example, environment
variables such as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by
\texttt{CILK\_NWORKERS}. We discuss some of these issues below, and return to 
address others in the future work section.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir as described above. Generally, a program has to contain other
OpenMP pragmas in order to run parallel code. A standard pattern for building
task-parallel OpenMP programs is to start a \texttt{parallel} pragma to spin up
the necessary hardware threads, then have a \texttt{single} pragma to have only
one thread continue on the specified statement. Then the other threads will get
work from spawned tasks, instead of implicitly executing the same code in a
data parallel style. For the purpose of this paper, we took the shortcut of
replacing these pragmas with no-ops. This worked well for all examples except
for one, in which after the \texttt{parallel} and \texttt{single} pragmas the
top level task was called with an unnecessary \texttt{task} pragma. Because the
wait was implicit and unhandled by our implementation, the simple fix was to
remove the \texttt{task} pragma. This was the only change required to the 
source code. Our temporary no-op shortcut, of course, doesn't follow the OpenMP
specification, and should be addressed by future work.

\subsection{Example} \label{Sec:Example}

To better understand how the OpenMP to Tapir compiler works, we turn to the
\texttt{fib} example from Section~\ref{Sec:OpenMP}. In Figure~\ref{Fig:Example}
we show how the code generated differs between the original Clang
implementation, which generates vanilla LLVM with calls into the OpenMP
runtime, and.

\section{Evaluation}

For evaluation we compared performance on the Barcelona OpenMP task suite to 
existing OpenMP implementations. The Barcelona OpenMP task suite is a set of 
tests intended to test performance of OpenMP implementations using both
irregular and regular tasking. The benchmarks are described in
Figure~\ref{Fig:benchmark_desc}.

\begin{itemize}
\item \texttt{Alignment}: aligns all protein sequences from  an  input
file  against  every  other  sequence  using  the Myers  and Miller \cite{}
algorithm. The alignments are scored and the best score for each pair is
provided as a result. The scoring method is  a  full  dynamic  programming
algorithm. It uses  a  weight matrix to score mismatches, and assigns
penalties for opening and extending gaps. The output is the best score for each
pair of them.
\item \texttt{FFT}: computes the one-dimensional Fast Fourier Transform
of a vector of n complex values using the Cooley-Tukey \cite{cooley-tukey}
algorithm. This is a divide and conquer algorithm that  recursively  breaks
down a Discrete Fourier Transform (DFT) into many smaller DFT’s. In each of the
divisions multiple tasks are generated.
\item \texttt{Fibonacci}: computes the n'th fibonacci number using a  recursive
paralellization. While  not  representative  of  an efficient  fibonacci
computation  it  is  still  useful  because  it  is a simple test case of a
deep tree composed of very fine grain tasks.  
\item \texttt{Floorplan}: kernel computes the optimal floorplan distribution
of a number of cells. The algorithm gets an input file with  cell’s
description  and  it  returns  the  minimum  area  size which includes all
cells. This minimum area is found through a recursive branch and bound search.
We hierarchically generate tasks  for  each  branch  of  the  solution  space.
The  state  of  the algorithm needs to be copied into each newly created task
so they can proceed. This implies that additional synchronizations have been
introduced in the code to maintain the parent state alive.
\item \texttt{Health}: simulates de Columbian Health Care System \cite{?}. It
uses multilevel lists where each element in the structure  represents  a
village with  a  list  of  potential patients and one hospital. The hospital
has several double-linked lists representing the possible status of a patient
inside it (waiting, in assessment,   in   treatment   or   waiting   for
reallocation).  At  each time step  all  patients  are  simulated  according
with several probabilities (of getting sick, needing a convalescence treatment,
or  being  reallocated to  an  upper  level  hospital).  A  task  is  created
for  each  village being  simulated.  Once  the lower   levels   have   been
simulated synchronization   occurs. 
\item \texttt{NQueens}: computes  all  solutions  of  the n-queens
problem, whose objective is to find a placement for $n$ queens on an $n \;
\times \; n$ chessboard such that none of the queens attack any other. It uses
a backtracking search algorithm with pruning. A task is created for each step
of the solution.
\item \texttt{Sort}: sorts a random permutation of n 32-bit numbers with  a
fast  parallel  sorting  variation  \cite{mergesort}  of  the  ordinary
mergesort.  First, it divides an array of elements in two halves, sorting  each
half  recursively,  and  then  merging  the  sorted halves with a parallel
divide-and-conquer method rather than the  conventional  serial  merge.  Tasks
are  used  for  each  split and merge. When the array is too small, a serial
quicksort is used  so increase  the  task  granularity.  To  avoid  the
overhead of  quicksort,  an insertion  sort  is  used  for  very  small  arrays
(below a threshold of 20 elements).
\item \texttt{SparseLU}: computes an LU matrix factorization over
sparse matrices. A first level matrix is composed by pointers to  small
submatrices  that  may  not  be  allocated.  Due  to  the sparseness  of  the
matrix,  a  lot  of  imbalance  exists.  Matrix size and submatrix size can be
set at execution time. While a dynamic schedule can reduce the imbalance, a
solution with tasks parallelism seems to obtain better results \cite{}. In each
of the sparseLU  phases,  a  task  is  created  for  each  block  of  the
matrix that is not empty.
\item \texttt{Strassen}: algorithm  uses  hierarchical  decomposition of a
matrix for multiplication of large dense matrices \cite{}. Decomposition is
done by dividing each dimension of the matrix into  two  sections  of  equal
size. For each decomposition a task is created. 
\end{itemize}

In particular, we compare to GCC 7.1, Clang 4.0.1, and Intel 17.0.0. The
machine used is a two socket Intel Xeon E5-2693 v3 machine with 132GB of
memory, running Linux 4.11.4. It's worth noting that each of the other 
implementations comes with it's own runtime. In this sense, performance is
a function of both the compiler and any optimizations it's able to perform, 
and the runtime. Understanding the interaction of these two parts is non-trivial,
as we will see when discussing results.

As mentioned in the Section~\ref{Sec:Implementation}, the gaps in the implementation
changed program behaviour in a couple cases. In the case of \texttt{FFT}, we were
forced to remove an unecessary \texttt{task} pragma, as the current implementation
doesn't insert the implicit barrier at the end of a \texttt{parallel} region. This 
was a one line fix in the benchmark, and would be fixed properly by handling
the \texttt{parallel} and \texttt{single} pragmas correctly in our implementation.

The second change in program behaviour due to our incomplete implementation was
caused by the lack \texttt{critical} pragma and \texttt{atomic} pragmas. This
issue showed up in the \texttt{Floorplan} benchmark. The \texttt{critical} pragma
ensures that only one thread can executed the referred statement at a time. 
This should be fixable in the implementation by adding a simple code-localized
synchronization generation in the IR. Similarly, the \texttt{atomic} pragma 
can be addressed by retaining a pointer to the shared stack variable, much like
existing OpenMP implementations. It is worth noting that this behaviour was
non-deterministic, and that roughly half of the time the Floorplan still
returned correct results. As we will see, this makes for an interesting
trade-off given the performance increase witnessed.

In our evaluation, we caught three kinds of errors:

\begin{itemize}
\item Non-determinism in result correctness
\item Segmentation faults
\item Timeouts (10 Minutes)
\end{itemize}

We set a timeout of 10 minutes, as at least one implementation was always
finishing within ~10 seconds, anything running more than 60 times slower
becomes irrelevant. DUe to time constraints, correctness checks were not
performed on every run, so it is possible some non-determinism was missed.

\section{Results} \label{Sec:Results}

In this section we discuss the results of running our evaluation on our
implementation, and how its performance compared to the existing
implementations listed in Section~\ref{Sec:Evaluation}. See
Figure~\ref{Fig:results} for a simple visualization of the performance. Each
graph represents a single benchmark from the previously enumerated Barcelona
benchmarks, and each bar represents one of the implementations.

\begin{figure*}
\begin{multicols}{2}
  \includegraphics[width=\linewidth]{alignment.pdf} \par
  \includegraphics[width=\linewidth]{fib.pdf} 
  \includegraphics[width=\linewidth]{fft.pdf} \par
  \includegraphics[width=\linewidth]{floorplan.pdf} 
  \includegraphics[width=\linewidth]{sort.pdf} \par
  \includegraphics[width=\linewidth]{nqueens.pdf}
\end{multicols}
\caption{Barcelona OpenMP Task Suite Results}
\end{figure*}

\begin{figure*}
\begin{multicols}{2}
  \includegraphics[width=\linewidth]{health.pdf} \par
  \includegraphics[width=\linewidth]{sparselu.pdf} 
\end{multicols}
\centering
\includegraphics[width=0.45\linewidth]{strassen.pdf}
\caption{Barcelona OpenMP Task Suite Results (continued)}
\end{figure*}

The failure modes enumerated in Section~\ref{Sec:Evaluation} are denoted by 
asterisks in the implementation labels on the X axis. One asterisk refers to
some incorrect answers. Two asterisks refers to time-out. Three asterisks
refers to segmentation faults.

First, we can see that there is one failure for our implementation, on
\texttt{Floorplan}. Note that the result is non-zero. In fact, while ICC and
Clang finished in roughly 1-2 seconds, the our Tapir implementation was
finishing in \emph{0.02} seconds, and computing the correct result roughly half
the time. This raises interesting questions on the cost/benefit relation of the
OpenMP behaviour that our implementation is missing.

Other failures include ICC failing with segmentation faults on \texttt{FFT} and
\texttt{Fibonacci}. While we haven't carefully diagnosed these failures, we 
suspect that the number of tasks spawned exceeded a limit in Intel's compiler.
Note that the \texttt{FFT} failures were non-deterministic, occurring on
roughly one in three runs. Clang failed with segmentation faults on every run 
of \texttt{SparseLU}. Again, we didn't have the time to investigate these
failures, and are unsure what was causing them.

The final failures we saw were timeout failures. GCC was the only culprit for these
failures. Recall that the timeout was set at 10 minutes, so any timeout means GCC 
was running at least approximately 60 times slower than the fastest implementation. 
For example, on the \texttt{FFT} benchmark, while our Tapir implementation was running
in under 5 seconds, the GCC implementation was timing out at 600 seconds, so
was at least 100 times slower.

\subsection{Forensics}

In this section we attempt to understand \emph{why} the performance varies in the ways 
it does. While we can't hope to figure out every discrepancy in performance, we
can hope to get some ideas for why the performance of our implementation is
generally better, and where each of the implementations is spending its time.

Our primary tool for this task will be performance counters. It's worth noting
that some information is difficult to infer from performance counters, such as
sources of contention in runtimes, reasons for memory locality issues, etc.
Still, it will give us some insight into performance bottlenecks in the
program.

Before we begin understanding why, it's worth noting again that the scope of the 
work that this paper is responsible for is only the frontend. All performance
gains due to the Cilk runtime we sadly cannot claim credit for.

Another import consideration is that we can break reasons for performance
discrepencies into a few categories:

\begin{itemize}
\item Front-end code generation
\item IR Optimizations
\item Runtime efficiency
\end{itemize}

While we will try out best to distinguish between these in our forensics, it is 
difficult to do so with only performance counter information. A more thorough
investigation would require careful knowledge of each of the frontends and
runtimes. One thing we can guess at is that despite the focus of the paper,
\emph{IR optimisations are unlikely to be the cause of performance
improvements}. Our evidence for this comes from Shardl et al. \cite{}. While
some IR optimisations have been implemented, they seem to have significantly
more effect on parallel loop code , while the tasking code remains
relatively unaffected by IR optimizations \cite{}.

\section{Discussion} \label{Sec:Discussion}

\subsection{Future Work} \label{Sec:Future}

\subsection{Related Work} \label{Sec:Related}

\section{Conclusion} \label{Sec:Conclusion}
We have shown that Tapir is an excellent IR target for OpenMP tasks. While not
all of the semantics are covered, we have a path forward for many of them. We have
shown that compiling to Tapir instructions allows for a straightforward compilation of 
OpenMP tasking programs to use the Cilk runtime system. We've also shown that this
combination leads to better performance than existing OpenMP tasking implementations
on the Barcelona OpenMP Tasking benchmark suite. Moving forward, we hope efforts like
this one help to reduce the fragmentation of parallel runtimes, as well as make
it easier to write optimization for parallel programs.

\bibliographystyle{ACM-Reference-Format}
\bibliography{annotated}

\end{document}

### 2017-08-31 plot.py

@@ -1,0 +1,39 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import glob
+
+pdat = {}
+pdat['alignment'] = {'title' : "Alignment"}
+pdat['fft'] = {'title' : "FFT"}
+pdat['fib'] = {'title' : "Fibonacci"}
+pdat['floorplan'] = {'title' : "Floorplan"}
+pdat['health'] = {'title' : "Health"}
+pdat['nqueens'] = {'title' : "NQueens"}
+pdat['sort'] = {'title' : "Sort"}
+pdat['sparselu'] = {'title' : "SparseLU"}
+pdat['strassen'] = {'title' : "Strassen"}
+
+types = ['tapir', 'gcc', 'icc', 'clang']
+
+for p in pdat:
+    for t in types:
+        print(p, t)
+        fs = glob.glob('./data/%s.%s.*' %(p,t))
+        pdat[p][t] = [np.nan if len([l.split()[3] for l in open(f).readlines() if len(l.split()) > 3 and l.split()[0] == "Time"]) == 0
+                         else float([l.split()[3] for l in open(f).readlines() if len(l.split()) > 3 and l.split()[0] == "Time"][0]) 
+                         for f in fs]
+        print(np.mean(pdat[p][t]))
+
+width=0.75
+ind = np.arange(len(types))
+
+for p in pdat:
+    fig, ax = plt.subplots()
+    ax.bar(ind, [np.mean(pdat[p][t]) for t in types], width, yerr=[np.std(pdat[p][t]) for t in types])
+    ax.set_ylabel("Time (Seconds, lower is better)")
+    ax.set_xticks(ind)
+    ax.set_xticklabels(types)
+    ax.set_title(pdat[p]['title'])
+    plt.savefig('%s.pdf' %(p))
+    
+

### 2017-09-01 paper.tex
Fixed typo on correctness
Added example codegen

@@
 \usepackage{multicol}
-\usepackage{graphicx}
+\graphicspath{{./figs/}}
 
@@
 will change of variable semantics, e.g. \texttt{shared} vs. \texttt{private}.
-Surprisingly, this had very effect on the correctness of the entire Barcelona
+Surprisingly, this had little effect on the correctness of the entire Barcelona
 OpenMP task suite. Indeed, it is likely that fixing this issue would increase
@@
 
+All code used for the implementation is available at
+\texttt{https://github.com/lanl/openmpir}. The version used for this paper is
+tagged as LLVM17.
+
 \subsection{Example} \label{Sec:Example}
@@
 \texttt{fib} example from Section~\ref{Sec:OpenMP}. In Figure~\ref{Fig:Example}
-we show how the code generated differs between the original Clang
-implementation, which generates vanilla LLVM with calls into the OpenMP
-runtime, and.
+we show what Tapir code is generated.
 
-\section{Evaluation}
+\begin{figure*}
+\begin{tabular}{| c c c | c c c |}
+\hline
+ & 
+\begin{lstlisting}
+int fib(int n) 
+{                    // A
+  if (n < 2)
+    return n;
+  else
+  {
+    int x, y;
+    #pragma omp task
+      x = fib(n-1);  // B
+    #pragma omp task
+      y = fib(n-2);  // C
+    #pragma omp taskwait
+    return (x+y);    // D
+  }
+}
+\end{lstlisting}
+& & &
+\begin{lstlisting}
+if.end:                                 
+  detach within %syncreg, label %det.achd, label %det.cont
+
+det.achd:                                 
+  %2 = load i32, i32* %n.addr, align 4
+  %sub = sub nsw i32 %2, 1
+  %call = call i32 @fib(i32 %sub)
+  store i32 %call, i32* %x, align 4
+  reattach within %syncreg, label %det.cont
+
+det.cont:                                   
+  detach within %syncreg, label %det.achd1, label %det.cont4
+
+det.achd1:                                
+  %3 = load i32, i32* %n.addr, align 4
+  %sub2 = sub nsw i32 %3, 2
+  %call3 = call i32 @fib(i32 %sub2)
+  store i32 %call3, i32* %y, align 4
+  reattach within %syncreg, label %det.cont4
+
+det.cont4:                                  
+  sync within %syncreg, label %sync.continue
+\end{lstlisting}
+ &  \\
+\hline
+\end{tabular}
+
+\caption{Compilation to Tapir}
+\label{Fig:Example}
+\end{figure*}
+
+\section{Evaluation} \label{Sec:Evaluation}
 
@@
 irregular and regular tasking. The benchmarks are described in
-Figure~\ref{Fig:benchmark_desc}.
+Figure~\ref{Fig:benchmark_desc}. The benchmark suite has added an unbalanced
+tree search benchmark since the paper was published.  Unfortunately, this seems
+to be buggy, and failed on every implementation for even medium input sizes. We
+have therefore left it out of our evaluation.
 
@@
 
-In particular, we compare to GCC 7.1, Clang 4.0.1, and Intel 17.0.0. The
-machine used is a two socket Intel Xeon E5-2693 v3 machine with 132GB of
-memory, running Linux 4.11.4. It's worth noting that each of the other 
-implementations comes with it's own runtime. In this sense, performance is
-a function of both the compiler and any optimizations it's able to perform, 
-and the runtime. Understanding the interaction of these two parts is non-trivial,
-as we will see when discussing results.
+There are two classes of input sizes for the benchmarks. Some, like Floorplan,
+require input files. For each of these, we chose the largest provided. For the
+benchmarks with a parameter to adjust the size of the input, we attempted to 
+choose a parameter size so that the fastest implementation was on the order of
+seconds. The exact inputs used are:
+
+\begin{itemize}
+\item \texttt{Alignment: -f prot.100.aa} 
+\item \texttt{FFT: -n 335544320}
+\item \texttt{Fibonacci: -n 48}
+\item \texttt{Floorplan: -f input.20}
+\item \texttt{Health: -f large.input}
+\item \texttt{NQueens: -n 15}
+\item \texttt{Sort -n 335544320}
+\item \texttt{SparseLU: -n 100 -m 100}
+\item \texttt{Strassen: -n 8192}
+\end{itemize}
+
+For other implementations, we compare to GCC 7.1, Clang 4.0.1, and Intel
+17.0.0. The machine used is a two socket Intel Xeon E5-2693 v3 machine with
+132GB of memory, running Linux 4.11.4. All benchmarks were run using all 64
+hyper-threads. 32 thread tests were done as well and showed qualitatively
+similar results. It's worth noting that each of the other implementations comes
+with it's own runtime. In this sense, performance is a function of both the
+compiler and any optimizations it's able to perform, and the runtime.
+Understanding the interaction of these two parts is non-trivial, as we will see
+when discussing results.
 
@@
 \begin{itemize}
-\item Non-determinism in result correctness
 \item Segmentation faults
 \item Timeouts (10 Minutes)
+\item Non-determinism in result correctness
 \end{itemize}
@@
 finishing within ~10 seconds, anything running more than 60 times slower
-becomes irrelevant. DUe to time constraints, correctness checks were not
+becomes irrelevant. Due to time constraints, correctness checks were not
 performed on every run, so it is possible some non-determinism was missed.
+
+Each variant was run 10 times, with the height of the bars representing the
+mean and the error bars representing standard deviation. In cases where there
+was timeout or segmentation fault, no time is reported, and the run is marked
+as faulty.
 
@@
 
+Putting failures aside for the moment, performance for our implementation is
+quite strong, generally outperforming or matching the best of existing
+implementations. For benchmarks where the overhead to work ratio is high is 
+where the Tapir implementation really shines. For example, for the
+\texttt{Fibonacci} benchmark, our implementation finishes significantly faster
+than any of the others. While ICC failed for that sized input, we did test with
+smaller inputs and found its similarly outpaced by the Tapir implementation.
+\texttt{NQueens} and \texttt{FFT} show similar behaviour, to lesser degrees. In 
+contrast, benchmarks that have a lower overhead to work ratio unsurprisingly 
+differ less in performance. For example,
+
 \subsection{Forensics}
@@
 Another import consideration is that we can break reasons for performance
-discrepencies into a few categories:
+discrepancies into a few categories:
 

### 2017-09-01 paper.tex
Added Stephen's extra tex

@@
 
-All code used for the implementation is available at
+All code used for the implementation is available at \\
 \texttt{https://github.com/lanl/openmpir}. The version used for this paper is
@@
 
+\section*{Acknowledgment}
+Sandia National Laboratories is a multimission laboratory managed and operated 
+by National Technology and Engineering Solutions of Sandia, LLC., a wholly 
+owned subsidiary of Honeywell International, Inc., for the U.S. Department of 
+Energy’s National Nuclear Security Administration under contract DE-NA-0003525.
+
 \bibliographystyle{ACM-Reference-Format}

### 2017-09-05 paper.tex
Updated latest figures, added some discussion

@@
 the results of evaluation in Section~\ref{Sec:Results}, including possible
-explanations for discrepancies. Finally, we discuss how the many ways the work can
-be extended in Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.
+explanations for discrepancies. We discuss the implications of the results ans the
+significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
+we cover some of the many ways the work can be extended in
+Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.
 
@@
 choose a parameter size so that the fastest implementation was on the order of
-seconds. The exact inputs used are:
+seconds. The one exception was for \texttt{Fibonacci}. Because, as discussed
+further below, the Intel implementation required large stacksize, we stopped at
+42. The exact inputs used are:
 
@@
 \item \texttt{FFT: -n 335544320}
-\item \texttt{Fibonacci: -n 48}
+\item \texttt{Fibonacci: -n 42}
 \item \texttt{Floorplan: -f input.20}
@@
 
+The Intel compiler required increasing the stacksize for \texttt{Fibonacci} and
+\texttt{FFT}. The implementation seemed to use significantly more stack space
+than the others, requiring setting \texttt{OMP\_STACKSIZE} to \texttt{256M} for
+computing the 42nd Fibonacci number. \texttt{FFT} ran fine with an increase to
+\texttt{16M}.
+
 For other implementations, we compare to GCC 7.1, Clang 4.0.1, and Intel
-17.0.0. The machine used is a two socket Intel Xeon E5-2693 v3 machine with
-132GB of memory, running Linux 4.11.4. All benchmarks were run using all 64
-hyper-threads. 32 thread tests were done as well and showed qualitatively
-similar results. It's worth noting that each of the other implementations comes
-with it's own runtime. In this sense, performance is a function of both the
-compiler and any optimizations it's able to perform, and the runtime.
-Understanding the interaction of these two parts is non-trivial, as we will see
-when discussing results.
+17.0.0. The machine used is a two socket Intel Xeon E5-2683 v3 machine with
+132GB of memory, running Linux 4.11.4. Each Xeon has 16 cores running at
+2.1GHz, with 20M of L3 cache. All benchmarks were run using all 64
+hyper-threads. 32 thread tests were run as a sanity check and showed
+qualitatively similar results. It's worth noting that each of the other
+implementations comes with it's own runtime. In this sense, performance is a
+function of both the compiler and any optimizations it's able to perform, and
+the runtime.  Understanding the interaction of these two parts is non-trivial,
+as we will see when discussing results. Each compiler was run with the flags
+\texttt{-O3 -fopenmp}. Our compiler was also run with \texttt{-ftapir} to
+enable Tapir instructions.
 
@@
 existing OpenMP implementations. It is worth noting that this behaviour was
-non-deterministic, and that roughly half of the time the Floorplan still
-returned correct results. As we will see, this makes for an interesting
+non-deterministic, and that roughly half of the time the \texttt{Floorplan}
+still returned correct results. As we will see, this makes for an interesting
 trade-off given the performance increase witnessed.
 
-In our evaluation, we caught three kinds of errors:
-
-\begin{itemize}
-\item Segmentation faults
-\item Timeouts (10 Minutes)
-\item Non-determinism in result correctness
-\end{itemize}
-
 We set a timeout of 10 minutes, as at least one implementation was always
-finishing within ~10 seconds, anything running more than 60 times slower
-becomes irrelevant. Due to time constraints, correctness checks were not
-performed on every run, so it is possible some non-determinism was missed.
+finishing within 10 seconds, anything running more than 60 times slower becomes
+irrelevant. Due to time constraints, correctness checks were not performed on
+every run, so it is possible some non-determinism was missed.
 
@@
 
-The failure modes enumerated in Section~\ref{Sec:Evaluation} are denoted by 
-asterisks in the implementation labels on the X axis. One asterisk refers to
-some incorrect answers. Two asterisks refers to time-out. Three asterisks
-refers to segmentation faults.
+As mentioned earlier on \texttt{Floorplan} our implementation was sometimes
+giving incorrect results. It is difficult to tell, but the time to completion
+is non-zero. In fact, while ICC and Clang finished in roughly 1-2 seconds, the
+our Tapir implementation was finishing in 0.02 seconds, and computing the
+correct result roughly half the time. This raises interesting questions on the
+cost/benefit relation of the OpenMP behaviour that our implementation is
+missing. We'll discuss this further in Section~\ref{Sec:Discussion}.
 
-First, we can see that there is one failure for our implementation, on
-\texttt{Floorplan}. Note that the result is non-zero. In fact, while ICC and
-Clang finished in roughly 1-2 seconds, the our Tapir implementation was
-finishing in \emph{0.02} seconds, and computing the correct result roughly half
-the time. This raises interesting questions on the cost/benefit relation of the
-OpenMP behaviour that our implementation is missing.
+GCC was the only culprit for timeouts. Recall that the timeout was set at 10
+minutes, so any timeout means GCC was running at least approximately 60 times
+slower than the fastest implementation. For example, on the \texttt{FFT}
+benchmark, while our Tapir implementation was running in under 5 seconds, the
+GCC implementation was timing out at 600 seconds, so was at least 100 times
+slower. We've left GCC results off of the graphs for the three failure cases, 
+\texttt{FFT}, \texttt{Fibonacci}, and \texttt{NQueens}.
 
-Other failures include ICC failing with segmentation faults on \texttt{FFT} and
-\texttt{Fibonacci}. While we haven't carefully diagnosed these failures, we 
-suspect that the number of tasks spawned exceeded a limit in Intel's compiler.
-Note that the \texttt{FFT} failures were non-deterministic, occurring on
-roughly one in three runs. Clang failed with segmentation faults on every run 
-of \texttt{SparseLU}. Again, we didn't have the time to investigate these
-failures, and are unsure what was causing them.
+Putting failures aside, performance for our implementation is quite strong,
+generally outperforming or matching the best of existing implementations. For
+benchmarks where the overhead to work ratio is high is where the Tapir
+implementation really shines. For example, for the \texttt{Fibonacci}
+benchmark, our implementation finishes significantly faster than any of the
+others. While ICC failed for that sized input, we did test with smaller inputs
+and found its similarly outpaced by the Tapir implementation.  \texttt{NQueens}
+and \texttt{FFT} show similar behaviour, to lesser degrees. In contrast,
+benchmarks that have a lower overhead to work ratio unsurprisingly differ less
+in performance. For example,
 
-The final failures we saw were timeout failures. GCC was the only culprit for these
-failures. Recall that the timeout was set at 10 minutes, so any timeout means GCC 
-was running at least approximately 60 times slower than the fastest implementation. 
-For example, on the \texttt{FFT} benchmark, while our Tapir implementation was running
-in under 5 seconds, the GCC implementation was timing out at 600 seconds, so
-was at least 100 times slower.
+\subsection{Profiling}
 
-Putting failures aside for the moment, performance for our implementation is
-quite strong, generally outperforming or matching the best of existing
-implementations. For benchmarks where the overhead to work ratio is high is 
-where the Tapir implementation really shines. For example, for the
-\texttt{Fibonacci} benchmark, our implementation finishes significantly faster
-than any of the others. While ICC failed for that sized input, we did test with
-smaller inputs and found its similarly outpaced by the Tapir implementation.
-\texttt{NQueens} and \texttt{FFT} show similar behaviour, to lesser degrees. In 
-contrast, benchmarks that have a lower overhead to work ratio unsurprisingly 
-differ less in performance. For example,
-
-\subsection{Forensics}
-
-In this section we attempt to understand \emph{why} the performance varies in the ways 
-it does. While we can't hope to figure out every discrepancy in performance, we
-can hope to get some ideas for why the performance of our implementation is
-generally better, and where each of the implementations is spending its time.
+In this section we attempt to understand \emph{why} the performance varies in
+the ways it does. While we can't hope to figure out every discrepancy in
+performance, we can hope to get some ideas for why the performance of our
+implementation is generally better, and where each of the implementations is
+spending its time.
 
@@
 sources of contention in runtimes, reasons for memory locality issues, etc.
-Still, it will give us some insight into performance bottlenecks in the
-program.
+Still, it will give us some insight into performance bottlenecks.
 
@@
 work that this paper is responsible for is only the frontend. All performance
-gains due to the Cilk runtime we sadly cannot claim credit for.
+gains due to the existing Tapir backend and Cilk runtime we sadly cannot claim
+credit for.
 
-Another import consideration is that we can break reasons for performance
-discrepancies into a few categories:
+We can break reasons for performance discrepancies into a few categories:
 
@@
 
-While we will try out best to distinguish between these in our forensics, it is 
-difficult to do so with only performance counter information. A more thorough
-investigation would require careful knowledge of each of the frontends and
-runtimes. One thing we can guess at is that despite the focus of the paper,
-\emph{IR optimisations are unlikely to be the cause of performance
-improvements}. Our evidence for this comes from Shardl et al. \cite{}. While
-some IR optimisations have been implemented, they seem to have significantly
-more effect on parallel loop code , while the tasking code remains
-relatively unaffected by IR optimizations \cite{}.
+While we would like to distinguish between these, it is difficult to do so with
+only performance counter information. A more thorough investigation would
+require careful knowledge of each of the frontends and runtimes. One thing we
+can guess at is that despite the focus of the paper, \emph{IR optimisations are
+unlikely to be the cause of performance improvements}. Our evidence for this
+comes from Shardl et al. \cite{}. While some IR optimisations have been
+implemented, they seem to have significantly more effect on parallel loop code,
+while the tasking code remains relatively unaffected by IR optimizations
+\cite{}.
+
+There are two cases we'd like to look at using the profiler. The first is a 
+case when there is a large difference in runtime between implementations. For 
+this case, our hypothesis is that in these cases the slower implementations
+are, for some reason, spending more time in the runtime. As mentioned above, 
+it is difficult to understand exactly why an implementation is spending more 
+time in the runtime without further investigation, but it is a useful 
+sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
+Indeed, we get roughly the following breakdowns of work to overhead ratio according
+to the Linux \texttt{perf} tool:
+
+\begin{itemize}
+\item \texttt{tapir: ~30\% runtime overhead, ~50\% work, ~20\% other}
+\item \texttt{clang: ~85\% runtime overhead, ~5\% work, ~10\% other}
+\item \texttt{icc: ~85\% runtime overhead, ~5\% work, ~10\% other}
+\item \texttt{gcc: ~100\% runtime overhead, ~0\% work, ~0\% other}
+\end{itemize}
+
+This confirms our suspicions that most of the slowdown incurred is overhead in
+the runtime. For \texttt{tapir}, the overhead is relatively small given the
+size of the work chunks, for \texttt{clang} and \texttt{icc} it is
+significantly larger, while for \texttt{gcc} it is completely overwhelming.
+This does let us rule out the possibility that somehow the only the codegen for
+other implementations is the culprit: it is almost certainly at least partially
+the runtime implementation.
+
+For the cases where performance is close between implementations, like
+\texttt{sparselu}, we expect runtimes to be dominated by actual task work. This 
+expectation is confirmed by investigating performance counters for \texttt{sparselu}:
+every implementation spends roughly 90\% of it's time doing work in the task
+bodies.
 
@@
 
+In this section we discuss the relevance of the results and this work in general
+in the context of the literature. As discussed in Section~\ref{Sec:Introduction}, 
+there have been many programming models for implementing parallel programs. While
+this paper does not take a stance on which one \emph{should} be used, it does
+accept that OpenMP is currently a widely used model, potentially the most
+widely used model that has a notion of tasks.
+
+The goal of this work is to show that a single intermediate representation can be
+used to implement multiple programming models, with good performance. Schardl et 
+al. showed that Tapir works well as an IR for Cilk, and surmised that it could
+be an effective IR for a large subset of OpenMP. This work has taken the first
+step in that direction, showing that it can work well as an IR for OpenMP tasks.
+
+One interesting aspect of this work is that it compiles a language extension intended
+for one runtime (OpenMP runtimes) to use another (the Cilk runtime). While in many
+ways this is a weakness of this work, it does open the possibility of doing this 
+in a general way. Having a real IR to work with enables, in the same way that LLVM
+does for serial code, the possibility of multiple input parallel languages and
+multiple output parallel runtimes. While the OpenMP spec's requirement to
+have access to runtime routines complicates things, for many programming models
+it seems at least conceivable to map them onto multiple different runtimes. This
+opens the possibility of mixing programming models at compile time by targeting
+the same runtime. For example, if both OpenMP and Cilk targeted Tapir
+instructions, one could have a large application where some parts were written
+using Cilk, while other used OpenMP, and the result would not suffer the
+incompatibility problems often seen today.
+
 \subsection{Future Work} \label{Sec:Future}
 
+One of the important questions for this work is whether or not the Tapir
+instructions can be extended to cover the full OpenMP semantics. As shown
+in this paper, even the full semantics of OpenMP tasks weren't covered. Runtime
+calls aside, implementing the full semantics of OpenMP pragmas would be a 
+significant undertaking.
+
+As two examples, let us consider the problematic \texttt{atomic} and
+\texttt{critical} pragmas from the \texttt{Floorplan} benchmark. The 
+\texttt{atomic} pragma would likely not be too difficult to translate
+from existing OpenMP codegen. One would have to simply keep around and then
+modify the contents of the original stack pointer to the shared variable. The
+\texttt{critical} section would likely be a little more difficult. Still, it
+should be possible to modify existing OpenMP code to block on that section.
+
+Still, whether or not it's \emph{possible} to implement these extra features
+is a separate question from whether or not it's \emph{useful}. If implementing
+extra features require using special synchronization primitives, or calls into
+runtimes, or any other tool other than the Tapir instructions, then the
+compiler cannot reason about code using those features, and many of the
+advantages we've discussed become moot, e.g. backend compatibility, generic 
+optimization, etc. Features like the OpenMP \texttt{atomic} pragma don't 
+require any changes to control flow, and therefore are likely to work well 
+with Tapir analyses and optimizations. On the other hand features like OpenMP's
+\texttt{critical} section \emph{do} require changes to control flow that seem
+challenging at best to map onto Tapir's instructions. Furthermore, even if it
+is possible to implement features using the Tapir instructions, it may require 
+such radical entangling of the code that even if analysis is technically
+possible, it becomes infeasible.
+
 \subsection{Related Work} \label{Sec:Related}
+
+There has been significant work on internal representations for parallel
+programs. These can be roughly broken into four categories.
+
+First, compilers can attach metadata to an existing IR. For example, LLVM has a
+parallel loop metadata construct \cite{}. This has the benefit of being
+flexible and requiring minimal work, but historically can be lost during
+optimizations, and any code movement can break the semantics. The flexibility
+of this approach can also be viewed as a negative, as it can compromise
+what can otherwise be a simple, well defined semantics for the IR.
+
+A second approach is to use intrinsic functions to define parallel tasks
+\cite{ares}. This, like the metadata approach, has the advantage of being
+flexible and relatively easy to implement, but at the cost of being less well
+defined. There has recently been proposals to standardize intrinsics extensions
+for LLVM to represent parallelism, but we would caution that a large, flexible
+set of intrinsics being added to an already only-partially defined
+\cite{verillvm} language could make reasoning about parallel programs, even
+in an IR settings, infeasible.
+
+A third approach is to have a separate instruction set for parallelism. This is
+the approach taken by projects like HPIR \cite{}, SPIRE \cite{}, and INSPIRE
+\cite{}. This approach has the downside of being unable to re-use existing
+optimization and analysis infrastructure of an existing IR.
+
+Finally, the last approach, and that taken by Tapir, is to implement parallel 
+instructions as an extension of an existing IR. This allows for integration
+into existing analyses and optimizations. In the case of Tapir, this allows one
+to leverage years of development into program analysis and optimization,
+extending only where necessary.
+
+While there is general agreement that there needs to be IR support for
+parallelism in parallel programs, there isn't consensus on which of these
+approaches is best. Approaches that are easier at first, such as metadata
+or intrinsic approaches, are similarly easy to extend to be unwieldy. There is
+a reason that it's difficult to add instructions to LLVM: any added IR
+construct needs to be considered in many locations. From debugging by printing
+out IR, to case analysis for different control flow constructs, having 
+a proper set of instructions to denote parallelism has a lot of advantages. 
+Additionally, any hope of having a formal semantics for an IR depends on a 
+simple, concise set of parallelism constructs, something that metadata and
+intrinsics approaches can easily avoid.
 

### 2017-09-05 plot.py

@@ -13,7 +13,7 @@ pdat['sort'] = {'title' : "Sort"}
 pdat['sparselu'] = {'title' : "SparseLU"}
 pdat['strassen'] = {'title' : "Strassen"}
 
-types = ['tapir', 'gcc', 'icc', 'clang']
+types = ['tapir', 'icc', 'clang', 'gcc']
 
 for p in pdat:
     for t in types:
@@ -25,15 +25,19 @@ for p in pdat:
         print(np.mean(pdat[p][t]))
 
 width=0.75
-ind = np.arange(len(types))
 
 for p in pdat:
     fig, ax = plt.subplots()
-    ax.bar(ind, [np.mean(pdat[p][t]) for t in types], width, yerr=[np.std(pdat[p][t]) for t in types])
+    means = [np.mean(pdat[p][t]) for t in types]
+    means = [m for m in means if not np.isnan(m)]
+    stds = [np.std(pdat[p][t]) for t in types]
+    stds = [m for m in stds if not np.isnan(m)]
+    ind = np.arange(len(means))
+    ax.bar(ind, means, width, yerr=stds)
     ax.set_ylabel("Time (Seconds, lower is better)")
     ax.set_xticks(ind)
     ax.set_xticklabels(types)
     ax.set_title(pdat[p]['title'])
-    plt.savefig('%s.pdf' %(p))
+    plt.savefig('figs/%s.pdf' %(p))
     
 

### 2017-09-06 paper.tex
Added lanl acks

@@
 
-\section*{Acknowledgment}
+\section*{Acknowledgments}
 Sandia National Laboratories is a multimission laboratory managed and operated 
@@
 
+This work was supported by the Director, Office of Advanced Scientific
+Computing Research, Office of Science, of the United States Department
+of Energy, under the guidance of Dr. Sonia Sachs. Los Alamos National
+Laboratory is operated by Los Alamos National Security LLC for the U.S.
+Department of Energy under contract DE-AC52-06NA25396.
+
+This research was supported by US National Science Foundation under CRII
+ACI 1565338, CNS 1527076, and CNS 1217948. Los Alamos National
+Laboratory is operated by Los Alamos National Security LLC for the US
+Department of Energy under contract DE-AC52-06NA25396.
+
+Work supported by the Advanced Simulation and Computing program of the
+U.S. Department of Energy's NNSA\@.  Los Alamos National Laboratory is
+managed and operated by Los Alamos National Security, LLC (LANS), under
+contract number DE-AC52-06NA25396 for the Department of Energy’s
+National Nuclear Security Administration (NNSA).
+
 \bibliographystyle{ACM-Reference-Format}

### 2017-09-06 paper.tex
Small intro change

@@
 support. There has been some recent work on specializing compilers to reason
-about libraries \cite{kokkos}, but the fragmentation among the models that
-presents another challenge for compiler writers, which must choose a single
-model to analyze and optimize. An ideal solution would be some some form of
-common representation of parallelism that all IRs can share.
+about libraries \cite{kokkos}, but fragmentation among models presents a
+significant challenge for compiler writers, which must choose a single model to
+analyze and optimize. An ideal solution would be some some form of common
+representation of parallelism that all IRs can share.
 
@@
 optimizations to concurrent code. The recent work of Schardl et al.
-\cite{shardl2017} has shown that standard compiler IRs can be extended to
-represent fork-join parallelism. In that work, by extending the LLVM
-instruction set with three instructions, they are able to capture all of the
-semantics to implement the semantics of Cilk, an extension to C. They suggested
-that a similar approach could be taken with OpenMP tasks.
+\cite{shardl2017} has shown that real world compiler IRs can be extended to
+represent fork-join parallelism. By extending the LLVM instruction set with
+three instructions, they were able to capture all of the semantics to implement
+the semantics of Cilk, an extension to C. They suggested that a similar
+approach could be taken with OpenMP tasks.
 

### 2017-09-06 paper.tex
Initial comments from WM

@@
 \documentclass[sigconf]{acmart}
+\newcommand{\wmnote}[1]{{\scriptsize \color{red} [[ Billy: #1]]}}
 
@@
 
-Internal representations (IRs) are the compiler objects on which almost all
+Internal representations (IRs) are the \wmnote{the phrase compiler objects seems a bit odd to me}compiler objects on which almost all
 analyses and optimizations are performed. Historically, a major drawback for
@@
 
-LLVM is arguably the most successful IR to date \cite{llvm}. By implementing a simple
+\wmnote{This feels like we're trying to convince someone about LLVM: perhaps LLVM has gained traction in the academic community for both its simplicity and relatively painless modification}LLVM is arguably the most successful IR to date \cite{llvm}.
+By implementing a simple
 single static assignment (SSA) representation, it enables powerful optimizations 
@@
 
-Addressing these concerns this is a major motivation for our turning to Tapir.
-By standardising a powerful but simple internal representation and building
-frontends and backends for it, one both make it easier to build optimizations
-and analyses for a given programming model, and avoid duplicating that work
+Addressing these concerns this is a major motivation for our \wmnote{use of?}turning to Tapir.
+\wmnote{this sentence feels very loose/run-on} By standardising a powerful but simple internal representation and building
+frontends and backends for it, one both makes it easier to build optimizations
+and analyses for a given programming model, and avoids duplicating work
 across different implementations. We'll return to this issue later in the paper
@@
 The surprising fact that program behaviour wasn't changed significantly is
-worth visiting here. We surmise that it is likely due to the variable semantics
+worth \wmnote{discussion?}visiting here. We surmise that it is likely due to the variable semantics
 borrowed from the Cilk codegen and their relation to the OpenMP specification

### 2017-09-06 paper.tex
Begin work on Tapir section

@@
 
+\author{William S. Moses}
+\affiliation{\institution{MIT CSAIL}}
+\email{wmoses@mit.edu}
+
 \begin{abstract}
@@
 This work implements a frontend which compiles a subset of OpenMP tasks to
-Tapir IR. This enables analyses and optimizations previously inaccessible to
+Tapir. This enables analyses and optimizations previously inaccessible to
 OpenMP codes. Initial performance results for the Barcelona OpenMP tasking
@@
 \section{Tapir} \label{Sec:Tapir}
+Tapir is an extension to LLVM proposed by Schardl, Moses, and Leiserson
+aiming to resolve issues regarding optimizing parallel code. Prior to
+the introduction of Tapir, compilers would represent parallel programs
+by directly translating their syntax to opaque runtime calls. This allowed
+for compilers to support parallel programs, but meant that traditional 
+optimizations such as code motion \cite{} or common-subexpression elimination \cite{}
+weren't able to reason about parallel programs. This often led to parallel
+programs running significantly slower than expected. In the Tapir project,
+the authors show that it is possible to represent and optimize fork-join
+parallel programs with relatively minimal modifications to the compiler: namely
+the introduction of three instructions designed to interface well with an
+existing compiler.
+
+Specifically, Tapir introduces the \texttt{detach}, \texttt{reattach}, and
+\texttt{sync} instructions. \wmnote{How in detail should I go here? I can talk
+about these instructions on a broad level as a target for parallel tasks, or
+more concretely define what they do. However, we don't seem discuss CFG's
+before this so I'm not sure what level of introduction you're thinking of.}
 

### 2017-09-06 paper.tex
Addressing Billy's comments

@@
 \newcommand{\wmnote}[1]{{\scriptsize \color{red} [[ Billy: #1]]}}
+\newcommand{\gsnote}[1]{{\scriptsize \color{blue} [[ George: #1]]}}
 
@@
 
-Internal representations (IRs) are the \wmnote{the phrase compiler objects seems a bit odd to me}compiler objects on which almost all
-analyses and optimizations are performed. Historically, a major drawback for
-parallel codes is that IRs don't have the ability to represent, and therefore
-reason about, parallelism. This prevents the application of common
-optimizations to concurrent code. The recent work of Schardl et al.
-\cite{shardl2017} has shown that real world compiler IRs can be extended to
-represent fork-join parallelism. By extending the LLVM instruction set with
-three instructions, they were able to capture all of the semantics to implement
-the semantics of Cilk, an extension to C. They suggested that a similar
-approach could be taken with OpenMP tasks.
+Compilers use internal representations (IRs) for implementing analyses and
+optimizations. Historically, a major drawback for parallel programs is that IRs
+don't have the ability to represent, and therefore reason about, parallelism.
+This prevents the application of common optimizations to concurrent code. The
+recent work of Schardl et al.  \cite{shardl2017} has shown that real world
+compiler IRs can be extended to represent fork-join parallelism. By extending
+the LLVM instruction set with three instructions, they were able to capture all
+of the semantics to implement the semantics of Cilk, an extension to C. They
+suggested that a similar approach could be taken with OpenMP tasks.
 
@@
 
-Internal representations are a crucial part of modern compilers. By representing
-code in a generic, hardware-agnostic format, they allow both multiple frontends for
-different source languages, and multiple backends for different hardware, to all
-take advantage of a series of compiler optimizations and analyses.
+\textbf{Internal representations (IRs)} are a crucial part of modern compilers.
+By representing code in a generic, hardware-agnostic format, they allow both
+multiple frontends for different source languages, and multiple backends for
+different hardware, to all take advantage of a series of compiler optimizations
+and analyses.
 
-\wmnote{This feels like we're trying to convince someone about LLVM: perhaps LLVM has gained traction in the academic community for both its simplicity and relatively painless modification}LLVM is arguably the most successful IR to date \cite{llvm}.
-By implementing a simple
-single static assignment (SSA) representation, it enables powerful optimizations 
-that are effective for a wide range of programming languages and hardware
-\cite{llvm imps}.
+\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
+LLVM has gained traction in the academic community for both its simplicity and
+relatively painless modification} 
+\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
+(LLVM in HPC). Is this any more satisfactory?} 
+Among IRs, LLVM is arguably the most ubiquitous \cite{llvm}. Winner of the
+2012 ACM Software System Award, it is in use by frontends ranging from Haskell
+to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
+extensible \textbf{single static assignment (SSA)} IR, LLVM enables powerful
+optimizations that are effective for a wide range of programming languages and
+hardware \cite{llvm}.
 
-Historically, LLVM has had a sequential semantics: there is no notion of
-concurrent or parallel processes. This resulted in existing compilers treating 
-parallelism constructs as thin wrappers for calls into runtime systems. These
-calls are essentially impossible for the compiler to reason about, due to the
-complexity of runtimes being called.
+Historically, LLVM has had a sequential semantics: there is no explicit notion
+of concurrent or parallel processes. This has resulted in existing compilers
+treating parallelism constructs as thin wrappers for calls into runtime
+systems. These calls are infeasible for the compiler to reason about, due to
+the complexity and flexibility of runtimes being called. This situation was
+recently remedied by Schardl et al.  \cite{tapir}. By adding three instructions
+to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
+Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
+\cite{tapir}.
 
-This unfortunate situation was recently remedied by Schardl et al.
-\cite{tapir}. By adding three instructions to LLVM, Tapir enables analyses and
-optimizations of parallel code. A full description
+Arguably the most ubiquitous parallel programming model for on-node parallelism
+in HPC, OpenMP has grown from being a way to implement parallel for loops to an
+enormous and complex specification for parallelism. This paper focuses on a 
+relatively recent addition to the OpenMP specification: the \texttt{task}
+and \texttt{taskwait} constructs. A detailed description of these constructs
+is presented in Section~\ref{Sec:OpenMP}. For a more complete description see
+\cite{spec}.
 
@@
 parallel programs \cite{....}. From libraries, to language extensions, to
-standalone languages, to combinations of the above, it can be hard to choose
-the right one. This means that when writing tooling, optimizations, or
-analyses, one has to choose which model to target. One common difficulty among
-all of these is that reasoning about parallel programs is hard \cite{}, and
-compilers can help \cite{}. Therefore any duplicated work is extremely costly.
-This is exactly analogous the problem LLVM helps to solve. Many optimizations
-and analyses can be applied across a wide range of programming languages and 
-hardware, but until there was an effective IR, those were often duplicated, 
-resulting in a lot of lost opportunity.
+standalone languages, to combinations of the above, fragmentation of parallel
+programming models is a problem that is only getting worse. This means that
+when writing tooling, optimizations, or analyses, one has to choose which model
+to target. One common difficulty among all of these is that reasoning about
+parallel programs is hard \cite{}, and compilers can help \cite{}. Therefore
+any duplicated work is extremely costly.  This is exactly analogous the problem
+LLVM helps to solve.  Many optimizations and analyses can be applied across a
+wide range of programming languages and hardware, but until there was an
+effective IR, those were often duplicated, resulting in a lot of lost
+opportunity.
 
-Addressing these concerns this is a major motivation for our \wmnote{use of?}turning to Tapir.
-\wmnote{this sentence feels very loose/run-on} By standardising a powerful but simple internal representation and building
-frontends and backends for it, one both makes it easier to build optimizations
-and analyses for a given programming model, and avoids duplicating work
-across different implementations. We'll return to this issue later in the paper
-when discussing potential limitations and challenges for Tapir for other
-hardware and
+Addressing these concerns this is a major motivation for our use of Tapir.
+By standardising an IR, in addition to easing the development of optimizations,
+one avoids duplicating work across different implementations. We'll return to
+this issue later in Section~\ref{Sec:Future} when discussing future work.
 
@@
 before this so I'm not sure what level of introduction you're thinking of.}
+\gsnote{I was thinking roughly a page should be enough. Maybe a summary of what
+Tapir is followed by an example of an optimization it enables. I like both the
+CSE and tail-call optimization examples from your PPoPP paper. Feel free to 
+define CFG and any other terms you need to.}
 
@@
 C family compiler that has support for OpenMP extensions \cite{clang} and targets
-LLVM. The existing Clang OpenMP implementation mapps OpenMP constructs directly
+LLVM. The existing Clang OpenMP implementation maps OpenMP constructs directly
 to OpenMP runtime library calls, by wrapping a C statement in a
 \texttt{CapturedStmt}. This replaces a C statement with a function call into
-the OpenMP runtime library. along with a machine generated function who's body
+the OpenMP runtime library, along with a machine generated function who's body
 contains that statement.
@@
 the performance of this work marginally, due to reducing the number of memory
-copies for variables declared \texttt{private} in OpenMP code.
+copies for variables declared \texttt{private} in OpenMP code. This property
+that behaviour wasn't changed significantly is worth revisiting, and we do so
+in Section~\ref{Sec:Discussion}.
 
-The surprising fact that program behaviour wasn't changed significantly is
-worth \wmnote{discussion?}visiting here. We surmise that it is likely due to the variable semantics
-borrowed from the Cilk codegen and their relation to the OpenMP specification
-of semantics.  In particular, \emph{OpenMP doesn't specify any memory model for
-shared variables}. This results in the Cilk style of only writing to a shared
-variable at the end of a parallel section as a perfectly valid implementation
-of OpenMP's specification. In contrast, existing OpenMP implementations often
-write to shared variables on every write. This is a potential for performance
-discrepancy due to differing implementation-defined behaviour, and we will 
-return to it when discussing performance on the Barcelona OpenMP task suite.
-There is of course the fact that technically, by copying the value of variables
-back to the surrounding context even for \texttt{private} variables, one isn't
-following the specification. We surmise that in practice this hasn't had an
-effect due to the fact that \texttt{private} variables are generally used for
-performance, rather than their semantic properties.
-
-A very important property to discuss of the current implementation is the
+A very important property of the current implementation to discuss is the
 backend. This work has \emph{only been on the frontend}, which means we use
-the only existing backend for Tapir. The existing backend is to generate 
-calls into the Cilk runtime, \texttt{libcilkrts}. This obviously has many 
-implications for adhering to the OpenMP specification. For example, environment
-variables such as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by
-\texttt{CILK\_NWORKERS}. We discuss some of these issues below, and return to 
-address others in the future work section.
+the only existing backend for Tapir. The existing backend generates calls into
+the Cilk runtime, \texttt{libcilkrts}. This obviously has many implications for
+adhering to the OpenMP specification. For example, environment variables such
+as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
+We discuss some of these issues below, and return to address others in the
+future work section.
 
 There are other issues that had to be overcome to run full OpenMP programs
-using Tapir as described above. Generally, a program has to contain other
-OpenMP pragmas in order to run parallel code. A standard pattern for building
-task-parallel OpenMP programs is to start a \texttt{parallel} pragma to spin up
-the necessary hardware threads, then have a \texttt{single} pragma to have only
-one thread continue on the specified statement. Then the other threads will get
-work from spawned tasks, instead of implicitly executing the same code in a
-data parallel style. For the purpose of this paper, we took the shortcut of
-replacing these pragmas with no-ops. This worked well for all examples except
-for one, in which after the \texttt{parallel} and \texttt{single} pragmas the
-top level task was called with an unnecessary \texttt{task} pragma. Because the
-wait was implicit and unhandled by our implementation, the simple fix was to
-remove the \texttt{task} pragma. This was the only change required to the 
-source code. Our temporary no-op shortcut, of course, doesn't follow the OpenMP
-specification, and should be addressed by future work.
+using Tapir as described above. Generally, in addition the the \texttt{task} 
+and \texttt{taskwait} pragmas, a program has to contain initialization OpenMP
+pragmas in order to run parallel code. A standard pattern for building
+task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
+the necessary hardware threads, followed immediately by a \texttt{single}
+pragma to have only one thread continue on the specified statement. Then the
+other threads will get work from spawned tasks, instead of implicitly executing
+the same code in a data parallel style. For the purpose of this paper, we took
+the shortcut of replacing these pragmas with no-ops. This worked well for all
+examples except for one, in which after the \texttt{parallel} and
+\texttt{single} pragmas the top level task was called with an unnecessary
+\texttt{task} pragma.  Because the wait was implicit and unhandled by our
+implementation, the simple fix was to remove the \texttt{task} pragma. This was
+the only change required to the source code. Our temporary no-op shortcut, of
+course, doesn't follow the OpenMP specification, and should be addressed by
+future work.
 
 All code used for the implementation is available at \\
-\texttt{https://github.com/lanl/openmpir}. The version used for this paper is
-tagged as LLVM17.
+\texttt{https://github.com/lanl/openmpir-clang}. The version used for this
+paper is tagged as LLVM17.
 
@@
 
+The surprising fact that program behaviour wasn't changed significantly is
+worth discussing here. We surmise that it is likely due to the variable
+semantics borrowed from the Cilk codegen and their relation to the OpenMP
+specification of semantics. In particular, \emph{OpenMP doesn't specify any
+memory model for shared variables}. This results in the Cilk style of only
+writing to a shared variable at the end of a parallel section as a perfectly
+valid implementation of OpenMP's specification. In contrast, existing OpenMP
+implementations often write to shared variables on every write. This is a
+potential for performance discrepancy due to differing implementation-defined
+behaviour, and we will return to it when discussing performance on the
+Barcelona OpenMP task suite. There is of course the fact that technically, by
+copying the value of variables back to the surrounding context even for
+\texttt{private} variables, one isn't following the specification. We surmise
+that in practice this hasn't had an effect due to the fact that
+\texttt{private} variables are generally used for performance, rather than
+their semantic properties.
+
 \subsection{Future Work} \label{Sec:Future}

### 2017-09-07 paper.tex
Made cilk backend more obvious

@@
 This work implements a frontend which compiles a subset of OpenMP tasks to
-Tapir. This enables analyses and optimizations previously inaccessible to
-OpenMP codes. Initial performance results for the Barcelona OpenMP tasking
-suite and show performance improvements over existing OpenMP compiler
-implementations. 
+Tapir, an extension to LLVM that represents fork-join parallelism. This enables
+analyses and optimizations previously inaccessible to OpenMP codes, as well as
+the ability to target other runtimes. Using a Cilk runtime backend, we compare
+to existing OpenMP implementations. Initial performance results for the
+Barcelona OpenMP tasking suite and show performance improvements over existing
+OpenMP compiler implementations. 
 \end{abstract}
@@
 compiling them to Tapir instructions. This enables us to target non-openmp
-runtimes and compare performance to existing OpenMP implementations. Specifically,
-we run our prototype implementation on the Barcelona OpenMP task suite
+runtimes and compare performance to existing OpenMP implementations. For this
+paper, our prototype implementation uses a Cilk runtime as a backend, and we run
+our prototype implementation on the Barcelona OpenMP task suite
 \cite{barcelona}. We discuss the kinds of optimizations this enables, and how
@@
 \item A prototype frontend from OpenMP task constructs to Tapir IR
-\item Benchmarks of the implementation, showing performance improvements over
-existing OpenMP implementations
+\item Benchmarks of the implementation using a Cilk runtime backend, showing
+performance improvements over existing OpenMP implementations
 \item An initial investigation into the reasons for the performance improvement
@@
 of work required. We argue that the benefits vastly outweigh the costs, and
-hope to see more work follow suite.
+hope to see more work follow suit.
 
@@
 to target. One common difficulty among all of these is that reasoning about
-parallel programs is hard \cite{}, and compilers can help \cite{}. Therefore
-any duplicated work is extremely costly.  This is exactly analogous the problem
-LLVM helps to solve.  Many optimizations and analyses can be applied across a
-wide range of programming languages and hardware, but until there was an
-effective IR, those were often duplicated, resulting in a lot of lost
-opportunity.
+parallel programs is hard \cite{}, and compilers can help \cite{}. This is
+exactly analogous the problem LLVM solves: compiler optimizations are
+hard to do well, so sharing them is valuable. Tapir is a natural extension
+
+With fragmentation comes questions of compatibility, or composability, of 
+different programming models. To illustrate this issue, consider the example of
+a program using a library that uses OpenMP to implement parallelism internally. 
+If said program uses a different programming model for parallelism, e.g. Cilk
+, there is currently no good solution for
 
@@
 CSE and tail-call optimization examples from your PPoPP paper. Feel free to 
-define CFG and any other terms you need to.}
+define CFG and any other terms you need to. If that's too much to ask on such 
+short notice just let me know and I'll help out.}
 
@@
 
+While OpenMP and Cilk have similar fork-join style semantics that Tapir
+implements naturally, many allow a more flexible mechanism for synchronization:
+futures \cite{qthreads, hpx, chapel, charm++}. The primary difference
+between the fork-join parallelism represented by Tapir and these approaches
+is that \emph{child tasks can outlive their parents}. Figuring out how
+to represent this class of parallelism in Tapir represents a significant
+challenge for future work.
+
 \subsection{Related Work} \label{Sec:Related}

### 2017-09-07 paper.tex
Fixed small typo.
Various improvements

@@
 \usepackage{autofe}
+
+\lstset{numbers=left,
+         numberstyle=\tiny,
+         numbersep=5pt,
+         xleftmargin=0.1in
+}
 
@@
 support. There has been some recent work on specializing compilers to reason
-about libraries \cite{kokkos}, but fragmentation among models presents a
-significant challenge for compiler writers, which must choose a single model to
-analyze and optimize. An ideal solution would be some some form of common
-representation of parallelism that all IRs can share.
+about parallel programming libraries \cite{Moss_2016}, but fragmentation among
+models presents a significant challenge for compiler writers, which must choose
+a single model to analyze and optimize. An ideal solution would be some form of
+common representation of parallelism in an internal representation \textbf{(IR)}.
 
@@
 optimizations. Historically, a major drawback for parallel programs is that IRs
-don't have the ability to represent, and therefore reason about, parallelism.
-This prevents the application of common optimizations to concurrent code. The
-recent work of Schardl et al.  \cite{shardl2017} has shown that real world
+haven't had the ability to represent, and therefore reason about, parallelism.
+This has prevented the application of common optimizations to concurrent code.
+The recent work of Schardl et al.  \cite{shardl2017} has shown that real world
 compiler IRs can be extended to represent fork-join parallelism. By extending
@@
 
-\textbf{Internal representations (IRs)} are a crucial part of modern compilers.
-By representing code in a generic, hardware-agnostic format, they allow both
+Internal representations (IRs) are a crucial part of modern compilers.  By
+representing code in a generic, hardware-agnostic format, they allow both
 multiple frontends for different source languages, and multiple backends for
@@
 to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
-extensible \textbf{single static assignment (SSA)} IR, LLVM enables powerful
-optimizations that are effective for a wide range of programming languages and
-hardware \cite{llvm}.
+extensible single static assignment IR, LLVM enables powerful optimizations
+that are effective for a wide range of programming languages and hardware
+\cite{llvm}.
 
@@
 
-In this section we discuss a significant challenge in HPC: fragmentation of
-parallel programming models. There are many, many, many ways to write
-parallel programs \cite{....}. From libraries, to language extensions, to
-standalone languages, to combinations of the above, fragmentation of parallel
-programming models is a problem that is only getting worse. This means that
-when writing tooling, optimizations, or analyses, one has to choose which model
-to target. One common difficulty among all of these is that reasoning about
-parallel programs is hard \cite{}, and compilers can help \cite{}. This is
-exactly analogous the problem LLVM solves: compiler optimizations are
-hard to do well, so sharing them is valuable. Tapir is a natural extension
+In this section we further discuss a significant challenge in HPC:
+fragmentation of parallel programming models. There are many, many, many ways
+to write parallel programs \cite{....}. From libraries, to language extensions,
+to standalone languages, to combinations of the above, fragmentation of
+parallel programming models is a problem that is only getting worse. This means
+that when writing tooling, optimizations, or analyses, one has to choose which
+model to target. One common difficulty among all of these is that reasoning
+about parallel programs is hard \cite{}, and compilers can help \cite{}. This
+is exactly analogous the problem LLVM solves: compiler optimizations are hard
+to do well, so sharing them is valuable. Tapir is a demonstration that the 
+the solutions in the case of serial code can be effectively extended to the 
+case of parallel code.
 
@@
 If said program uses a different programming model for parallelism, e.g. Cilk
-, there is currently no good solution for
+, there is currently no good solution for ensuring that the two different backends
+will cooperate with management of hardware resources.
 
@@
 By standardising an IR, in addition to easing the development of optimizations,
-one avoids duplicating work across different implementations. We'll return to
-this issue later in Section~\ref{Sec:Future} when discussing future work.
+one avoids duplicating work across different implementations. It also enables
+the possibility of multiple programming models coexisting peacefully. We'll
+return to this issue later in Section~\ref{Sec:Discussion} when discussing
+future work.
 
@@
 
-\lstset{basicstyle=\tt\small, language=C, 
-morekeywords={omp, for, pragma, omp, task, taskwait}}
 \begin{figure*}
@@
 
-Tapir is implemented as an extension to the LLVM instruction set. Clang is a
-C family compiler that has support for OpenMP extensions \cite{clang} and targets
+Tapir is implemented as an extension to the LLVM instruction set. Clang is a C
+family compiler that has support for OpenMP extensions \cite{clang} and targets
 LLVM. The existing Clang OpenMP implementation maps OpenMP constructs directly
@@
 
-A very important property of the current implementation to discuss is the
-backend. This work has \emph{only been on the frontend}, which means we use
-the only existing backend for Tapir. The existing backend generates calls into
-the Cilk runtime, \texttt{libcilkrts}. This obviously has many implications for
-adhering to the OpenMP specification. For example, environment variables such
-as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
+This work has \emph{only been on the frontend}, which means we use the only
+existing backend for Tapir, which generates code that calls into the Cilk
+runtime, \texttt{libcilkrts}. This has implications for adhering to the OpenMP
+specification. For example, environment variables such as
+\texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
 We discuss some of these issues below, and return to address others in the
@@
 course, doesn't follow the OpenMP specification, and should be addressed by
-future work.
+future work. See Section~\ref{Sec:Future} for a further discussion how how the 
+implementation can be improved.
 
@@
 & & &
-\begin{lstlisting}
+\begin{lstlisting}[language=llvm]
 if.end:                                 
@@
 \end{figure*}
+
+The listed Tapir LLVM IR code corresponds to lines 7 through 12 of the C source
+code.  Upon entering the \texttt{else} branch, the code immediately detaches
+the basic block corresponding to the call to \texttt{fib(n-1)}, labeled
+\texttt{det.achd}. The continuation for the first call also immediately
+detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
+Finally, the continuation for the second call immediately calls sync. This is
+actually a nice example of code that will be optimized by Tapir. The LLVM code
+shown in Figure~\ref{Fig:Example} is not optimized. With optimizataions turned
+on, Tapir will replace a \texttt{detach} of a function call followed
+immediately by \texttt{sync} with a simple function call, resulting in faster
+code.
 

### 2017-09-07 paper.tex
Continue on Tapir section

@@
 \usepackage{multicol}
+\usepackage{minted}
 \graphicspath{{./figs/}}
@@
 \usepackage{listings}
+\usepackage{subfigure}
+\newcommand{\figlabel}[1]   {\label{fig:#1}}
+\newcommand{\figref}[1]         {Figure~\ref{fig:#1}}
+\newcommand{\figreftwo}[2]      {Figures \ref{fig:#1} and~\ref{fig:#2}}
+\newcommand{\figrefthree}[3]      {Figures \ref{fig:#1}, \ref{fig:#2} and~\ref{fig:#3}}
+\newcommand{\subfiglabel}[1]    {\textbf{#1}}
+\newcommand{\subfigcap}[1]      {\textbf{~~#1}}
+\newcommand{\subfigref}[2]      {Figure~\ref{fig:#1}#2}
+\newcommand{\subfigreftwo}[3]      {Figures~\ref{fig:#1}#2 and~\ref{fig:#1}#3}
+
+\definecolor{CodeColor}{RGB}{98,39,26}
+\def\code{\lstinline[basicstyle=\ttfamily\color{CodeColor}]}
 
@@
 \begin{document}
+\lstset{basicstyle=\tt\small, language=C, 
+morekeywords={omp, for, pragma, omp, task, taskwait}}
 
@@
 existing compiler.
+
+\subsection{Compilation without Tapir}
+
+\setminted{fontsize=\small,baselinestretch=1}
+\begin{figure}[t]
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l@{}l}
+
+\subfiglabel{a}\\
+\begin{lstlisting}
+void search(int low, int high) {
+  if (low == high) search_base(low);
+  else {
+    #pragma omp task
+    search(low, (low+high)/2);
+    #pragma omp task
+    search((low+high)/2 + 1, high);
+    #pragma omp taskwait
+  } 
+}
+\end{lstlisting}
+\\
+\subfiglabel{b}\\
+\begin{lstlisting}
+void search(int low, int high) {
+  if (low == high) search_base(low);
+  else {
+    int mid = (low+high)/2;
+    #pragma omp task
+    search(low, mid);
+    #pragma omp task
+    search(mid + 1, high);
+    #pragma omp taskwait
+  } 
+}
+\end{lstlisting}
+\vspace{0.1ex}
+\end{tabular*}
+
+\caption[Example of common-subexpression elimination on an OpenMP
+    program.]{Example of common-subexpression elimination on an OpenMP
+    program.  \subfigcap{a}~The function \code{search}, which uses
+    parallel divide-and-conquer to apply the function
+    \code{search_base} to every integer in the closed interval
+    [\code{low}, \code{high}].  \subfigcap{b}~An optimized version of
+    \code{search}, where the common subexpression \code{(low+high)/2}
+    of the original version
+    is computed only once and stored in the variable \code{mid} in
+    the optimized version.}
+  \label{fig:search}
+\end{figure}
+
+Consider the first code segment defined in \figref{search}.
+Here, we find a simple OpenMP program which performs a divide-and-conquer search.
+However, as written, there is a simple optimization that can be performed --
+namely, that the value \code{(low+high)/2} may be computed once before either
+parallel task, allowing the program to avoid some redundant computation. On
+the serial version of this program, optimizations like this are performed
+automatically. However, without the use of Tapir, these sorts of optimizations
+are not run on parallel programs.
+
+If one were to compile this program using the traditional technique of converting
+to runtime calls, one would find a program similar to \figref{runtime_calls}.
+The OpenMP pragmas are effectively treated as syntactic sugar for such runtime calls
+because vanilla LLVM has no concise way to represent parallel programs.
+These runtime calls obfuscate the program, making it impossible to analyze or
+optimize parallel code.
+
+\begin{figure}[t]
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l@{}l}
+\begin{lstlisting}
+void task_1(int* clos) {
+  search(*clos[0], (*clos[0] + *clos[1])/2);
+}
+
+void task_2(int* clos) {
+  search(*clos[0], (*clos[0] + *clos[1])/2);
+}
+
+void search(int low, int high) {
+  if (low == high) search_base(low);
+  else {
+    int* closure_1[] = {&low, &high};
+    omp_run_task(task_1, closure_1);
+    int* closure_2[] = {&low, &high};
+    omp_run_task(task_2, closure_2);
+    omp_taskwait();
+  } 
+}
+\end{lstlisting}
+\vspace{0.1ex}
+\end{tabular*}
+
+\caption[Simplified compilation of the unoptimized search code from \figref{search}.]{Simplified compilation of the unoptimized search code from \figref{search}.
+\subfigcap{a}~The function \code{search}.
+The parallel tasks are moved into their own functions and accompanying closures.
+These functions are then passed to OpenMP runtime calls.
+This system obfuscates the program, making it impossible for an optimizer to reason about what is happening.}
+  \label{fig:runtime_calls}
+\end{figure}
+
+\subsection{Tapir Internals}
+Tapir is a compelling solution to this problem because it allows existing
+analysis and optimizations to work on parallel programs without much modification.
+Tapir does this by extending LLVM's instruction set to natively represent
+fork-join parallel programs in a way that
+
+... TODO ...
 

### 2017-09-07 paper.tex
Added citations

@@
 This has prevented the application of common optimizations to concurrent code.
-The recent work of Schardl et al.  \cite{shardl2017} has shown that real world
+The recent work of Schardl et al. \cite{tapir} has shown that real world
 compiler IRs can be extended to represent fork-join parallelism. By extending
@@
 
-The remainder of the paper is organized as follows: In Section~\ref{Sec:background} 
+The remainder of the paper is organized as follows: In Section~\ref{Sec:Background} 
 we give background and motivation for the paper. Following that, we describe OpenMP
@@
 (LLVM in HPC). Is this any more satisfactory?} 
-Among IRs, LLVM is arguably the most ubiquitous \cite{llvm}. Winner of the
-2012 ACM Software System Award, it is in use by frontends ranging from Haskell
-to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
-extensible single static assignment IR, LLVM enables powerful optimizations
-that are effective for a wide range of programming languages and hardware
-\cite{llvm}.
+Among IRs, LLVM is arguably the most ubiquitous \cite{lattner2004llvm}. Winner
+of the 2012 ACM Software System Award, it is in use by frontends ranging from
+Haskell to C++, and backends ranging from x86 to PTX \cite{lattner2004llvm}.
+Implementing a simple, extensible single static assignment IR, LLVM enables
+powerful optimizations that are effective for a wide range of programming
+languages and hardware \cite{lattner2004llvm}.
 
@@
 In this section we further discuss a significant challenge in HPC:
-fragmentation of parallel programming models. There are many, many, many ways
-to write parallel programs \cite{....}. From libraries, to language extensions,
-to standalone languages, to combinations of the above, fragmentation of
-parallel programming models is a problem that is only getting worse. This means
-that when writing tooling, optimizations, or analyses, one has to choose which
-model to target. One common difficulty among all of these is that reasoning
-about parallel programs is hard \cite{}, and compilers can help \cite{}. This
-is exactly analogous the problem LLVM solves: compiler optimizations are hard
-to do well, so sharing them is valuable. Tapir is a demonstration that the 
-the solutions in the case of serial code can be effectively extended to the 
-case of parallel code.
+fragmentation of parallel programming models. There are many ways to write
+parallel programs \cite{spec, qthreads, chapel, cilk, kokkos, legion}. From
+libraries, to language extensions, to standalone languages, to combinations of
+the above, fragmentation of parallel programming models is a problem that is
+only getting worse. This means that when writing tooling, optimizations, or
+analyses, one has to choose which model to target. One common difficulty among
+all of these is that reasoning about parallel programs is hard, and compilers
+can help. This is exactly analogous the problem LLVM solves: compiler
+optimizations are hard , so sharing them is valuable. Tapir is a demonstration
+that the the LLVM approach to compilation of serial code can be effectively
+extended to the case of parallel code.
 
@@
 
-\section{Implementation}
+\section{Implementation} \label{Sec:Implementation}
 
@@
 
-For evaluation we compared performance on the Barcelona OpenMP task suite to 
-existing OpenMP implementations. The Barcelona OpenMP task suite is a set of 
-tests intended to test performance of OpenMP implementations using both
-irregular and regular tasking. The benchmarks are described in
-Figure~\ref{Fig:benchmark_desc}. The benchmark suite has added an unbalanced
-tree search benchmark since the paper was published.  Unfortunately, this seems
-to be buggy, and failed on every implementation for even medium input sizes. We
-have therefore left it out of our evaluation.
+For evaluation we compared performance on the Barcelona OpenMP task suite
+\cite{barcelona} to existing OpenMP implementations. The Barcelona OpenMP task
+suite is a set of tests intended to test performance of OpenMP implementations
+using both irregular and regular tasking. The benchmark suite has added an
+unbalanced tree search benchmark since the paper was published. Unfortunately
+we were unable to succesfully run it on any implementation, even for even
+medium input sizes. We have therefore left it out of our evaluation. The
+following descriptions are abridged versions taken verbatim from
+\cite{barcelona}:
 
@@
 \item \texttt{Alignment}: aligns all protein sequences from  an  input
-file  against  every  other  sequence  using  the Myers  and Miller \cite{}
+file  against  every  other  sequence  using  the Myers and Miller 
 algorithm. The alignments are scored and the best score for each pair is
@@
 \item \texttt{FFT}: computes the one-dimensional Fast Fourier Transform
-of a vector of n complex values using the Cooley-Tukey \cite{cooley-tukey}
+of a vector of n complex values using the Cooley-Tukey 
 algorithm. This is a divide and conquer algorithm that  recursively  breaks
@@
 introduced in the code to maintain the parent state alive.
-\item \texttt{Health}: simulates de Columbian Health Care System \cite{?}. It
+\item \texttt{Health}: simulates the Columbian Health Care System. It
 uses multilevel lists where each element in the structure  represents  a
@@
 or  being  reallocated to  an  upper  level  hospital).  A  task  is  created
-for  each  village being  simulated.  Once  the lower   levels   have   been
-simulated synchronization   occurs. 
+for  each  village being  simulated. Once the lower levels have been simulated
+synchronization occurs. 
 \item \texttt{NQueens}: computes  all  solutions  of  the n-queens
@@
 \item \texttt{Sort}: sorts a random permutation of n 32-bit numbers with  a
-fast  parallel  sorting  variation  \cite{mergesort}  of  the  ordinary
-mergesort.  First, it divides an array of elements in two halves, sorting  each
-half  recursively,  and  then  merging  the  sorted halves with a parallel
-divide-and-conquer method rather than the  conventional  serial  merge.  Tasks
-are  used  for  each  split and merge. When the array is too small, a serial
-quicksort is used  so increase  the  task  granularity.  To  avoid  the
-overhead of  quicksort,  an insertion  sort  is  used  for  very  small  arrays
-(below a threshold of 20 elements).
+fast  parallel  sorting  variation of  the  ordinary mergesort.  First, it
+divides an array of elements in two halves, sorting  each half  recursively,
+and  then  merging  the  sorted halves with a parallel divide-and-conquer
+method rather than the  conventional  serial  merge.  Tasks are  used  for
+each  split and merge. When the array is too small, a serial quicksort is used
+so increase  the  task  granularity.  To  avoid  the overhead of  quicksort, an
+insertion  sort  is  used  for  very  small  arrays (below a threshold of 20
+elements).
 \item \texttt{SparseLU}: computes an LU matrix factorization over
@@
 set at execution time. While a dynamic schedule can reduce the imbalance, a
-solution with tasks parallelism seems to obtain better results \cite{}. In each
+solution with tasks parallelism seems to obtain better results. In each
 of the sparseLU  phases,  a  task  is  created  for  each  block  of  the
@@
 \item \texttt{Strassen}: algorithm  uses  hierarchical  decomposition of a
-matrix for multiplication of large dense matrices \cite{}. Decomposition is
-done by dividing each dimension of the matrix into  two  sections  of  equal
-size. For each decomposition a task is created. 
+matrix for multiplication of large dense matrices. Decomposition is done by
+dividing each dimension of the matrix into  two  sections  of  equal size. For
+each decomposition a task is created. 
 \end{itemize}

### 2017-09-07 paper.tex
Added citations

@@
 
-\section{Introduction}
+\section{Introduction} \label{Sec:Introduction}
 
@@
 
-This work has \emph{only been on the frontend}, which means we use the only
-existing backend for Tapir, which generates code that calls into the Cilk
-runtime, \texttt{libcilkrts}. This has implications for adhering to the OpenMP
-specification. For example, environment variables such as
-\texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
-We discuss some of these issues below, and return to address others in the
-future work section.
+This work represented in this paper is only frontend implementation, which
+means we use the only existing backend for Tapir. The existing backend
+generates code that calls into the Cilk runtime, \texttt{libcilkrts}. This has
+implications for adhering to the OpenMP specification. For example, environment
+variables such as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by
+\texttt{CILK\_NWORKERS}.  We discuss some of these issues below, and return to
+address others in the future work section.
 
@@
 implementations listed in Section~\ref{Sec:Evaluation}. See
-Figure~\ref{Fig:results} for a simple visualization of the performance. Each
-graph represents a single benchmark from the previously enumerated Barcelona
-benchmarks, and each bar represents one of the implementations.
+Figures~\ref{Fig:results} and \ref{Fig:results2} for a simple visualization of
+the performance. Each graph represents a single benchmark from the previously
+enumerated Barcelona benchmarks, and each bar represents one of the
+implementations.
 
@@
 \caption{Barcelona OpenMP Task Suite Results}
+\label{Fig:results}
 \end{figure*}
@@
 \caption{Barcelona OpenMP Task Suite Results (continued)}
+\label{Fig:results2}
 \end{figure*}
@@
 unlikely to be the cause of performance improvements}. Our evidence for this
-comes from Shardl et al. \cite{}. While some IR optimisations have been
+comes from Shardl et al. \cite{tapir}. While some IR optimisations have been
 implemented, they seem to have significantly more effect on parallel loop code,
 while the tasking code remains relatively unaffected by IR optimizations
-\cite{}.
+\cite{tapir}.
 
@@
 implements naturally, many allow a more flexible mechanism for synchronization:
-futures \cite{qthreads, hpx, chapel, charm++}. The primary difference
-between the fork-join parallelism represented by Tapir and these approaches
-is that \emph{child tasks can outlive their parents}. Figuring out how
-to represent this class of parallelism in Tapir represents a significant
-challenge for future work.
+futures \cite{qthreads, chapel, hpx}. The primary difference between the
+fork-join parallelism represented by Tapir and these approaches is that
+child tasks can outlive their parents. Figuring out how to represent this class
+of parallelism in Tapir represents a significant challenge for future work.
 
@@
 A third approach is to have a separate instruction set for parallelism. This is
-the approach taken by projects like HPIR \cite{}, SPIRE \cite{}, and INSPIRE
-\cite{}. This approach has the downside of being unable to re-use existing
-optimization and analysis infrastructure of an existing IR.
+the approach taken by projects like HPIR \cite{zhao2011intermediate}, SPIRE
+\cite{khaldi2012spire}, and INSPIRE \cite{jordan2013inspire}. This approach has
+the downside of being unable to re-use existing optimization and analysis
+infrastructure of an existing IR.
 

### 2017-09-07 paper.tex
Billy to be second author
Small fixes in intro

@@
 
+\author{William S. Moses}
+\affiliation{\institution{MIT CSAIL}}
+\email{wmoses@mit.edu}
+
 \author{Stephen Olivier}
@@
 \email{pat@lanl.gov}
-
-\author{William S. Moses}
-\affiliation{\institution{MIT CSAIL}}
-\email{wmoses@mit.edu}
 
@@
 a single model to analyze and optimize. An ideal solution would be some form of
-common representation of parallelism in an internal representation \textbf{(IR)}.
-
-Compilers use internal representations (IRs) for implementing analyses and
-optimizations. Historically, a major drawback for parallel programs is that IRs
-haven't had the ability to represent, and therefore reason about, parallelism.
-This has prevented the application of common optimizations to concurrent code.
-The recent work of Schardl et al. \cite{tapir} has shown that real world
-compiler IRs can be extended to represent fork-join parallelism. By extending
-the LLVM instruction set with three instructions, they were able to capture all
-of the semantics to implement the semantics of Cilk, an extension to C. They
-suggested that a similar approach could be taken with OpenMP tasks.
+common representation of parallelism.
+
+Compilers use internal representations \textbf{(IRs)} for implementing the
+majority of their analyses and optimizations \cite{llvm}. Historically, a major
+drawback for parallel programs is that IRs haven't had the ability to
+represent, and therefore reason about, parallelism.  This has prevented the
+application of common optimizations to concurrent code.  The recent work of
+Schardl et al. \cite{tapir} has shown that real world compiler IRs can be
+extended to represent fork-join parallelism. By extending the LLVM instruction
+set with three instructions, they were able to capture all of the semantics to
+implement the semantics of Cilk, an extension to C. They suggested that a
+similar approach could be taken with OpenMP tasks.
 

### 2017-09-07 paper.tex
Small internal formatting
Removed 'within syncreg' constructs

@@
 if.end:                                 
-  detach within %syncreg, label %det.achd, label %det.cont
+  detach label %det.achd, label %det.cont
 
@@
   store i32 %call, i32* %x, align 4
-  reattach within %syncreg, label %det.cont
+  reattach label %det.cont
 
 det.cont:                                   
-  detach within %syncreg, label %det.achd1, label %det.cont4
+  detach label %det.achd1, label %det.cont4
 
@@
   store i32 %call3, i32* %y, align 4
-  reattach within %syncreg, label %det.cont4
+  reattach label %det.cont4
 
 det.cont4:                                  
-  sync within %syncreg, label %sync.continue
+  sync label %sync.continue
 \end{lstlisting}

### 2017-09-07 paper.tex
Revise intro / continue to flesh out tapir

@@
 
+\usepackage{tikz}
+\usetikzlibrary{decorations.pathreplacing,shapes,arrows,positioning,calc}
+
+\tikzset{
+  basic block/.style={
+    rectangle,
+    rounded corners,
+    draw=black, thin,
+    % minimum width=3.2cm,
+    inner sep=3pt,
+    % outer sep=4pt,
+    align=left,
+    anchor=west,
+    fill=codebgcolor, 
+    % font={\ttfamily\fontsize{8.5}{10.2}\selectfont},
+    font={\ttfamily\fontsize{7.5}{9}\selectfont},
+    text=terminalcolor,
+    draw=codeheadcolor,
+    thick,
+    auto,
+    node distance=.4cm and 1.25cm,
+  },
+  cfedge/.style={
+    font=\itshape,
+    draw=black,
+    ->,
+    >=stealth'
+  },
+  process/.style={
+    draw,
+    fill=orange!50,
+    rectangle,
+    minimum height=1.5em,
+    minimum width=5em,
+    align=center,
+    font=\small,
+  },
+  form/.style={
+    draw,
+    fill=blue!30,
+    ellipse,
+    minimum height=1em,
+    align=center,
+    font=\footnotesize,
+  },
+  slave/.style={
+    execute at end picture={
+      \coordinate (lower right) at (current bounding box.south east);
+      \coordinate (upper left) at (current bounding box.north west);
+      \pgfresetboundingbox
+      \path (upper left) rectangle (lower right);
+    }
+  }
+}
+
 \definecolor{CodeColor}{RGB}{98,39,26}
@@
 programming models. Unfortunately, despite having significant overlap in
-semantics, parallel programming models like OpenMP, Cilk, Kokkos, HPX, Charm++,
-Qthreads, pthreads, MPI, Chapel, UPC, etc. have virtually no ability to
-interoperate, both at compile time and run time. In addition, many of these
-programming models are implemented as libraries, and therefore have no compiler
-support. There has been some recent work on specializing compilers to reason
+semantics, parallel programming models such as OpenMP\cite{}, Cilk\cite{},
+Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
+Qthreads\cite{}, pthreads\cite{}, MPI\cite{}, Chapel\cite{}, UPC\cite{},
+among others have virtually no ability to interoperate either at compile
+time or run time.
+\wmnote{Not sure if relevant, but there was some experiment run a while back of someone composing a library that internally used OpenMP
+with another framework that had terribly results compared with either used alone.}
+Many of these programming models are implemented as runtime libraries with only
+primitive compiler support. As a result, most of these parallel frameworks
+aren't able to take advantage of the compiler's analysis and optimizations.
+There has been some recent work on specializing compilers to reason
 about parallel programming libraries \cite{Moss_2016}, but fragmentation among
@@
 
-Compilers use internal representations \textbf{(IRs)} for implementing the
-majority of their analyses and optimizations \cite{llvm}. Historically, a major
-drawback for parallel programs is that IRs haven't had the ability to
-represent, and therefore reason about, parallelism.  This has prevented the
-application of common optimizations to concurrent code.  The recent work of
+Compilers implement the majority of their analyses and optimizations on
+intermediate representations \textbf{(IRs)} which allow such transformations
+to be written once with relative ease and apply to a variety of source languages
+such as C or C++.
+Historically, the IR of mainstream compilers such as LLVM\cite{llvm} or GCC's Gimple\cite{}
+haven't suppored parallel programs. As a consequence, compilers relying on these
+IRs haven't had the ability to
+represent, and therefore reason about, parallelism. The recent work of
 Schardl et al. \cite{tapir} has shown that real world compiler IRs can be
-extended to represent fork-join parallelism. By extending the LLVM instruction
-set with three instructions, they were able to capture all of the semantics to
-implement the semantics of Cilk, an extension to C. They suggested that a
-similar approach could be taken with OpenMP tasks.
+extended to represent fork-join parallelism without requiring a major rewrite.
+They do so by extending the LLVM instruction set with three instructions
+capable of representing fork-join parallelism. To benchmark performance and test
+accuracy, they wrote a frontend which translated programs written with Cilk, a parallel
+programming model for C/C++, into Tapir. They also wrote a backend to \textbf{lower}
+take Tapir programs to vanilla LLVM with Cilk runtime calls.
+They suggested that a similar approach could be taken with OpenMP and other
+such frameworks.
 
 In this work, we put that suggestion to the test, implementing OpenMP tasks by
-compiling them to Tapir instructions. This enables us to target non-openmp
-runtimes and compare performance to existing OpenMP implementations. For this
-paper, our prototype implementation uses a Cilk runtime as a backend, and we run
-our prototype implementation on the Barcelona OpenMP task suite
+compiling them to Tapir instructions. This allows us to both analyze and optimize
+OpenMP programs as well as compile programs written in OpenMP to use other runtimes.
+For this paper, our prototype implementation uses a Cilk runtime as a backend,
+and we run our prototype implementation on the Barcelona OpenMP task suite
 \cite{barcelona}. We discuss the kinds of optimizations this enables, and how
@@
 
-\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
-LLVM has gained traction in the academic community for both its simplicity and
-relatively painless modification} 
 \gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
 (LLVM in HPC). Is this any more satisfactory?}
-Among IRs, LLVM is arguably the most ubiquitous \cite{lattner2004llvm}. Winner
-of the 2012 ACM Software System Award, it is in use by frontends ranging from
-Haskell to C++, and backends ranging from x86 to PTX \cite{lattner2004llvm}.
-Implementing a simple, extensible single static assignment IR, LLVM enables
-powerful optimizations that are effective for a wide range of programming
-languages and hardware \cite{lattner2004llvm}.
-
+\wmnote{It still feels like we're trying to advocate for LLVM. I'd prefer some
+more objective-sounding language like perhaps the one below.}
+LLVM has gained traction in the academic community for its simplicity,
+relatively painless modification, and extensibility\cite{lattner2004llvm}.
 Historically, LLVM has had a sequential semantics: there is no explicit notion
@@
 
-\setminted{fontsize=\small,baselinestretch=1}
 \begin{figure}[t]
@@
 analysis and optimizations to work on parallel programs without much modification.
-Tapir does this by extending LLVM's instruction set to natively represent
-fork-join parallel programs in a way that
-
-... TODO ...
-
-Specifically, Tapir introduces the \texttt{detach}, \texttt{reattach}, and
-\texttt{sync} instructions. \wmnote{How in detail should I go here? I can talk
-about these instructions on a broad level as a target for parallel tasks, or
-more concretely define what they do. However, we don't seem discuss CFG's
-before this so I'm not sure what level of introduction you're thinking of.}
-\gsnote{I was thinking roughly a page should be enough. Maybe a summary of what
-Tapir is followed by an example of an optimization it enables. I like both the
-CSE and tail-call optimization examples from your PPoPP paper. Feel free to 
-define CFG and any other terms you need to. If that's too much to ask on such 
-short notice just let me know and I'll help out.}
+To properly describe Tapir, one must briefly introduce key aspects of LLVM.
+LLVM represents control flow through a control-flow graph (CFG) of blocks,
+which are sequences of instructions. Blocks always end with terminator instructions
+which define the edges of the control flow graph. For example, \figref{CFG} shows a simple
+function and its corresponding LLVM code.
+
+\begin{figure*}[h!]
+  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}ll}
+
+    \subfiglabel{a} & \subfiglabel{b} \\
+\begin{minipage}[T]{0.45\linewidth}
+      \begin{lstlisting}
+int f(int x) {
+  int y = 10;
+  if (x == 0) {
+    y = 12;
+  }
+  return y;
+}
+      \end{lstlisting}
+    \end{minipage}
+&
+    \begin{minipage}[T]{0.45\linewidth}
+\begin{lstlisting}[language=llvm]
+define i32 @f(i32 %x) {
+entry:
+  %y = alloca i32, align 4
+  store i32 10, i32* %y, align 4
+  %cmp = icmp eq i32 %x, 0
+  br i1 %cmp, label %if.then, label %if.end
+
+if.then:
+  store i32 12, i32* %y, align 4
+  br label %if.end
+
+if.end:
+  %1 = load i32, i32* %y, align 4
+  ret i32 %1
+}
+\end{lstlisting}
+    \end{minipage}\\
+    \addlinespace[2ex]
+    \bottomrule
+  \end{tabular*}
+  \caption{Example code snippet and the corresponding LLVM code.
+This code segment contains three blocks: \code{entry}, \code{if.then}, and \code{if.end}.
+The conditional statement is implemented in LLVM by the conditional branch at the end of the entry block.
+}
+  \figlabel{CFG}
+  \vspace{-.4cm}
+\end{figure*}
+
+To represent parallelism, Tapir introduces the \code{detach}, \code{reattach}, and
+\code{sync} instructions. The \code{detach} statement is syntactically similar to
+the conditional branch as it is also a terminator instruction with with two blocks.
+However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
+the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
+\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
+is analagous to a \code{fork} in most fork-join frameworks.
+
+At the end of any task created by a \code{detach} instruction, one must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
+is used to denote the end of a parallel task.
+
+Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
+by \code{detach} instructions within the current function to complete. The \code{sync}
+instruction is analagous to a \code{join} in most fork-join frameworks.
 
@@
 an implicit directed acyclic graph (DAG) that allows both reasoning about and 
-visualization of the program execution.  Figure~\ref{Fig:fib-graph-to-schedule}
+visualization of the program execution.  \figref{fib-graph-to-schedule}
 shows some example code, a view of the task DAG, and a simplified execution 
@@
 Fibonacci number calculation on two threads.}
-\label{Fig:fib-graph-to-schedule}
+\label{fig:fib-graph-to-schedule}
 \end{figure*}
@@
 To better understand how the OpenMP to Tapir compiler works, we turn to the
-\texttt{fib} example from Section~\ref{Sec:OpenMP}. In Figure~\ref{Fig:Example}
+\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:Example}
 we show what Tapir code is generated.
@@
 \caption{Compilation to Tapir}
-\label{Fig:Example}
+\label{fig:Example}
 \end{figure*}
@@
 owned subsidiary of Honeywell International, Inc., for the U.S. Department of 
-Energy’s National Nuclear Security Administration under contract DE-NA-0003525.
+Energy's National Nuclear Security Administration under contract DE-NA-0003525.
 
@@
 managed and operated by Los Alamos National Security, LLC (LANS), under
-contract number DE-AC52-06NA25396 for the Department of Energy’s
+contract number DE-AC52-06NA25396 for the Department of Energy's
 National Nuclear Security Administration (NNSA).

### 2017-09-07 paper.tex
Addressing Billy's concerns re: performance conclusions

@@
 only performance counter information. A more thorough investigation would
-require careful knowledge of each of the frontends and runtimes. One thing we
-can guess at is that despite the focus of the paper, \emph{IR optimisations are
-unlikely to be the cause of performance improvements}. Our evidence for this
-comes from Shardl et al. \cite{tapir}. While some IR optimisations have been
-implemented, they seem to have significantly more effect on parallel loop code,
-while the tasking code remains relatively unaffected by IR optimizations
-\cite{tapir}.
+require deep knowledge of each of the frontends and runtimes. While the
+importance of optimizations of parallel code is clear, we hypothesize that most
+of the disparity present in these results is a function of runtime
+implementation differences. Our evidence for this comes from Shardl et al.
+\cite{tapir}. Generally, the effect of the optimizations that have been
+implemented for Tapir IR are outweighed by runtime contention, memory accesses,
+etc. for parallel runs. \gsnote{Billy, this OK? Don't want to sound too
+negative about the importance of the optimizations you've implemented.}
 
@@
 sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
-Indeed, we get roughly the following breakdowns of work to overhead ratio according
-to the Linux \texttt{perf} tool:
+Indeed, we get roughly the following breakdowns of work to overhead ratio
+according to the Linux \texttt{perf} tool:
 
@@
 
-This confirms our suspicions that most of the slowdown incurred is overhead in
+This supports our hypothesis that most of the slowdown incurred is overhead in
 the runtime. For \texttt{tapir}, the overhead is relatively small given the
@@
 significantly larger, while for \texttt{gcc} it is completely overwhelming.
-This does let us rule out the possibility that somehow the only the codegen for
-other implementations is the culprit: it is almost certainly at least partially
-the runtime implementation.
 
@@
 al. showed that Tapir works well as an IR for Cilk, and surmised that it could
-be an effective IR for a large subset of OpenMP. This work has taken the first
-step in that direction, showing that it can work well as an IR for OpenMP tasks.
+be an effective IR for a large subset of OpenMP \cite{tapir}. This work has
+taken the first step in that direction, showing that it can work well as an IR
+for OpenMP tasks.
 
@@
 
+An obvious first step towards better OpenMP coverage, other than the features
+listed above, is OpenMP \texttt{parallel for} loops. These are similar enough
+to \texttt{cilk\_for} loops in semantics that this also should be a relatively
+painless next step to adapt from the existing Cilk implementation.
+
 While OpenMP and Cilk have similar fork-join style semantics that Tapir

### 2017-09-07 paper.tex
Minor fixes

--- 2017-09-07 paper.tex
@@
 \texttt{-O3 -fopenmp}. Our compiler was also run with \texttt{-ftapir} to
-enable Tapir instructions.
+enable Tapir instructions to be lowered to a parallel runtime (in this case,
+the default Cilk runtime).
 
@@
 unlikely to be the cause of performance improvements}. Our evidence for this
-comes from Shardl et al. \cite{tapir}. While some IR optimisations have been
+comes from Schardl et al. \cite{tapir}. While some IR optimisations have been
 implemented, they seem to have significantly more effect on parallel loop code,

### 2017-09-07 paper.tex

--- 2017-09-07 paper.tex
@@
 of the disparity present in these results is a function of runtime
-implementation differences. Our evidence for this comes from Shardl et al.
+implementation differences. Our evidence for this comes from Schardl et al.
 \cite{tapir}. Generally, the effect of the optimizations that have been
@@
 negative about the importance of the optimizations you've implemented.}
+\wmnote{So actually the memory accesses are happily an optimization Tapir does
+-- mainly some mem2reg optimization that may not make much of a different (though still something)
+on one core, is a lot more impactful on multicores because of issues regarding contention
+forcing a cache line flush.
+Also I'm not convinced there wasn't an impact on FFT/nqueens because we did see a
+decent improvement from tapir on that -- though I wouldn't be suprised to see the runtime
+contribute a large chunk of that.
+Fibonacci is certainly runtime based because there's basically no work inside of
+it so its almost entirely a measurement of the runtime efficiency.
+Sort I'm not sure on depending on the implementation
+}
 

### 2017-09-07 paper.tex
Continue to respond to comments

@@
 decent improvement from tapir on that -- though I wouldn't be suprised to see the runtime
-contribute a large chunk of that.
+contribute a large chunk of that, especially for nqueens because that seems like a much
+larger improvement than the serial optimizations applied to parallel program alone (when we tested).
 Fibonacci is certainly runtime based because there's basically no work inside of

### 2017-09-08 paper.tex
Changed the discussion of optimization vs runtime

--- 2017-09-07 paper.tex
@@
 with another framework that had terribly results compared with either used alone.} 
+\gsnote{Hmm. A quick googling turned up a project out of berkeley combining OpenMP and TBB, 
+but performance didn't seem too bad. I'll take another look tomorrow.}
 Many of these programming models are implemented as runtime libraries with only
@@
 
-\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
-(LLVM in HPC). Is this any more satisfactory?}
-\wmnote{It still feels like we're trying to advocate for LLVM. I'd prefer some
-more objective-sounding language like perhaps the one below.}
 LLVM has gained traction in the academic community for its simplicity,
@@
 because vanilla LLVM has no concise way to represent parallel programs.
-These runtime calls obfuscate the program, making it impossible to analyze or
+These runtime calls obfuscate the program, making it infeasable to analyze or
 optimize parallel code.
@@
 These functions are then passed to OpenMP runtime calls.
-This system obfuscates the program, making it impossible for an optimizer to reason about what is happening.}
+This system obfuscates the program, making it infeasable for an optimizer to reason about what is happening.}
   \label{fig:runtime_calls}
@@
 of the disparity present in these results is a function of runtime
-implementation differences. Our evidence for this comes from Schardl et al.
-\cite{tapir}. Generally, the effect of the optimizations that have been
-implemented for Tapir IR are outweighed by runtime contention, memory accesses,
-etc. for parallel runs. \gsnote{Billy, this OK? Don't want to sound too
-negative about the importance of the optimizations you've implemented.}
-\wmnote{So actually the memory accesses are happily an optimization Tapir does
--- mainly some mem2reg optimization that may not make much of a different (though still something)
-on one core, is a lot more impactful on multicores because of issues regarding contention
-forcing a cache line flush.
-Also I'm not convinced there wasn't an impact on FFT/nqueens because we did see a
-decent improvement from tapir on that -- though I wouldn't be suprised to see the runtime
-contribute a large chunk of that.
-Fibonacci is certainly runtime based because there's basically no work inside of
-it so its almost entirely a measurement of the runtime efficiency.
-Sort I'm not sure on depending on the implementation
-}
+implementation differences. As justification for this hypothesis, consider
+the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
+and the code leaves little room for optimization, algorithmic changes
+notwithstanding. In other words, it is effectively a benchmark of how fast a
+runtime can execute fork-join parallelism. \gsnote{still worth looking into the
+optimization I mention in the implemtnation section.}
+
+%\wmnote{So actually the memory accesses are happily an optimization Tapir does
+%-- mainly some mem2reg optimization that may not make much of a different
+%(though still something) on one core, is a lot more impactful on multicores
+%because of issues regarding contention forcing a cache line flush.  Also I'm
+%not convinced there wasn't an impact on FFT/nqueens because we did see a decent
+%improvement from tapir on that -- though I wouldn't be suprised to see the
+%runtime contribute a large chunk of that, especially for nqueens because that
+%seems like a much larger improvement than the serial optimizations applied to
+%parallel program alone (when we tested).  Fibonacci is certainly runtime based
+%because there's basically no work inside of it so its almost entirely a
+%measurement of the runtime efficiency.  Sort I'm not sure on depending on the
+%implementation
+%}
 

### 2017-09-08 paper.tex
Some edits and clean up of things.  Pushing here to sync up w/ George about Rice & Cray work on extending LLVM for PGAS-like constructs.  We should probably call this out...

@@
 \newcommand{\gsnote}[1]{{\scriptsize \color{blue} [[ George: #1]]}}
+\newcommand{\pmnote}[1]{{\scriptsize \color{magenta} [[ Pat: #1]]}}
 
@@
 \begin{abstract}
+\pmnote{Overhauled this some -- take a look...}
 Optimizing compilers for task-level parallelism are still in their infancy.
-This work implements a frontend which compiles a subset of OpenMP tasks to
-Tapir, an extension to LLVM that represents fork-join parallelism. This enables
-analyses and optimizations previously inaccessible to OpenMP codes, as well as
-the ability to target other runtimes. Using a Cilk runtime backend, we compare
-to existing OpenMP implementations. Initial performance results for the
-Barcelona OpenMP tasking suite and show performance improvements over existing
-OpenMP compiler implementations. 
+This work explores a frontend that compiles OpenMP tasking semantics to Tapir, 
+an extension to LLVM IR that represents fork-join parallelism. This enables
+analyses and optimizations that were previously inaccessible to OpenMP codes, as well as
+the ability to target additional runtimes at code generation. Using a Cilk runtime 
+backend, we compare results to existing OpenMP implementations. Initial performance 
+results for the Barcelona OpenMP tasking suite and show performance improvements over 
+these existing implementations.
 \end{abstract}
@@
 When writing task-parallel programs today, one has a large selection of
-programming models. Unfortunately, despite having significant overlap in
-semantics, parallel programming models such as OpenMP\cite{}, Cilk\cite{},
-Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
+potential programming models and implementations to consider. Unfortunately, 
+despite having some significant overlaps in semantics, parallel programming models 
+such as OpenMP\cite{}, Cilk\cite{},Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
 Qthreads\cite{}, pthreads\cite{}, MPI\cite{}, Chapel\cite{}, UPC\cite{},
-among others have virtually no ability to interoperate either at compile
-time or run time.
+among others have limited ability to interoperate either at the level of 
+compile-time analysis or at run-time execution.
 \wmnote{Not sure if relevant, but there was some experiment run a while back of someone composing a library that internally used OpenMP
@@
 but performance didn't seem too bad. I'll take another look tomorrow.}
-Many of these programming models are implemented as runtime libraries with only
-primitive compiler support. As a result, most of these parallel frameworks
-aren't able to take advantage of the compiler's analysis and optimizations.
-There has been some recent work on specializing compilers to reason
-about parallel programming libraries \cite{Moss_2016}, but fragmentation among
+In addition, many of these programming models are implemented as runtime libraries with only
+primitive compiler support. As a result, most are not able to take advantage of the compiler's 
+analysis and optimization capabilities.
+While there has been some recent work on specializing compilers to reason
+about parallel programming libraries~\cite{Moss_2016}, fragmentation among
 models presents a significant challenge for compiler writers, which must choose
-a single model to analyze and optimize. An ideal solution would be some form of
-common representation of parallelism.
+a single model, or multiple intermediate forms, to use for analysis and optimization. An ideal 
+solution would be a form of common representation of parallel semantics and constructs.
 
@@
 intermediate representations \textbf{(IRs)} which allow such transformations
-to be written once with relative ease and apply to a variety of source languages
-such as C or C++.
+to be written once with relative ease and apply to a variety of source front-end
+languages such as C or C++.
 Historically, the IR of mainstream compilers such as LLVM\cite{llvm} or GCC's Gimple\cite{}
-haven't suppored parallel programs. As a consequence, compilers relying on these
-IRs haven't had the ability to
-represent, and therefore reason about, parallelism. The recent work of
-Schardl et al. \cite{tapir} has shown that real world compiler IRs can be
+haven't suppored parallel constructs. As a consequence, compilers utilizing these
+IRs haven't had the ability to reason about parallelism. The recent work of
+Schardl et al.~\cite{tapir} has shown that the LLVM IR can be
 extended to represent fork-join parallelism without requiring a major rewrite.
-They do so by extending the LLVM instruction set with three instructions
+They did so by extending the LLVM instruction set with three instructions
 capable of representing fork-join parallelism. To benchmark performance and test
 accuracy, they wrote a frontend which translated programs written with Cilk, a parallel
-programming model for C/C++, into Tapir. They also wrote a backend to \textbf{lower}
-take Tapir programs to vanilla LLVM with Cilk runtime calls.
+programming model for C/C++, into the Tapir IR. They also implemented a backend to \textbf{lower}
+Tapir programs to vanilla LLVM IR with embedded Cilk runtime calls.
 They suggested that a similar approach could be taken with OpenMP and other
@@
 OpenMP programs as well as compile programs written in OpenMP to use other runtimes.
-For this paper, our prototype implementation uses a Cilk runtime as a backend,
-and we run our prototype implementation on the Barcelona OpenMP task suite
-\cite{barcelona}. We discuss the kinds of optimizations this enables, and how
-the work could be extended to include other parallelism constructs and semantics
+For this paper, our prototype implementation uses the Cilk runtime as a backend,
+and we run our prototype implementation on the Barcelona OpenMP task suite~\cite{barcelona}. 
+We discuss the kinds of optimizations this enables, and how
+the work can be extended to include other parallelism constructs and semantics
 in OpenMP. We also discuss the possibility of extending this work to other
@@
 \begin{itemize}
-\item A prototype frontend from OpenMP task constructs to Tapir IR
+  \item A prototype frontend that transforms OpenMP task constructs to Tapir IR.
   \item Benchmarks of the implementation using a Cilk runtime backend, showing
-performance improvements over existing OpenMP implementations
-\item An initial investigation into the reasons for the performance improvement
+        performance improvements over existing OpenMP implementations.
+  \item An initial investigation into the reasons for these performance improvements.
 \end{itemize}
 
-Our hope for this paper is that it moves the community towards a shared IR that
-can be used for many programming models. This work presented in this paper only
-represents a small step in that direction, and there is still a herculean amount
-of work required. We argue that the benefits vastly outweigh the costs, and
-hope to see more work follow suit.
-
-\subsection{Outline}
+Our hope for this paper is to move the community towards further conversations on a 
+shared IR that can be used for many programming models. The work presented in this paper only
+represents a small step in that direction and there is still a herculean amount
+of design and implementaton details to be considered. We argue that the benefits vastly outweigh 
+the costs and we hope to see additional future work explore additional possibilities.
+
+% I'd skip the subsection...  Not critical, just personal prefs... 
+% \subsection{Outline}
 
@@
 
-Internal representations (IRs) are a crucial part of modern compilers.  By
+Internal representations are a crucial part of modern compiler design and implementation.  By
 representing code in a generic, hardware-agnostic format, they allow both
-multiple frontends for different source languages, and multiple backends for
-different hardware, to all take advantage of a series of compiler optimizations
-and analyses.
-
-LLVM has gained traction in the academic community for its simplicity,
-relatively painless modification, and extensibility\cite{lattner2004llvm}.
-Historically, LLVM has had a sequential semantics: there is no explicit notion
+multiple frontends and backends to all take advantage of a common/shared series of compiler 
+analyses and optimizations.
+
+LLVM has gained traction in the broad community for its simplicity,
+relative ease of both experimentation and modifications, and extensibility~\cite{lattner2004llvm}.
+Historically, LLVM has only had a sequential semantics: there is no explicit notion
 of concurrent or parallel processes. This has resulted in existing compilers
@@
 systems. These calls are infeasible for the compiler to reason about, due to
-the complexity and flexibility of runtimes being called. This situation was
-recently remedied by Schardl et al. \cite{tapir}. By adding three instructions
-to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
-Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
-\cite{tapir}.
+the complexity and flexibility of the runtimes being called \pmnote{and the challenge of
+reasoning about the corresponding semantics in a uniform fashion?}. The addition of parallel
+constructs were recently explored by Schardl et al.~\cite{tapir}. By adding three first-class 
+instructions to LLVM, Tapir enables analyses and optimizations of parallel constructs. We briefly 
+describe the key aspects of Tapir in Section~\ref{Sec:Tapir}.
+% I removed this as the reference was already given...   PM 
+% For a more complete description see \cite{tapir}.
 
 Arguably the most ubiquitous parallel programming model for on-node parallelism
-in HPC, OpenMP has grown from being a way to implement parallel for loops to an
-enormous and complex specification for parallelism. This paper focuses on a 
+in HPC, OpenMP has grown from being a way to implement parallel loops into a
+large and complex specification for parallelism. This paper focuses on a 
 relatively recent addition to the OpenMP specification: the \texttt{task}
 and \texttt{taskwait} constructs. A detailed description of these constructs
-is presented in Section~\ref{Sec:OpenMP}. For a more complete description see
-\cite{spec}.
+is presented in Section~\ref{Sec:OpenMP} and a more complete description can 
+be found in the OpenMP specification~\cite{spec}.
 
@@
 fragmentation of parallel programming models. There are many ways to write
-parallel programs \cite{spec, qthreads, chapel, cilk, kokkos, legion}. From
+parallel programs~\cite{spec, qthreads, chapel, cilk, kokkos, legion} \pmnote{I
+would add MPI and UPC/PGAS here too}. From
 libraries, to language extensions, to standalone languages, to combinations of
@@
 only getting worse. This means that when writing tooling, optimizations, or
-analyses, one has to choose which model to target. One common difficulty among
-all of these is that reasoning about parallel programs is hard, and compilers
-can help. This is exactly analogous the problem LLVM solves: compiler
-optimizations are hard , so sharing them is valuable. Tapir is a demonstration
-that the the LLVM approach to compilation of serial code can be effectively
-extended to the case of parallel code.
+analyses, one has to choose which model to target as well as understand points
+of contention and interoperability between them. A common difficulty among
+all of these is that reasoning about parallel programs is hard and compilers
+can potentially help reduce this burden. This is exactly analogous the problem 
+LLVM solves: compiler optimizations are hard, so sharing them is valuable. Tapir 
+is a demonstration that the the LLVM approach to compilation of serial code can 
+be effectively extended to help reason about parallel code.
 
@@
 a program using a library that uses OpenMP to implement parallelism internally. 
-If said program uses a different programming model for parallelism, e.g. Cilk
-, there is currently no good solution for ensuring that the two different backends
+If said program uses a different programming model for parallelism, e.g. Cilk,
+there is currently no good solution for ensuring that the two different backends
 will cooperate with management of hardware resources.
@@
 Addressing these concerns this is a major motivation for our use of Tapir.
+\pmnote{This made me remember that Rice and Cray folks did some work on extending LLVM
+to reason about PGAS-like constructs We should probably call that out in 
+this discussion -- that said, I don't recall their approach being general so it 
+does not quite fit the discussion here but I'd have to think a bit harder if it
+would be possible to capture something like MPI operations in some fashion... 
+I will email George a pointer to the paper...} 
 By standardising an IR, in addition to easing the development of optimizations,

### 2017-09-08 paper.tex
More citations

@@
 potential programming models and implementations to consider. Unfortunately, 
-despite having some significant overlaps in semantics, parallel programming models 
-such as OpenMP\cite{}, Cilk\cite{},Kokkos\cite{}, HPX\cite{}, Charm++\cite{},
-Qthreads\cite{}, pthreads\cite{}, MPI\cite{}, Chapel\cite{}, UPC\cite{},
-among others have limited ability to interoperate either at the level of 
-compile-time analysis or at run-time execution.
-\wmnote{Not sure if relevant, but there was some experiment run a while back of someone composing a library that internally used OpenMP
-with another framework that had terribly results compared with either used alone.} 
-\gsnote{Hmm. A quick googling turned up a project out of berkeley combining OpenMP and TBB, 
-but performance didn't seem too bad. I'll take another look tomorrow.}
-In addition, many of these programming models are implemented as runtime libraries with only
-primitive compiler support. As a result, most are not able to take advantage of the compiler's 
-analysis and optimization capabilities.
-While there has been some recent work on specializing compilers to reason
-about parallel programming libraries~\cite{Moss_2016}, fragmentation among
-models presents a significant challenge for compiler writers, which must choose
-a single model, or multiple intermediate forms, to use for analysis and optimization. An ideal 
-solution would be a form of common representation of parallel semantics and constructs.
+despite having some significant overlaps in semantics, parallel programming
+models such as OpenMP~\cite{openmp}, Cilk~\cite{cilk}, Kokkos~\cite{kokkos},
+HPX~\cite{hpx}, Charm++~\cite{charm}, Qthreads~\cite{qthreads},
+pthreads\cite{pthreads}, MPI~\cite{mpi}, Chapel~\cite{chapel}, UPC~\cite{upc},
+OpenCL~\cite{opencl}, OpenACC~\cite{openacc} among others have limited ability to
+interoperate either at the level of compile-time analysis or at run-time
+execution.  \wmnote{Not sure if relevant, but there was some experiment run a
+while back of someone composing a library that internally used OpenMP with
+another framework that had terribly results compared with either used alone.}
+\gsnote{Hmm. A quick googling turned up a project out of berkeley combining
+OpenMP and TBB, but performance didn't seem too bad. I'll take another look
+tomorrow.} In addition, many of these programming models are implemented as
+runtime libraries with only primitive compiler support. As a result, most are
+not able to take advantage of the compiler's analysis and optimization
+capabilities.  While there has been some recent work on specializing
+compilers to reason about parallel programming libraries~\cite{Moss_2016},
+fragmentation among models presents a significant challenge for compiler
+writers, which must choose a single model, or multiple intermediate forms, to
+use for analysis and optimization. An ideal solution would be a form of
+common representation of parallel semantics and constructs.
 
@@
 is presented in Section~\ref{Sec:OpenMP} and a more complete description can 
-be found in the OpenMP specification~\cite{spec}.
+be found in the OpenMP specification~\cite{openmp}.
 
@@
 fragmentation of parallel programming models. There are many ways to write
-parallel programs~\cite{spec, qthreads, chapel, cilk, kokkos, legion} \pmnote{I
-would add MPI and UPC/PGAS here too}. From
-libraries, to language extensions, to standalone languages, to combinations of
-the above, fragmentation of parallel programming models is a problem that is
-only getting worse. This means that when writing tooling, optimizations, or
-analyses, one has to choose which model to target as well as understand points
-of contention and interoperability between them. A common difficulty among
-all of these is that reasoning about parallel programs is hard and compilers
-can potentially help reduce this burden. This is exactly analogous the problem 
-LLVM solves: compiler optimizations are hard, so sharing them is valuable. Tapir 
-is a demonstration that the the LLVM approach to compilation of serial code can 
-be effectively extended to help reason about parallel code.
+parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc, mpi}
+\pmnote{I would add MPI and UPC/PGAS here too}. From libraries, to language
+extensions, to standalone languages, to combinations of the above,
+fragmentation of parallel programming models is a problem that is only getting
+worse. This means that when writing tooling, optimizations, or analyses, one
+has to choose which model to target as well as understand points of contention
+and interoperability between them. A common difficulty among all of these is
+that reasoning about parallel programs is hard and compilers can potentially
+help reduce this burden. This is exactly analogous the problem LLVM solves:
+compiler optimizations are hard, so sharing them is valuable. Tapir is a
+demonstration that the the LLVM approach to compilation of serial code can be
+effectively extended to help reason about parallel code.
 
@@
 
-Initially, the cross-vendor OpenMP~\cite{spec} shared memory programming model 
+Initially, the cross-vendor OpenMP~\cite{openmp} shared memory programming model 
 focused on the execution of data parallelism by a cooperating team of threads, 
@@
 
+A seperate but related area in LLVM extensions is the area of using memory
+
 \subsection{Related Work} \label{Sec:Related}

### 2017-09-08 paper.tex
Hacking away...

--- 2017-09-08 paper.tex
@@
 
-Consider the first code segment defined in \figref{search}.
+Consider the first code segment defined in~\figref{search}.
 Here, we find a simple OpenMP program which performs a divide-and-conquer search.
@@
 namely, that the value \code{(low+high)/2} may be computed once before either
-parallel task, allowing the program to avoid some redundant computation. On
+parallel task, avoiding redundant computations. For
 the serial version of this program, optimizations like this are performed
@@
 
-If one were to compile this program using the traditional technique of converting
-to runtime calls, one would find a program similar to \figref{runtime_calls}.
-The OpenMP pragmas are effectively treated as syntactic sugar for such runtime calls
-because vanilla LLVM has no concise way to represent parallel programs.
-These runtime calls obfuscate the program, making it infeasable to analyze or
-optimize parallel code.
+If one were to compile this program using the traditional technique of lowering 
+\texttt{\#pragma} directives into runtime calls, one would find a program similar 
+to~\figref{runtime_calls}.
+The OpenMP pragmas are effectively treated as syntactic sugar for runtime calls
+because LLVM has no concise way to represent such parallel constructs.
+Importantly, these runtime calls tend to obfuscate the program, making it 
+infeasible to analyze and/or optimize parallel code.
 
@@
 
-\caption[Simplified compilation of the unoptimized search code from \figref{search}.]{Simplified compilation of the unoptimized search code from \figref{search}.
-\subfigcap{a}~The function \code{search}.
-The parallel tasks are moved into their own functions and accompanying closures.
-These functions are then passed to OpenMP runtime calls.
-This system obfuscates the program, making it infeasable for an optimizer to reason about what is happening.}
+\caption[Simplified compilation of the unoptimized search code from \figref{search}.]
+{Simplified compilation of the unoptimized search code from~\figref{search}.  
+\subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own 
+functions and accompanying closures. These functions are then passed to OpenMP runtime 
+calls. This system obfuscates the program, making it infeasable for an optimizer to 
+reason about what is happening.}
   \label{fig:runtime_calls}
@@
 Tapir is a compelling solution to this problem because it allows existing
-analysis and optimizations to work on parallel programs without much modification.
+analysis and optimizations to work on parallel programs without significant modifications.
 To properly describe Tapir, one must briefly introduce key aspects of LLVM.
-LLVM represents control flow through a control-flow graph (CFG) of blocks,
-which are sequences of instructions. Blocks always end with terminator instructions
-which define the edges of the control flow graph. For example, \figref{CFG} shows a simple
+LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
+sequences of instructions. Blocks always end with terminator instructions that
+define the edges of the control flow graph. For example, \figref{CFG} shows a simple
 function and its corresponding LLVM code.
@@
 \textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
-is analagous to a \code{fork} in most fork-join frameworks.
+is analagous to a \code{fork} in most fork-join models.
 
@@
 designated by the application developer and scheduled onto the team of threads 
-by the OpenMP run time implementation.  The \texttt{task} construct applied to 
+by the OpenMP runtime implementation.  The \texttt{task} construct applied to 
 a structured block of code creates an explicit task, and the \texttt{taskwait} 
@@
 clause codifies data dependences among tasks, indicating that a data location 
-is an input or output of a task.  The run time system ensures that a task is 
+is an input or output of a task.  The runtime system ensures that a task is 
 not scheduled until its input dependences are fulfilled.  The \texttt{taskloop} 
@@
 To better understand how the OpenMP to Tapir compiler works, we turn to the
-\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:Example}
+\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:tapir-example}
 we show what Tapir code is generated.
@@
 \end{tabular}
-
-\caption{Compilation to Tapir}
-\label{fig:Example}
+\caption{Transformation from OpenMP task constructs to Tapir.}
+\label{fig:tapir-example}
 \end{figure*}
@@
 actually a nice example of code that will be optimized by Tapir. The LLVM code
-shown in Figure~\ref{Fig:Example} is not optimized. With optimizataions turned
+shown in Figure~\ref{Fig:tapir-example} is not optimized. With optimizataions turned
 on, Tapir will replace a \texttt{detach} of a function call followed

### 2017-09-08 paper.tex
Small results discussion change

--- 2017-09-08 paper.tex
@@
 
-As mentioned earlier on \texttt{Floorplan} our implementation was sometimes
-giving incorrect results. It is difficult to tell, but the time to completion
-is non-zero. In fact, while ICC and Clang finished in roughly 1-2 seconds, the
-our Tapir implementation was finishing in 0.02 seconds, and computing the
-correct result roughly half the time. This raises interesting questions on the
-cost/benefit relation of the OpenMP behaviour that our implementation is
-missing. We'll discuss this further in Section~\ref{Sec:Discussion}.
+As mentioned earlier on \texttt{Floorplan} our implementation returned correct
+results nondeterministically due to not having an implementation of the OpenMP
+\texttt{critical} and \texttt{atomic} pragmas. While ICC and Clang finished in
+roughly 1-2 seconds, the Tapir implementation was finishing in 0.02 seconds,
+and computing the correct result roughly half the time. This raises interesting
+questions on the cost/benefit relation of the missing OpenMP pragmas.
 
@@
 GCC implementation was timing out at 600 seconds, so was at least 100 times
-slower. We've left GCC results off of the graphs for the three failure cases, 
+slower. We've left GCC results off of the graphs for the three timeout cases,
 \texttt{FFT}, \texttt{Fibonacci}, and \texttt{NQueens}.

### 2017-09-08 paper.tex
Fixed example figure reference

@@
 intermediate representations \textbf{(IRs)} which allow such transformations
-to be written once with relative ease and apply to a variety of source front-end
-languages such as C or C++.
-Historically, the IR of mainstream compilers such as LLVM\cite{llvm} or GCC's Gimple\cite{}
-haven't suppored parallel constructs. As a consequence, compilers utilizing these
-IRs haven't had the ability to reason about parallelism. The recent work of
-Schardl et al.~\cite{tapir} has shown that the LLVM IR can be
-extended to represent fork-join parallelism without requiring a major rewrite.
-They did so by extending the LLVM instruction set with three instructions
-capable of representing fork-join parallelism. To benchmark performance and test
-accuracy, they wrote a frontend which translated programs written with Cilk, a parallel
-programming model for C/C++, into the Tapir IR. They also implemented a backend to \textbf{lower}
-Tapir programs to vanilla LLVM IR with embedded Cilk runtime calls.
-They suggested that a similar approach could be taken with OpenMP and other
-such frameworks.
+to be written once with relative ease and apply to a variety of source
+front-end languages such as C or C++.  Historically, the IR of mainstream
+compilers such as LLVM~\cite{llvm} or GCC's Gimple~\cite{} haven't suppored
+parallel constructs. As a consequence, compilers utilizing these IRs haven't
+had the ability to reason about parallelism. The recent work of Schardl et
+al.~\cite{tapir} has shown that the LLVM IR can be extended to represent
+fork-join parallelism without requiring a major rewrite.  They did so by
+extending the LLVM instruction set with three instructions capable of
+representing fork-join parallelism. To benchmark performance and test accuracy,
+they wrote a frontend which translated programs written with Cilk, a parallel
+programming model for C/C++, into the Tapir IR. They also implemented a backend
+to \textbf{lower} Tapir programs to vanilla LLVM IR with embedded Cilk runtime
+calls. They suggested that a similar approach could be taken with OpenMP and
+other such frameworks.
 
@@
 To better understand how the OpenMP to Tapir compiler works, we turn to the
-\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:Example}
+\texttt{fib} example from Section~\ref{Sec:OpenMP}. In Figure~\ref{Fig:Example}
 we show what Tapir code is generated.
@@
 \caption{Compilation to Tapir}
-\label{fig:Example}
+\label{Fig:Example}
 \end{figure*}

### 2017-09-08 paper.tex
Figure update.

--- 2017-09-08 paper.tex
@@
 To better understand how the OpenMP to Tapir compiler works, we turn to the
-\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:tapir-example}
+\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{tapir-example}
 we show what Tapir code is generated.
@@
 actually a nice example of code that will be optimized by Tapir. The LLVM code
-shown in Figure~\ref{Fig:tapir-example} is not optimized. With optimizataions turned
+shown in Figure~\ref{tapir-example} is not optimized. With optimizataions turned
 on, Tapir will replace a \texttt{detach} of a function call followed

### 2017-09-08 paper.tex
More updates.

@@
 \texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{tapir-example}
-we show what Tapir code is generated.
+we show the input OpenMP program and a portion of the corresponding Tapir code 
+that is generated.
 
@@
 
-The listed Tapir LLVM IR code corresponds to lines 7 through 12 of the C source
+The listed Tapir IR code corresponds to lines 7 through 12 of the C source
 code.  Upon entering the \texttt{else} branch, the code immediately detaches
@@
 actually a nice example of code that will be optimized by Tapir. The LLVM code
-shown in Figure~\ref{tapir-example} is not optimized. With optimizataions turned
+shown in \figref{tapir-example} is not optimized. With optimizataions turned
 on, Tapir will replace a \texttt{detach} of a function call followed
@@
 
-For evaluation we compared performance on the Barcelona OpenMP task suite
-\cite{barcelona} to existing OpenMP implementations. The Barcelona OpenMP task
-suite is a set of tests intended to test performance of OpenMP implementations
-using both irregular and regular tasking. The benchmark suite has added an
-unbalanced tree search benchmark since the paper was published. Unfortunately
-we were unable to succesfully run it on any implementation, even for even
-medium input sizes. We have therefore left it out of our evaluation. The
-following descriptions are abridged versions taken verbatim from
-\cite{barcelona}:
+For evaluating the use of Tapir we compared performance on the Barcelona OpenMP Task 
+Suite~\cite{barcelona} to several existing OpenMP implementations. The suite is a set 
+of tests intended to test performance of OpenMP implementations
+using both irregular and regular tasking.\footnote{An unbalanced tree search benchmark was
+very recently added to the suite. Unfortunately we were unable to succesfully run 
+it on any implementation, even for even medium input sizes. We have therefore left 
+it out of our evaluation.  We will continue to work on this benchmark and will 
+include the results if the paper is accepted to the workshop.} The following list provides abridged descriptions of 
+each benchmark -- they are taken verbatim from~\cite{barcelona}:
 

### 2017-09-08 paper.tex
citations and modified example

@@
 front-end languages such as C or C++.  Historically, the IR of mainstream
-compilers such as LLVM~\cite{llvm} or GCC's Gimple~\cite{} haven't suppored
-parallel constructs. As a consequence, compilers utilizing these IRs haven't
-had the ability to reason about parallelism. The recent work of Schardl et
-al.~\cite{tapir} has shown that the LLVM IR can be extended to represent
-fork-join parallelism without requiring a major rewrite.  They did so by
-extending the LLVM instruction set with three instructions capable of
-representing fork-join parallelism. To benchmark performance and test accuracy,
-they wrote a frontend which translated programs written with Cilk, a parallel
-programming model for C/C++, into the Tapir IR. They also implemented a backend
-to \textbf{lower} Tapir programs to vanilla LLVM IR with embedded Cilk runtime
-calls. They suggested that a similar approach could be taken with OpenMP and
-other such frameworks.
+compilers such as LLVM~\cite{lattner2004llvm} or GCC's
+Gimple~\cite{merrill2003generic} haven't suppored parallel constructs. As a
+consequence, compilers utilizing these IRs haven't had the ability to reason
+about parallelism. The recent work of Schardl et al.~\cite{tapir} has shown
+that the LLVM IR can be extended to represent fork-join parallelism without
+requiring a major rewrite.  They did so by extending the LLVM instruction set
+with three instructions capable of representing fork-join parallelism. To
+benchmark performance and test accuracy, they wrote a frontend which translated
+programs written with Cilk, a parallel programming model for C/C++, into the
+Tapir IR. They also implemented a backend to \textbf{lower} Tapir programs to
+vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
+approach could be taken with OpenMP and other such frameworks.
 
@@
 fragmentation of parallel programming models. There are many ways to write
-parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc, mpi}
-\pmnote{I would add MPI and UPC/PGAS here too}. From libraries, to language
-extensions, to standalone languages, to combinations of the above,
-fragmentation of parallel programming models is a problem that is only getting
-worse. This means that when writing tooling, optimizations, or analyses, one
-has to choose which model to target as well as understand points of contention
-and interoperability between them. A common difficulty among all of these is
-that reasoning about parallel programs is hard and compilers can potentially
-help reduce this burden. This is exactly analogous the problem LLVM solves:
-compiler optimizations are hard, so sharing them is valuable. Tapir is a
-demonstration that the the LLVM approach to compilation of serial code can be
-effectively extended to help reason about parallel code.
+parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
+mpi} . From libraries, to language extensions, to standalone languages, to
+combinations of the above, fragmentation of parallel programming models is a
+problem that is only getting worse. This means that when writing tooling,
+optimizations, or analyses, one has to choose which model to target as well as
+understand points of contention and interoperability between them. A common
+difficulty among all of these is that reasoning about parallel programs is hard
+and compilers can potentially help reduce this burden. This is exactly
+analogous the problem LLVM solves: compiler optimizations are hard, so sharing
+them is valuable. Tapir is a demonstration that the the LLVM approach to
+compilation of serial code can be effectively extended to help reason about
+parallel code.
 
@@
 To better understand how the OpenMP to Tapir compiler works, we turn to the
-\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{tapir-example}
-we show the input OpenMP program and a portion of the corresponding Tapir code 
-that is generated.
+parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
+Figure~\ref{Fig:Example} we show what the input OpenMP code and the Tapir code
+is generated.
 
@@
 \begin{lstlisting}
-int fib(int n) 
-{                    // A
-  if (n < 2)
-    return n;
-  else
-  {
-    int x, y;
   #pragma omp task
-      x = fib(n-1);  // B
+    x = fib(n-1);  
   #pragma omp task
-      y = fib(n-2);  // C
+    y = fib(n-2);  
   #pragma omp taskwait
-    return (x+y);    // D
-  }
-}
 \end{lstlisting}
@@
 
-The listed Tapir IR code corresponds to lines 7 through 12 of the C source
-code.  Upon entering the \texttt{else} branch, the code immediately detaches
-the basic block corresponding to the call to \texttt{fib(n-1)}, labeled
+Upon entering the \texttt{else} branch, the code immediately detaches the basic
+block corresponding to the call to \texttt{fib(n-1)}, labeled
 \texttt{det.achd}. The continuation for the first call also immediately
@@
 the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
-and the code leaves little room for optimization, algorithmic changes
-notwithstanding. In other words, it is effectively a benchmark of how fast a
-runtime can execute fork-join parallelism. \gsnote{still worth looking into the
-optimization I mention in the implemtnation section.}
+and the code leaves little room for optimization (algorithmic rewrites
+notwithstanding). In other words, it is effectively a benchmark of how fast a
+runtime can execute fork-join parallelism .
 
@@
 
-A seperate but related area in LLVM extensions is the area of using memory
+One interesting area for future work would be combining Tapir with work on
+reasoning about locality in LLVM. For example, Hayashi et al.
+\cite{hayashi2015llvm} use address spaces in LLVM to reason about memory
+locality in partitioned global address space (PGAS) systems. By combining
+reasoning about code concurrency and locality, one could potentially apply
+optimizations to better schedule and co-locate code and data.
 

### 2017-09-08 paper.tex
More tweaks.

--- 2017-09-08 paper.tex
@@
 generally outperforming or matching the best of existing implementations. For
-benchmarks where the overhead to work ratio is high is where the Tapir
-implementation really shines. For example, for the \texttt{Fibonacci}
+benchmarks where the overhead-to-work ratio is high is where the Tapir (and 
+Cilk) implementation shows significant benefit. For example, for the \texttt{Fibonacci}
 benchmark, our implementation finishes significantly faster than any of the
 others. While ICC failed for that sized input, we did test with smaller inputs
-and found its similarly outpaced by the Tapir implementation.  \texttt{NQueens}
+and found it was similarly outpaced by the Tapir implementation.  \texttt{NQueens}
 and \texttt{FFT} show similar behaviour, to lesser degrees. In contrast,
@@
 
-In this section we attempt to understand \emph{why} the performance varies in
-the ways it does. While we can't hope to figure out every discrepancy in
-performance, we can hope to get some ideas for why the performance of our
-implementation is generally better, and where each of the implementations is
-spending its time.
-
-Our primary tool for this task will be performance counters. It's worth noting
-that some information is difficult to infer from performance counters, such as
-sources of contention in runtimes, reasons for memory locality issues, etc.
-Still, it will give us some insight into performance bottlenecks.
+In this section we speculate \emph{why} the performance varies in
+the ways it does. While we have yet to figure out every discrepancy in
+performance, we hope to get some ideas for why the performance of our
+implementation is generally better and where each of the implementations are
+spending a majority of the time. Our primary tool for this task will be performance 
+counters. It's worth noting that some information is difficult to infer from 
+performance counters, such as sources of contention, reasons for memory locality 
+issues, etc. Still, they do give us some insight into performance bottlenecks.
 
@@
 gains due to the existing Tapir backend and Cilk runtime we sadly cannot claim
-credit for.
-
-We can break reasons for performance discrepancies into a few categories:
+credit for. We can break reasons for performance discrepancies into a few categories:
 
@@
 importance of optimizations of parallel code is clear, we hypothesize that most
-of the disparity present in these results is a function of runtime
+of the disparity present in these results is more a function of runtime
 implementation differences. As justification for this hypothesis, consider
 the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
-and the code leaves little room for optimization, algorithmic changes
+and the code leaves little room for optimization --- algorithmic changes
 notwithstanding. In other words, it is effectively a benchmark of how fast a
@@
 This supports our hypothesis that most of the slowdown incurred is overhead in
-the runtime. For \texttt{tapir}, the overhead is relatively small given the
+the runtime. For Tapir the overhead is relatively small given the
 size of the work chunks, for \texttt{clang} and \texttt{icc} it is
-significantly larger, while for \texttt{gcc} it is completely overwhelming.
-
-For the cases where performance is close between implementations, like
+significantly larger, and for \texttt{gcc} it is overwhelming.
+
+For the cases where performance is close between the implementations, like
 \texttt{sparselu}, we expect runtimes to be dominated by actual task work. This 

### 2017-09-08 paper.tex
re-fixed example, added lithe

--- 2017-09-08 paper.tex
@@
 \begin{abstract}
-\pmnote{Overhauled this some -- take a look...}
 Optimizing compilers for task-level parallelism are still in their infancy.
@@
 an extension to LLVM IR that represents fork-join parallelism. This enables
-analyses and optimizations that were previously inaccessible to OpenMP codes, as well as
-the ability to target additional runtimes at code generation. Using a Cilk runtime 
-backend, we compare results to existing OpenMP implementations. Initial performance 
-results for the Barcelona OpenMP tasking suite and show performance improvements over 
-these existing implementations.
+analyses and optimizations that were previously inaccessible to OpenMP codes,
+as well as the ability to target additional runtimes at code generation. Using
+a Cilk runtime backend, we compare results to existing OpenMP implementations.
+Initial performance results for the Barcelona OpenMP tasking suite and show
+performance improvements over these existing implementations.
 \end{abstract}
@@
 interoperate either at the level of compile-time analysis or at run-time
-execution.  \wmnote{Not sure if relevant, but there was some experiment run a
-while back of someone composing a library that internally used OpenMP with
-another framework that had terribly results compared with either used alone.}
-\gsnote{Hmm. A quick googling turned up a project out of berkeley combining
-OpenMP and TBB, but performance didn't seem too bad. I'll take another look
-tomorrow.} In addition, many of these programming models are implemented as
+execution. In addition, many of these programming models are implemented as
 runtime libraries with only primitive compiler support. As a result, most are
@@
 common representation of parallel semantics and constructs.
+
+%\wmnote{Not sure if relevant, but there was some experiment run a
+%while back of someone composing a library that internally used OpenMP with
+%another framework that had terribly results compared with either used alone.}
+%\gsnote{I've added a quick note on Lithe, an MIT/Berkeley project that attempted
+% combine OpenMP and TBB, in the related work section}
 
@@
 systems. These calls are infeasible for the compiler to reason about, due to
-the complexity and flexibility of the runtimes being called \pmnote{and the challenge of
-reasoning about the corresponding semantics in a uniform fashion?}. The addition of parallel
-constructs were recently explored by Schardl et al.~\cite{tapir}. By adding three first-class 
-instructions to LLVM, Tapir enables analyses and optimizations of parallel constructs. We briefly 
-describe the key aspects of Tapir in Section~\ref{Sec:Tapir}.
-% I removed this as the reference was already given...   PM 
-% For a more complete description see \cite{tapir}.
+the complexity and flexibility of the runtimes being called, and the challenge
+of reasoning about the corresponding semantics in a uniform fashion. The
+addition of parallel constructs were recently explored by Schardl et
+al.~\cite{tapir}. By adding three first-class instructions to LLVM, Tapir
+enables analyses and optimizations of parallel constructs. We briefly describe
+the key aspects of Tapir in Section~\ref{Sec:Tapir}.
+% I removed this as the reference was already given...   PM For a more complete
+% description see \cite{tapir}.
 
@@
 parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
-Figure~\ref{Fig:Example} we show what the input OpenMP code and the Tapir code
+Figure~\ref{fig:tapir-example} we show what the input OpenMP code and the Tapir code
 is generated.
@@
 implementations listed in Section~\ref{Sec:Evaluation}. See
-Figures~\ref{Fig:results} and \ref{Fig:results2} for a simple visualization of
+Figures~\ref{fig:results} and \ref{fig:results2} for a simple visualization of
 the performance. Each graph represents a single benchmark from the previously
@@
 \caption{Barcelona OpenMP Task Suite Results}
-\label{Fig:results}
+\label{fig:results}
 \end{figure*}
@@
 \caption{Barcelona OpenMP Task Suite Results (continued)}
-\label{Fig:results2}
+\label{fig:results2}
 \end{figure*}
@@
 
+A related approach to composing programming models is to keep the same code
+generation approach, but replace calls into built in runtimes with calls into a
+shared runtime. This is the approach taken by Lithe \cite{lithe}, where Pan et
+al. compose programs using both GNU OpenMP and Intel TBB. This approach succeeds
+in composing different programming models at run-time, but doesn't enable the
+static analyses that enabled by Tapir.
+
 \section{Conclusion} \label{Sec:Conclusion}

### 2017-09-08 paper.tex
More references
Removed addressed note

@@
 
-Addressing these concerns this is a major motivation for our use of Tapir.
-\pmnote{This made me remember that Rice and Cray folks did some work on extending LLVM
-to reason about PGAS-like constructs We should probably call that out in 
-this discussion -- that said, I don't recall their approach being general so it 
-does not quite fit the discussion here but I'd have to think a bit harder if it
-would be possible to capture something like MPI operations in some fashion... 
-I will email George a pointer to the paper...} 
-By standardising an IR, in addition to easing the development of optimizations,
+Addressing these concerns this is a major motivation for our use of Tapir.  By
+standardising an IR, in addition to easing the development of optimizations,
 one avoids duplicating work across different implementations. It also enables
@@
 First, compilers can attach metadata to an existing IR. For example, LLVM has a
-parallel loop metadata construct \cite{}. This has the benefit of being
+parallel loop metadata construct \cite{llvmref}. This has the benefit of being
 flexible and requiring minimal work, but historically can be lost during

### 2017-09-08 paper.tex
Rework Figure 5 to be consistent with other code figures
Unnecessary capitalization
More clean up suggestions.

--- 2017-09-08 paper.tex
@@
 For the cases where performance is close between the implementations, like
-\texttt{sparselu}, we expect runtimes to be dominated by actual task work. This 
-expectation is confirmed by investigating performance counters for \texttt{sparselu}:
-every implementation spends roughly 90\% of it's time doing work in the task
+\texttt{sparselu}, we expect runtimes to be dominated by actual task workloads. This 
+expectation is confirmed by investigating performance counters for \texttt{sparselu}
+where every implementation spends roughly 90\% of it's time in the task
 bodies.
@@
 
-In this section we discuss the relevance of the results and this work in general
-in the context of the literature. As discussed in Section~\ref{Sec:Introduction}, 
+In this section we discuss the relevance of the results and in the context of the
+literature. As discussed in Section~\ref{Sec:Introduction}, 
 there have been many programming models for implementing parallel programs. While
 this paper does not take a stance on which one \emph{should} be used, it does
-accept that OpenMP is currently a widely used model, potentially the most
-widely used model that has a notion of tasks.
-
-The goal of this work is to show that a single intermediate representation can be
-used to implement multiple programming models, with good performance. Schardl et 
-al. showed that Tapir works well as an IR for Cilk, and surmised that it could
+accept that OpenMP is currently/potentially the most widely used model that has 
+a notion of tasks. \pmnote{but it isn't clear to me that tasks are really being 
+used much! :-)}.  The goal of this work is to show that a single intermediate representation can be
+used to implement multiple programming paradigms with good performance. Schardl et 
+al. showed that Tapir works well as an IR target for Cilk and surmised that it could
 be an effective IR for a large subset of OpenMP \cite{tapir}. This work has
-taken the first step in that direction, showing that it can work well as an IR
-for OpenMP tasks.
+taken the first step in that direction.
 
 One interesting aspect of this work is that it compiles a language extension intended
-for one runtime (OpenMP runtimes) to use another (the Cilk runtime). While in many
-ways this is a weakness of this work, it does open the possibility of doing this 
-in a general way. Having a real IR to work with enables, in the same way that LLVM
-does for serial code, the possibility of multiple input parallel languages and
-multiple output parallel runtimes. While the OpenMP spec's requirement to
-have access to runtime routines complicates things, for many programming models
+for one runtime (OpenMP) to use another (the Cilk runtime). While in many
+ways this is a weakness of this work, it does open the possibility of utilizing this 
+in a more general way to address many aspects including significant challenges 
+such as performance portability. Having a parallel-aware IR to work with enables, 
+in the same way that LLVM does for serial code, the possibility of multiple high-level 
+parallel languages and multiple parallel runtime targets. While the OpenMP specification's 
+requirement to have access to runtime routines complicates things, for many programming models
 it seems at least conceivable to map them onto multiple different runtimes. This
@@
 semantics borrowed from the Cilk codegen and their relation to the OpenMP
-specification of semantics. In particular, \emph{OpenMP doesn't specify any
+specification of semantics. In particular, \emph{OpenMP doesn't specify a
 memory model for shared variables}. This results in the Cilk style of only
@@
 potential for performance discrepancy due to differing implementation-defined
-behaviour, and we will return to it when discussing performance on the
+behaviour and we will return to it when discussing performance on the
 Barcelona OpenMP task suite. There is of course the fact that technically, by

### 2017-09-08 paper.tex

@@
 
+%\wmnote{Not sure if relevant, but there was some experiment run a
+%while back of someone composing a library that internally used OpenMP with
+%another framework that had terribly results compared with either used alone.}
+%\gsnote{I've added a quick note on Lithe, an MIT/Berkeley project that attempted
+% combine OpenMP and TBB, in the related work section}
+
 Compilers implement the majority of their analyses and optimizations on
@@
 To better understand how the OpenMP to Tapir compiler works, we turn to the
-\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{tapir-example}
-we show the input OpenMP program and a portion of the corresponding Tapir code 
-that is generated.
-
-\begin{figure*}
-\begin{tabular}{| c c c | c c c |}
-\hline
+parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
+Figure~\ref{fig:tapir-example} we show what the input OpenMP code and the Tapir code
+is generated.
+
+\begin{figure*}[h!]
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}ll}
+\subfiglabel{a} & \subfiglabel{b} \\
+\begin{minipage}[T]{0.45\linewidth}
+\begin{lstlisting}
+  #pragma omp task
+    x = fib(n-1);  
+  #pragma omp task
+    y = fib(n-2);  
+  #pragma omp taskwait
+\end{lstlisting}
+\end{minipage}
 &
-\begin{lstlisting}
-int fib(int n) 
-{                    // A
-  if (n < 2)
-    return n;
-  else
-  {
-    int x, y;
-    #pragma omp task
-      x = fib(n-1);  // B
-    #pragma omp task
-      y = fib(n-2);  // C
-    #pragma omp taskwait
-    return (x+y);    // D
-  }
-}
-\end{lstlisting}
-& & &
+\begin{minipage}[T]{0.45\linewidth}
 \begin{lstlisting}[language=llvm]
@@
 \end{lstlisting}
- & \\
-\hline
-\end{tabular}
+\end{minipage}\\
+\addlinespace[2ex]
+\bottomrule
+\end{tabular*}
 \caption{Transformation from OpenMP task constructs to Tapir.}
@@
 \end{multicols}
-\caption{Barcelona OpenMP Task Suite Results}
-\label{Fig:results}
+\caption{Barcelona OpenMP Task Suite results}
+\label{fig:results}
 \end{figure*}
@@
 \includegraphics[width=0.45\linewidth]{strassen.pdf}
-\caption{Barcelona OpenMP Task Suite Results (continued)}
-\label{Fig:results2}
+\caption{Barcelona OpenMP Task Suite results (continued)}
+\label{fig:results2}
 \end{figure*}
@@
 the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
-and the code leaves little room for optimization --- algorithmic changes
-notwithstanding. In other words, it is effectively a benchmark of how fast a
-runtime can execute fork-join parallelism. \gsnote{still worth looking into the
-optimization I mention in the implemtnation section.}
+and the code leaves little room for optimization (algorithmic rewrites
+notwithstanding). In other words, it is effectively a benchmark of how fast a
+runtime can execute fork-join parallelism.
 

### 2017-09-08 paper.tex
Won't worry about openmp task popularity for now
rather shallow editing

@@
 \subtitle{Implementing OpenMP tasks with Tapir}
+
 \author{George Stelle}
@@
 Optimizing compilers for task-level parallelism are still in their infancy.
-This work explores a frontend that compiles OpenMP tasking semantics to Tapir,
+This work explores a compiler front end that translates OpenMP tasking semantics to Tapir,
 an extension to LLVM IR that represents fork-join parallelism. This enables
@@
 as well as the ability to target additional runtimes at code generation. Using
-a Cilk runtime backend, we compare results to existing OpenMP implementations.
-Initial performance results for the Barcelona OpenMP tasking suite and show
-performance improvements over these existing implementations.
+a Cilk runtime back end, we compare results to existing OpenMP implementations.
+Initial performance results for the Barcelona OpenMP tasking suite show
+performance improvements over existing implementations.
 \end{abstract}
@@
 
-When writing task-parallel programs today, one has a large selection of
+When writing task-parallel programs today, there is a large selection of
 potential programming models and implementations to consider. Unfortunately, 
@@
 pthreads\cite{pthreads}, MPI~\cite{mpi}, Chapel~\cite{chapel}, UPC~\cite{upc},
-OpenCL~\cite{opencl}, OpenACC~\cite{openacc} among others have limited ability to
+OpenCL~\cite{opencl}, and OpenACC~\cite{openacc}, among others, have limited ability to
 interoperate either at the level of compile-time analysis or at run-time
 execution. In addition, many of these programming models are implemented as
-runtime libraries with only primitive compiler support. As a result, most are
-not able to take advantage of the compiler's analysis and optimization
+runtime libraries with only primitive compiler support. As a consequence, most are
+not able to take advantage of potential compiler analysis and optimization
 capabilities.  While there has been some recent work on specializing
@@
 fragmentation among models presents a significant challenge for compiler
-writers, which must choose a single model, or multiple intermediate forms, to
-use for analysis and optimization. An ideal solution would be a form of
+writers, which must choose a single model, or multiple intermediate forms, for
+analysis and optimization. An ideal solution would be a 
 common representation of parallel semantics and constructs.
@@
 Gimple~\cite{merrill2003generic} haven't suppored parallel constructs. As a
-consequence, compilers utilizing these IRs haven't had the ability to reason
-about parallelism. The recent work of Schardl et al.~\cite{tapir} has shown
+consequence, compilers using these IRs haven't had the ability to reason
+about parallelism. The recent work of Schardl et al.\ has shown
 that the LLVM IR can be extended to represent fork-join parallelism without
-requiring a major rewrite.  They did so by extending the LLVM instruction set
+requiring a major rewrite~\cite{tapir}.  They did so by extending the LLVM instruction set
 with three instructions capable of representing fork-join parallelism. To
-benchmark performance and test accuracy, they wrote a frontend which translated
+benchmark performance and test accuracy they wrote a front end which translated
 programs written with Cilk, a parallel programming model for C/C++, into the
-Tapir IR. They also implemented a backend to \textbf{lower} Tapir programs to
+Tapir IR. They also implemented a back end to \textbf{lower} Tapir programs to
 vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
@@
 
-In this work, we put that suggestion to the test, implementing OpenMP tasks by
+In this work we put that suggestion to the test, implementing OpenMP tasks by
 compiling them to Tapir instructions. This allows us to both analyze and optimize
 OpenMP programs as well as compile programs written in OpenMP to use other runtimes.
-For this paper, our prototype implementation uses the Cilk runtime as a backend,
+For this paper our prototype implementation uses the Cilk runtime as a back end,
 and we run our prototype implementation on the Barcelona OpenMP task suite~\cite{barcelona}. 
-We discuss the kinds of optimizations this enables, and how
-the work can be extended to include other parallelism constructs and semantics
+We discuss the kinds of optimizations this enables and how
+the work can be extended to include other parallel constructs and semantics
 in OpenMP. We also discuss the possibility of extending this work to other
-programming models, which would help to alleviate some of the fragmentation
-issues discussed above.
-
-The contributions of this work include:
-
+programming models, which would help to alleviate some of the problems with fragmentation
+previously discussed.
+
+The contributions of this work include
 \begin{itemize}
-  \item A prototype frontend that transforms OpenMP task constructs to Tapir IR.
-  \item Benchmarks of the implementation using a Cilk runtime backend, showing
-        performance improvements over existing OpenMP implementations.
+  \item A prototype front end that transforms OpenMP task constructs to Tapir IR;
+  \item Benchmarks of the implementation using a Cilk runtime back end, showing
+        performance improvements over existing OpenMP implementations; and
   \item An initial investigation into the reasons for these performance improvements.
 \end{itemize}
-
-Our hope for this paper is to move the community towards further conversations on a 
-shared IR that can be used for many programming models. The work presented in this paper only
+Our hope is to move the community towards further conversations on a 
+shared IR that can be used for many programming models. The work presented here only
 represents a small step in that direction and there is still a herculean amount
@@
 
-The remainder of the paper is organized as follows: In Section~\ref{Sec:Background} 
-we give background and motivation for the paper. Following that, we describe OpenMP
+The remainder of the paper is organized as follows.  In Section~\ref{Sec:Background} 
+we give background and motivation for the presented work. Following that, we describe OpenMP
 tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
@@
 Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
-evaluation setup, including what implementations we compare to and how. We discuss
+evaluation setup, including which implementations we compare to and how. We discuss
 the results of evaluation in Section~\ref{Sec:Results}, including possible
-explanations for discrepancies. We discuss the implications of the results ans the
+explanations for discrepancies. We discuss the implications of the results and the
 significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
-we cover some of the many ways the work can be extended in
+we cover some of the many ways the work could be extended in
 Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.
@@
 of concurrent or parallel processes. This has resulted in existing compilers
-treating parallelism constructs as thin wrappers for calls into runtime
-systems. These calls are infeasible for the compiler to reason about, due to
+treating parallel constructs as thin wrappers for calls into run-time
+systems. These calls are infeasible for the compiler to reason about because of
 the complexity and flexibility of the runtimes being called, and the challenge
@@
 
-Arguably the most ubiquitous parallel programming model for on-node parallelism
-in HPC, OpenMP has grown from being a way to implement parallel loops into a
+Arguably the most common parallel programming model for on-node parallelism
+in HPC, OpenMP has grown from being a method for implementing parallel loops into a
 large and complex specification for parallelism. This paper focuses on a 
@@
 parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
-mpi} . From libraries, to language extensions, to standalone languages, to
-combinations of the above, fragmentation of parallel programming models is a
+mpi}.  From libraries, to language extensions, to standalone languages, to
+combinations of the these, the fragmentation of parallel programming models is a
 problem that is only getting worse. This means that when writing tooling,
-optimizations, or analyses, one has to choose which model to target as well as
+optimizations, or analyses, one much choose which model to target as well as
 understand points of contention and interoperability between them. A common
@@
 
-With fragmentation comes questions of compatibility, or composability, of 
-different programming models. To illustrate this issue, consider the example of
+With fragmentation comes questions of compatibility or composability of 
+different programming models. To illustrate this, consider the example of
 a program using a library that uses OpenMP to implement parallelism internally. 
-If said program uses a different programming model for parallelism, e.g. Cilk,
-there is currently no good solution for ensuring that the two different backends
+If this program uses a different programming model for parallelism, e.g.\ Cilk,
+there is currently no good solution for ensuring that the two different back ends
 will cooperate with management of hardware resources. 
-
-Addressing these concerns this is a major motivation for our use of Tapir.  By
-standardising an IR, in addition to easing the development of optimizations,
+%
+Addressing these concerns is a major motivation for our use of Tapir.  By
+standardizing an IR, in addition to easing the development of optimizations,
 one avoids duplicating work across different implementations. It also enables
-the possibility of multiple programming models coexisting peacefully. We'll
-return to this issue later in Section~\ref{Sec:Discussion} when discussing
+the possibility of multiple programming models coexisting peacefully. We will
+return to this consideration later in Section~\ref{Sec:Discussion} when discussing
 future work.
@@
 Tapir is an extension to LLVM proposed by Schardl, Moses, and Leiserson
-aiming to resolve issues regarding optimizing parallel code. Prior to
-the introduction of Tapir, compilers would represent parallel programs
+seeking to resolve issues in optimizing parallel code. Prior to
+the introduction of Tapir compilers would represent parallel programs
 by directly translating their syntax to opaque runtime calls. This allowed
-for compilers to support parallel programs, but meant that traditional 
-optimizations such as code motion \cite{} or common-subexpression elimination \cite{}
+compilers to support parallel programs but meant that traditional 
+optimizations such as code motion~\cite{} or common-subexpression elimination~\cite{}
 weren't able to reason about parallel programs. This often led to parallel
-programs running significantly slower than expected. In the Tapir project,
+programs running significantly slower than expected. In the Tapir project
 the authors show that it is possible to represent and optimize fork-join
-parallel programs with relatively minimal modifications to the compiler: namely
+parallel programs with relatively minimal modifications to the compiler, namely
 the introduction of three instructions designed to interface well with an
@@
 
-Consider the first code segment defined in~\figref{search}.
-Here, we find a simple OpenMP program which performs a divide-and-conquer search.
-However, as written, there is a simple optimization that can be performed --
-namely, that the value \code{(low+high)/2} may be computed once before either
-parallel task, avoiding redundant computations. For
-the serial version of this program, optimizations like this are performed
+Consider the first code segment defined in~\figref{search},
+a simple OpenMP program which performs divide-and-conquer search.
+However, as written, there is a simple optimization that can be performed,
+that the value \code{(low+high)/2} may be computed once before either
+parallel task, thereby avoiding redundant computations. For
+the serial version of this program optimizations like this are performed
 automatically. However, without the use of Tapir, these sorts of optimizations
@@
 analysis and optimizations to work on parallel programs without significant modifications.
-To properly describe Tapir, one must briefly introduce key aspects of LLVM.
+To properly describe Tapir requires a brief introduction to key aspects of LLVM.
 LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
@@
 
-To represent parallelism, Tapir introduces the \code{detach}, \code{reattach}, and
+To represent parallelism Tapir introduces the \code{detach}, \code{reattach}, and
 \code{sync} instructions. The \code{detach} statement is syntactically similar to
-the conditional branch as it is also a terminator instruction with with two blocks.
+a conditional branch in that it is a terminator instruction with with two blocks.
 However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
@@
 
-At the end of any task created by a \code{detach} instruction, one must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
+The end of a task created by a \code{detach} instruction must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
 is used to denote the end of a parallel task.
-
+%
 Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
@@
 
-The initial design of the OpenMP task model~\cite{ayguade09design} established 
-a basic framework for asynchronous task parallel execution in OpenMP programs.
-Subsequent versions of the OpenMP specification up to the current version 4.5 
-have added additional new features to the tasking model.  The \texttt{depend} 
+The initial design of the OpenMP task model established 
+a basic framework for asynchronous task parallel execution in OpenMP programs~\cite{ayguade09design}.
+Subsequent versions of the OpenMP specification, up to the current version 4.5, 
+have added new features to the tasking model.  The \texttt{depend} 
 clause codifies data dependences among tasks, indicating that a data location 
 is an input or output of a task.  The runtime system ensures that a task is 
-not scheduled until its input dependences are fulfilled.  The \texttt{taskloop} 
+not scheduled until its input dependencies are fulfilled.  The \texttt{taskloop} 
 construct combines groups of independent loop iterations into explicit tasks, 
@@
 work on other tasks, as may be desired for long-running tasks that generate 
-many others.  The task concept was also leveraged to provide for asynchronous 
+many others.  The task concept was also leveraged to provide asynchronous 
 offload of data and computation to accelerators by applying the \texttt{nowait} 
-clause to device constructs like the \texttt{target} construct to generate an 
+clause to device constructs such as the \texttt{target} construct to generate an 
 asynchronous \textit{target task}.
@@
 Tapir is implemented as an extension to the LLVM instruction set. Clang is a C
-family compiler that has support for OpenMP extensions \cite{clang} and targets
-LLVM. The existing Clang OpenMP implementation maps OpenMP constructs directly
-to OpenMP runtime library calls, by wrapping a C statement in a
+family compiler that has support for OpenMP extensions and targets
+LLVM~ \cite{clang}. The existing Clang OpenMP implementation maps OpenMP constructs directly
+to OpenMP runtime library calls by wrapping a C statement in a
 \texttt{CapturedStmt}. This replaces a C statement with a function call into
-the OpenMP runtime library, along with a machine generated function who's body
+the OpenMP runtime library, along with a machine generated function whose body
 contains that statement.
 
-For this work, we replaced a subset of the OpenMP implementation to generate
+For this work we replaced a subset of the OpenMP implementation to generate
 Tapir IR instead of vanilla LLVM IR with OpenMP runtime calls. Specifically, we
-replace the two primary pragmas for task parallelism: \texttt{task\_spawn} and
-\texttt{task\_wait}. By re-using code from Schardl et al. for code generation
+replace the two primary \texttt{task\_spawn} and
+\texttt{task\_wait} pragmas for task parallelism. By re-using code from Schardl et al.\ for code generation
 of Cilk constructs, we were able to easily generate Tapir code for these OpenMP
@@
 While we did implement the codegen for the \texttt{task\_spawn} and
-\texttt{task\_wait} constructs, it's worth noting here that even for these
-pragmas, the implementation is incomplete. Currently, any clauses modifying the
-behaviour of the pragmas is ignored. Probably the most common semantics this
-will change of variable semantics, e.g. \texttt{shared} vs. \texttt{private}.
+\texttt{task\_wait} constructs, it's worth noting that even for these
+pragmas the implementation is incomplete. Currently, any clauses modifying the
+behavior of the pragmas is ignored. Probably the most common semantics this
+will change is for the semantcis of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
 Surprisingly, this had little effect on the correctness of the entire Barcelona
-OpenMP task suite. Indeed, it is likely that fixing this issue would increase
-the performance of this work marginally, due to reducing the number of memory
-copies for variables declared \texttt{private} in OpenMP code. This property
+OpenMP task suite. Indeed, it is likely that fixing this problem would marginally increase
+performance by reducing the number of memory
+copies for variables declared \texttt{private} in OpenMP code.  The fact
 that behaviour wasn't changed significantly is worth revisiting, and we do so
@@
 
-This work represented in this paper is only frontend implementation, which
-means we use the only existing backend for Tapir. The existing backend
-generates code that calls into the Cilk runtime, \texttt{libcilkrts}. This has
+The work represented in this paper is only front-end implementation, which
+means we use the only existing back end for Tapir. The existing back end
+generates code that calls the Cilk runtime \texttt{libcilkrts}. This has
 implications for adhering to the OpenMP specification. For example, environment
-variables such as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by
+variables such as \texttt{OMP\_NUM\_THREADS} are ignored and replaced by
 \texttt{CILK\_NWORKERS}.  We discuss some of these issues below, and return to
-address others in the future work section.
+address others in the discussion of future work.
 
 There are other issues that had to be overcome to run full OpenMP programs
-using Tapir as described above. Generally, in addition the the \texttt{task} 
-and \texttt{taskwait} pragmas, a program has to contain initialization OpenMP
-pragmas in order to run parallel code. A standard pattern for building
+using Tapir. Generally, in addition the the \texttt{task} 
+and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP
+%pragmas in order to run parallel code.
+A standard pattern for building
 task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
@@
 pragma to have only one thread continue on the specified statement. Then the
-other threads will get work from spawned tasks, instead of implicitly executing
-the same code in a data parallel style. For the purpose of this paper, we took
+other threads will get work from spawned tasks instead of implicitly executing
+the same code in a data parallel manner. For the purpose of this work we took
 the shortcut of replacing these pragmas with no-ops. This worked well for all
-examples except for one, in which after the \texttt{parallel} and
+examples except for one in which after the \texttt{parallel} and
 \texttt{single} pragmas the top level task was called with an unnecessary
@@
 course, doesn't follow the OpenMP specification, and should be addressed by
-future work. See Section~\ref{Sec:Future} for a further discussion how how the 
+future work. Section~\ref{Sec:Future} discusses how how the 
 implementation can be improved.
@@
 \texttt{https://github.com/lanl/openmpir-clang}. The version used for this
-paper is tagged as LLVM17.
+paper is tagged LLVM17.
 
@@
 parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
-Figure~\ref{fig:tapir-example} we show what the input OpenMP code and the Tapir code
-is generated.
+Figure~\ref{fig:tapir-example} we show the input OpenMP code and the Tapir code
+that is generated.
 
@@
 Finally, the continuation for the second call immediately calls sync. This is
-actually a nice example of code that will be optimized by Tapir. The LLVM code
-shown in \figref{tapir-example} is not optimized. With optimizataions turned
-on, Tapir will replace a \texttt{detach} of a function call followed
+a nice example of code that will be optimized by Tapir. The LLVM code
+shown in \figref{tapir-example} is not optimized. With optimizations turned
+on Tapir will replace a \texttt{detach} of a function call followed
 immediately by \texttt{sync} with a simple function call, resulting in faster
@@
 
-For evaluating the use of Tapir we compared performance on the Barcelona OpenMP Task 
-Suite~\cite{barcelona} to several existing OpenMP implementations. The suite is a set 
+To evaluate the use of Tapir we compared performance on the Barcelona OpenMP Task 
+Suite to several existing OpenMP implementations. The suite is a set 
 of tests intended to test performance of OpenMP implementations
@@
 very recently added to the suite. Unfortunately we were unable to succesfully run 
-it on any implementation, even for even medium input sizes. We have therefore left 
+it on any implementation, even for medium input sizes. We have therefore left 
 it out of our evaluation.  We will continue to work on this benchmark and will 
 include the results if the paper is accepted to the workshop.} The following list provides abridged descriptions of 
-each benchmark -- they are taken verbatim from~\cite{barcelona}:
+each benchmark taken verbatim from the test suite distribution.
 
@@
 \item \texttt{FFT}: computes the one-dimensional Fast Fourier Transform
-of a vector of n complex values using the Cooley-Tukey 
+of a vector of $n$ complex values using the Cooley-Tukey 
 algorithm. This is a divide and conquer algorithm that  recursively  breaks
@@
 divisions multiple tasks are generated.
-\item \texttt{Fibonacci}: computes the n'th fibonacci number using a  recursive
+\item \texttt{Fibonacci}: computes the $n$th Fibonacci number using a  recursive
 paralellization. While  not  representative  of  an efficient  fibonacci
@@
 \item \texttt{Floorplan}: kernel computes the optimal floorplan distribution
-of a number of cells. The algorithm gets an input file with  cell’s
-description  and  it  returns  the  minimum  area  size which includes all
+of a number of cells. The algorithm gets an input file with  cells'
+descriptions  and  it  returns  the  minimum  area  size which includes all
 cells. This minimum area is found through a recursive branch and bound search.
@@
 of the solution.
-\item \texttt{Sort}: sorts a random permutation of n 32-bit numbers with  a
+\item \texttt{Sort}: sorts a random permutation of $n$ 32-bit numbers with  a
 fast  parallel  sorting  variation of  the  ordinary mergesort.  First, it
@@
 each  split and merge. When the array is too small, a serial quicksort is used
-so increase  the  task  granularity.  To  avoid  the overhead of  quicksort, an
+to increase  the  task  granularity.  To  avoid  the overhead of  quicksort, an
 insertion  sort  is  used  for  very  small  arrays (below a threshold of 20
@@
 There are two classes of input sizes for the benchmarks. Some, like Floorplan,
-require input files. For each of these, we chose the largest provided. For the
+require input files. For each of these we chose the largest provided. For the
 benchmarks with a parameter to adjust the size of the input, we attempted to 
-choose a parameter size so that the fastest implementation was on the order of
+choose a parameter size so that the fastest implementation ran on the order of
 seconds. The one exception was for \texttt{Fibonacci}. Because, as discussed
-further below, the Intel implementation required large stacksize, we stopped at
+further below, the Intel implementation required a large stack size, we stopped at
 42. The exact inputs used are:
@@
 
-The Intel compiler required increasing the stacksize for \texttt{Fibonacci} and
+The Intel compiler required increasing the stack size for \texttt{Fibonacci} and
 \texttt{FFT}. The implementation seemed to use significantly more stack space
 than the others, requiring setting \texttt{OMP\_STACKSIZE} to \texttt{256M} for
-computing the 42nd Fibonacci number. \texttt{FFT} ran fine with an increase to
+computing the 42nd Fibonacci number. \texttt{FFT} ran with an increase to
 \texttt{16M}.
@@
 17.0.0. The machine used is a two socket Intel Xeon E5-2683 v3 machine with
-132GB of memory, running Linux 4.11.4. Each Xeon has 16 cores running at
+132GB of memory running Linux 4.11.4. Each Xeon has 16 cores running at
 2.1GHz, with 20M of L3 cache. All benchmarks were run using all 64
-hyper-threads. 32 thread tests were run as a sanity check and showed
+hyper-threads. 32-thread tests were run as a sanity check and showed
 qualitatively similar results. It's worth noting that each of the other
-implementations comes with it's own runtime. In this sense, performance is a
-function of both the compiler and any optimizations it's able to perform, and
-the runtime.  Understanding the interaction of these two parts is non-trivial,
+implementations comes with its own runtime. In this sense, performance is a
+function of both the compiler and any optimizations it is able to perform, and
+the runtime.  Understanding the interaction of these two parts is non-trivial
 as we will see when discussing results. Each compiler was run with the flags
@@
 As mentioned in the Section~\ref{Sec:Implementation}, the gaps in the implementation
-changed program behaviour in a couple cases. In the case of \texttt{FFT}, we were
+changed program behavior in a couple cases. In the case of \texttt{FFT}, we were
 forced to remove an unecessary \texttt{task} pragma, as the current implementation
@@
 
-The second change in program behaviour due to our incomplete implementation was
-caused by the lack \texttt{critical} pragma and \texttt{atomic} pragmas. This
-issue showed up in the \texttt{Floorplan} benchmark. The \texttt{critical} pragma
-ensures that only one thread can executed the referred statement at a time. 
+The second change in program behavior due to our incomplete implementation was
+caused by the lack of the \texttt{critical} and \texttt{atomic} pragmas used
+in the \texttt{Floorplan} benchmark. The \texttt{critical} pragma
+ensures that only one thread can execute a particular statement at a time. 
 This should be fixable in the implementation by adding a simple code-localized
-synchronization generation in the IR. Similarly, the \texttt{atomic} pragma 
+synchronization generation in the IR\@. Similarly, the \texttt{atomic} pragma 
 can be addressed by retaining a pointer to the shared stack variable, much like
-existing OpenMP implementations. It is worth noting that this behaviour was
-non-deterministic, and that roughly half of the time the \texttt{Floorplan}
+existing OpenMP implementations. It is worth noting that this behavior was
+non-deterministic, and that roughly half of the time \texttt{Floorplan}
 still returned correct results. As we will see, this makes for an interesting
-trade-off given the performance increase witnessed.
-
-We set a timeout of 10 minutes, as at least one implementation was always
+trade-off given the witnessed performance increase.
+
+We set a time-out of 10 minutes, as at least one implementation was always
 finishing within 10 seconds, anything running more than 60 times slower becomes
-irrelevant. Due to time constraints, correctness checks were not performed on
-every run, so it is possible some non-determinism was missed.
+irrelevant.  Because of time constraints correctness checks were not performed on
+every run so it is possible some non-determinism was missed.
 
@@
 mean and the error bars representing standard deviation. In cases where there
-was timeout or segmentation fault, no time is reported, and the run is marked
+was timeout or segmentation fault no time is reported and the run is marked
 as faulty.
@@
 As mentioned earlier on \texttt{Floorplan} our implementation returned correct
-results nondeterministically due to not having an implementation of the OpenMP
+results nondeterministically because of not having an implementation of the OpenMP
 \texttt{critical} and \texttt{atomic} pragmas. While ICC and Clang finished in
-roughly 1-2 seconds, the Tapir implementation was finishing in 0.02 seconds,
-and computing the correct result roughly half the time. This raises interesting
+roughly 1-2 seconds, the Tapir implementation was finishing in 0.02 seconds
+and computing the correct result roughly half of the time. This raises interesting
 questions on the cost/benefit relation of the missing OpenMP pragmas.
@@
 GCC implementation was timing out at 600 seconds, so was at least 100 times
-slower. We've left GCC results off of the graphs for the three timeout cases,
+slower. GCC results are omitted from graphs for the three timeout cases,
 \texttt{FFT}, \texttt{Fibonacci}, and \texttt{NQueens}.
@@
 Cilk) implementation shows significant benefit. For example, for the \texttt{Fibonacci}
-benchmark, our implementation finishes significantly faster than any of the
-others. While ICC failed for that sized input, we did test with smaller inputs
-and found it was similarly outpaced by the Tapir implementation.  \texttt{NQueens}
-and \texttt{FFT} show similar behaviour, to lesser degrees. In contrast,
+benchmark, our implementation runs significantly faster than any of the
+others. While ICC failed for that size input, we did test with smaller inputs
+and found it was similarly outperformed by the Tapir implementation.  \texttt{NQueens}
+and \texttt{FFT} show similar behavior but to lesser degrees. In contrast,
 benchmarks that have a lower overhead to work ratio unsurprisingly differ less
@@
 
+%%% something missing here!
+
 \subsection{Profiling}
@@
 In this section we speculate \emph{why} the performance varies in
-the ways it does. While we have yet to figure out every discrepancy in
+the ways it does. While we have yet to identify every discrepancy in
 performance, we hope to get some ideas for why the performance of our
-implementation is generally better and where each of the implementations are
-spending a majority of the time. Our primary tool for this task will be performance 
+implementation is generally better and where each of the implementations is
+spending the bulk of the execution time. Our primary tool for this task is performance 
 counters. It's worth noting that some information is difficult to infer from 
-performance counters, such as sources of contention, reasons for memory locality 
-issues, etc. Still, they do give us some insight into performance bottlenecks.
+performance counters, such as sources of contention, reasons for memory locality,
+etc. Still, they do give us some insight into performance bottlenecks.
 
 Before we begin understanding why, it's worth noting again that the scope of the 
-work that this paper is responsible for is only the frontend. All performance
-gains due to the existing Tapir backend and Cilk runtime we sadly cannot claim
-credit for. We can break reasons for performance discrepancies into a few categories:
-
+work that this paper is responsible for is only the front end. All performance
+gains attributable to the existing Tapir back end and Cilk runtime represent prior work.
+We can break reasons for performance discrepancies into a few categories: 
 \begin{itemize}
-\item Front-end code generation
-\item IR Optimizations
-\item Runtime efficiency
+\item Front-end code generation;
+\item IR Optimizations; and
+\item Runtime efficiency.
 \end{itemize}
-
 While we would like to distinguish between these, it is difficult to do so with
 only performance counter information. A more thorough investigation would
-require deep knowledge of each of the frontends and runtimes. While the
+require deep knowledge of each of the front ends and runtimes. While the
 importance of optimizations of parallel code is clear, we hypothesize that most
@@
 and the code leaves little room for optimization (algorithmic rewrites
-notwithstanding). In other words, it is effectively a benchmark of how fast a
+aside), so it is effectively a benchmark of how fast a
 runtime can execute fork-join parallelism.
@@
 
-There are two cases we'd like to look at using the profiler. The first is a 
-case when there is a large difference in runtime between implementations. For 
-this case, our hypothesis is that in these cases the slower implementations
-are, for some reason, spending more time in the runtime. As mentioned above, 
+There are two cases we examing using the profiler. The first is the
+case when there is a large difference in run times between implementations. For 
+these our hypothesis is that the slower implementations
+are for some reason spending more time in the runtime. As mentioned, 
 it is difficult to understand exactly why an implementation is spending more 
@@
 sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
-Indeed, we get roughly the following breakdowns of work to overhead ratio
-according to the Linux \texttt{perf} tool:
-
+Indeed, we get roughly the following breakdowns of work-to-overhead ratio
+according to the Linux \texttt{perf} tool.
 \begin{itemize}
@@
 \end{itemize}
-
 This supports our hypothesis that most of the slowdown incurred is overhead in
@@
 
-For the cases where performance is close between the implementations, like
-\texttt{sparselu}, we expect runtimes to be dominated by actual task workloads. This 
+For the cases where performance is close between the implementations, such as
+\texttt{sparselu}, we expect run times to be dominated by actual task workloads. This 
 expectation is confirmed by investigating performance counters for \texttt{sparselu}
-where every implementation spends roughly 90\% of it's time in the task
+where every implementation spends roughly 90\% of its time in the task
 bodies.
@@
 accept that OpenMP is currently/potentially the most widely used model that has 
-a notion of tasks. \pmnote{but it isn't clear to me that tasks are really being 
-used much! :-)}.  The goal of this work is to show that a single intermediate representation can be
-used to implement multiple programming paradigms with good performance. Schardl et 
-al. showed that Tapir works well as an IR target for Cilk and surmised that it could
-be an effective IR for a large subset of OpenMP \cite{tapir}. This work has
-taken the first step in that direction.
-
-One interesting aspect of this work is that it compiles a language extension intended
-for one runtime (OpenMP) to use another (the Cilk runtime). While in many
-ways this is a weakness of this work, it does open the possibility of utilizing this 
-in a more general way to address many aspects including significant challenges 
-such as performance portability. Having a parallel-aware IR to work with enables, 
-in the same way that LLVM does for serial code, the possibility of multiple high-level 
-parallel languages and multiple parallel runtime targets. While the OpenMP specification's 
-requirement to have access to runtime routines complicates things, for many programming models
-it seems at least conceivable to map them onto multiple different runtimes. This
-opens the possibility of mixing programming models at compile time by targeting
-the same runtime. For example, if both OpenMP and Cilk targeted Tapir
-instructions, one could have a large application where some parts were written
-using Cilk, while other used OpenMP, and the result would not suffer the
-incompatibility problems often seen today.
-
-The surprising fact that program behaviour wasn't changed significantly is
-worth discussing here. We surmise that it is likely due to the variable
-semantics borrowed from the Cilk codegen and their relation to the OpenMP
+a notion of tasks. The goal of this work is to show that a single intermediate
+representation can be used to implement multiple programming paradigms with
+good performance. Schardl et al.\ showed that Tapir works well as an IR target
+for Cilk and surmised that it could be an effective IR for a large subset of
+  OpenMP~\cite{tapir}. This work has taken the first step in that direction.
+
+One interesting aspect of this work is that it compiles a language extension
+intended for one runtime (OpenMP) to use another (the Cilk runtime). While in
+many ways this is a weakness of this work, it does open the possibility of
+utilizing this in a more general way to address many significant challenges
+such as performance portability. Having a parallel-aware IR enables the
+possibility of multiple high-level parallel languages and multiple parallel
+runtime targets in the same way that LLVM does for serial code. While the
+OpenMP specification's requirement to have access to runtime routines
+complicates things, for many programming models it seems at least conceivable
+to map them onto multiple different runtimes. This opens the possibility of
+mixing programming models at compile time by targeting the same runtime. For
+example, if both OpenMP and Cilk targeted Tapir instructions, one could have a
+large application where some parts were written using Cilk, while other used
+OpenMP, and the result would not suffer the incompatibility problems often
+seen today.
+
+The surprising fact that program behavior wasn't changed significantly is
+worth discussing here. We surmise that it is likely due to the
+semantics of variables borrowed from the Cilk codegen and their relation to the OpenMP
 specification of semantics. In particular, \emph{OpenMP doesn't specify a
@@
 implementations often write to shared variables on every write. This is a
-potential for performance discrepancy due to differing implementation-defined
-behaviour and we will return to it when discussing performance on the
+potential for performance discrepancy because of differing implementation-defined
+behavior and we will return to it when discussing performance on the
 Barcelona OpenMP task suite. There is of course the fact that technically, by
@@
 \texttt{private} variables, one isn't following the specification. We surmise
-that in practice this hasn't had an effect due to the fact that
-\texttt{private} variables are generally used for performance, rather than
+that in practice this hasn't had an effect because
+\texttt{private} variables are generally used for performance rather than
 their semantic properties.
@@
 
-One of the important questions for this work is whether or not the Tapir
-instructions can be extended to cover the full OpenMP semantics. As shown
-in this paper, even the full semantics of OpenMP tasks weren't covered. Runtime
+One of the important questions in this work is whether the Tapir
+instructions can be extended to cover the full OpenMP semantics.
+%As we have shown, even the full semantics of OpenMP tasks weren't covered.
+Runtime
 calls aside, implementing the full semantics of OpenMP pragmas would be a
@@
 
-As two examples, let us consider the problematic \texttt{atomic} and
+For two examples, consider the problematic \texttt{atomic} and
 \texttt{critical} pragmas from the \texttt{Floorplan} benchmark. The 
 \texttt{atomic} pragma would likely not be too difficult to translate
-from existing OpenMP codegen. One would have to simply keep around and then
+from existing OpenMP codegen. One would need to simply maintain and then
 modify the contents of the original stack pointer to the shared variable. The
@@
 compiler cannot reason about code using those features, and many of the
-advantages we've discussed become moot, e.g. backend compatibility, generic 
+advantages we've discussed become moot, e.g., backend compatibility, generic 
 optimization, etc. Features like the OpenMP \texttt{atomic} pragma don't 
@@
 \texttt{critical} section \emph{do} require changes to control flow that seem
-challenging at best to map onto Tapir's instructions. Furthermore, even if it
+challenging to map onto Tapir's instructions. Furthermore, even if it
 is possible to implement features using the Tapir instructions, it may require 
 such radical entangling of the code that even if analysis is technically
-possible, it becomes infeasible.
+possible it becomes infeasible.
 
 An obvious first step towards better OpenMP coverage, other than the features
-listed above, is OpenMP \texttt{parallel for} loops. These are similar enough
+listed above, is handling OpenMP \texttt{parallel for} loops. These are similar enough
 to \texttt{cilk\_for} loops in semantics that this also should be a relatively
@@
 While OpenMP and Cilk have similar fork-join style semantics that Tapir
-implements naturally, many allow a more flexible mechanism for synchronization:
-futures \cite{qthreads, chapel, hpx}. The primary difference between the
+implements naturally, many allow a more flexible mechanism for synchronization, namely
+futures~\cite{qthreads, chapel, hpx}. The primary difference between the
 fork-join parallelism represented by Tapir and these approaches is that
-child tasks can outlive their parents. Figuring out how to represent this class
+child tasks can outlive their parents. Representing this class
 of parallelism in Tapir represents a significant challenge for future work.
@@
 One interesting area for future work would be combining Tapir with work on
-reasoning about locality in LLVM. For example, Hayashi et al.
-\cite{hayashi2015llvm} use address spaces in LLVM to reason about memory
-locality in partitioned global address space (PGAS) systems. By combining
+reasoning about locality in LLVM. For example, Hayashi et al.\ 
+use address spaces in LLVM to reason about memory
+locality in partitioned global address space (PGAS) systems~\cite{hayashi2015llvm}. By combining
 reasoning about code concurrency and locality, one could potentially apply
@@
 First, compilers can attach metadata to an existing IR. For example, LLVM has a
-parallel loop metadata construct \cite{llvmref}. This has the benefit of being
+parallel loop metadata construct~\cite{llvmref}. This has the benefit of being
 flexible and requiring minimal work, but historically can be lost during
@@
 
-A second approach is to use intrinsic functions to define parallel tasks
-\cite{ares}. This, like the metadata approach, has the advantage of being
+A second approach is to use intrinsic functions to define parallel tasks~\cite{ares}.
+This, like the metadata approach, has the advantage of being
 flexible and relatively easy to implement, but at the cost of being less well
@@
 for LLVM to represent parallelism, but we would caution that a large, flexible
-set of intrinsics being added to an already only-partially defined
-\cite{verillvm} language could make reasoning about parallel programs, even
-in an IR settings, infeasible.
+set of intrinsics being added to an already only-partially defined~\cite{verillvm} language could make reasoning about parallel programs, even
+in an IR setting, infeasible.
 
@@
 approaches is best. Approaches that are easier at first, such as metadata
-or intrinsic approaches, are similarly easy to extend to be unwieldy. There is
+or intrinsic approaches, are similarly easy to extend to become unwieldy. There is
 a reason that it's difficult to add instructions to LLVM: any added IR
@@
 out IR, to case analysis for different control flow constructs, having 
-a proper set of instructions to denote parallelism has a lot of advantages. 
+a proper set of instructions to denote parallelism has numerous advantages. 
 Additionally, any hope of having a formal semantics for an IR depends on a 
@@
 generation approach, but replace calls into built in runtimes with calls into a
-shared runtime. This is the approach taken by Lithe \cite{lithe}, where Pan et
-al. compose programs using both GNU OpenMP and Intel TBB. This approach succeeds
+shared runtime. This is the approach taken by Lithe, where Pan et
+al.\ compose programs using both GNU OpenMP and Intel TBB~\cite{lithe}. This approach succeeds
 in composing different programming models at run-time, but doesn't enable the

### 2017-09-08 paper.tex
Updated health plot, added clangs stacksize

@@
 than the others, requiring setting \texttt{OMP\_STACKSIZE} to \texttt{256M} for
-computing the 42nd Fibonacci number. \texttt{FFT} ran with an increase to
-\texttt{16M}.
+computing the 42nd Fibonacci number. \texttt{FFT} ran fine with an increase to
+\texttt{16M}. Clang required a similar increase in \texttt{OMP\_STACKSIZE} for
+\texttt{SparseLU}, needing to be increased to \texttt{64M} to avoid stack
+overflows. GCC and the Tapir implementation required no such modifications. 
+>>>>>>> Updated health plot, added clangs stacksize
 

### 2017-09-08 plot.py

@@ -42,7 +42,7 @@ for p in ['floorplan']: # pdat:
     plt.savefig('figs/%s.pdf' %(p))
 '''
 
-for p in ['floorplan']: 
+for p in ['health']: 
     fig = plt.figure()
     axb = fig.add_subplot(111)
     ax = fig.add_subplot(211)
@@ -55,8 +55,8 @@ for p in ['floorplan']:
     ind = np.arange(len(means))
     ax.bar(ind, means, width, yerr=stds)
     ax2.bar(ind, means, width, yerr=stds)
-    ax.set_ylim(170, 190)
-    ax2.set_ylim(0, 3)
+    ax.set_ylim(420, 480)
+    ax2.set_ylim(0, 20)
 
 # hide the spines between ax and ax2
     ax.spines['bottom'].set_visible(False)
@@ -91,6 +91,6 @@ for p in ['floorplan']:
     axb.tick_params(labelcolor='w', top='off', bottom='off', left='off', right='off')
 
     ax.set_xticks(ind)
-    ax.set_xticklabels(['tapir*'] + types[1:])
+    ax.set_xticklabels(types)
     ax.set_title(pdat[p]['title'])
     plt.savefig('figs/%s.pdf' %(p))

### 2017-09-08 paper.tex
changed back to barcelona openmp task suite

@@
 Optimizing compilers for task-level parallelism are still in their infancy.
-This work explores a compiler front end that translates OpenMP tasking semantics to Tapir,
-an extension to LLVM IR that represents fork-join parallelism. This enables
-analyses and optimizations that were previously inaccessible to OpenMP codes,
-as well as the ability to target additional runtimes at code generation. Using
-a Cilk runtime back end, we compare results to existing OpenMP implementations.
-Initial performance results for the Barcelona OpenMP tasking suite show
-performance improvements over existing implementations.
+This work explores a compiler front end that translates OpenMP tasking
+semantics to Tapir, an extension to LLVM IR that represents fork-join
+parallelism. This enables analyses and optimizations that were previously
+inaccessible to OpenMP codes, as well as the ability to target additional
+runtimes at code generation. Using a Cilk runtime back end, we compare results
+to existing OpenMP implementations.  Initial performance results for the
+Barcelona OpenMP task suite show performance improvements over existing
+implementations.
 \end{abstract}
@@
 will cooperate with management of hardware resources.
-%
+
 Addressing these concerns is a major motivation for our use of Tapir.  By
@@
 the possibility of multiple programming models coexisting peacefully. We will
-return to this consideration later in Section~\ref{Sec:Discussion} when discussing
-future work.
+return to this consideration later in Section~\ref{Sec:Discussion} when
+discussing future work.
 
@@
 compilers to support parallel programs but meant that traditional 
-optimizations such as code motion~\cite{} or common-subexpression elimination~\cite{}
-weren't able to reason about parallel programs. This often led to parallel
-programs running significantly slower than expected. In the Tapir project
-the authors show that it is possible to represent and optimize fork-join
-parallel programs with relatively minimal modifications to the compiler, namely
-the introduction of three instructions designed to interface well with an
-existing compiler.
+optimizations such as code motion or common-subexpression elimination weren't
+able to reason about parallel programs. This often led to parallel programs
+running significantly slower than expected. In the Tapir project the authors
+show that it is possible to represent and optimize fork-join parallel programs
+with relatively minimal modifications to the compiler, namely the introduction
+of three instructions designed to interface well with an existing compiler.
 

### 2017-09-09 paper.tex
spelling

@@
 compilers such as LLVM~\cite{lattner2004llvm} or GCC's
-Gimple~\cite{merrill2003generic} haven't suppored parallel constructs. As a
+Gimple~\cite{merrill2003generic} haven't supported parallel constructs. As a
 consequence, compilers using these IRs haven't had the ability to reason
@@
 analogous the problem LLVM solves: compiler optimizations are hard, so sharing
-them is valuable. Tapir is a demonstration that the the LLVM approach to
+them is valuable. Tapir is a demonstration that the LLVM approach to
 compilation of serial code can be effectively extended to help reason about
@@
 functions and accompanying closures. These functions are then passed to OpenMP runtime 
-calls. This system obfuscates the program, making it infeasable for an optimizer to 
+calls. This system obfuscates the program, making it infeasible for an optimizer to 
 reason about what is happening.}
@@
 \item \texttt{Fibonacci}: computes the $n$th Fibonacci number using a  recursive
-paralellization. While  not  representative  of  an efficient  fibonacci
+parallelization. While  not  representative  of  an efficient  fibonacci
 computation  it  is  still  useful  because  it  is a simple test case of a
@@
 overflows. GCC and the Tapir implementation required no such modifications.
->>>>>>> Updated health plot, added clangs stacksize
 
@@
 changed program behavior in a couple cases. In the case of \texttt{FFT}, we were
-forced to remove an unecessary \texttt{task} pragma, as the current implementation
+forced to remove an unnecessary \texttt{task} pragma, as the current implementation
 doesn't insert the implicit barrier at the end of a \texttt{parallel} region. This 
@@
 
-There are two cases we examing using the profiler. The first is the
+There are two cases we examine using the profiler. The first is the
 case when there is a large difference in run times between implementations. For 
@@
 \section*{Acknowledgments}
+
+Thanks to Kei Davis for his excellent editing.
+
 Sandia National Laboratories is a multimission laboratory managed and operated 

### 2017-09-09 paper.tex
Fixed weird typos in omp pragmas in text

@@
 
+\author{Kei Davis}
+\affiliation{\institution{Los Alamos National Laboratory}}
+\email{kei@lanl.gov}
+
 \author{Pat McCormick}
@@
 Tapir IR instead of vanilla LLVM IR with OpenMP runtime calls. Specifically, we
-replace the two primary \texttt{task\_spawn} and
-\texttt{task\_wait} pragmas for task parallelism. By re-using code from Schardl et al.\ for code generation
-of Cilk constructs, we were able to easily generate Tapir code for these OpenMP
-pragmas. The ease with which this was completed is a testament to the quality of
-the Tapir implementation.
-
-While we did implement the codegen for the \texttt{task\_spawn} and
-\texttt{task\_wait} constructs, it's worth noting that even for these
-pragmas the implementation is incomplete. Currently, any clauses modifying the
-behavior of the pragmas is ignored. Probably the most common semantics this
-will change is for the semantcis of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
+replace the two primary \texttt{task} and \texttt{taskwait} pragmas for task
+parallelism. By re-using code from Schardl et al.\ for code generation of Cilk
+constructs, we were able to easily generate Tapir code for these OpenMP
+pragmas. The ease with which this was completed is a testament to the quality
+of the Tapir implementation.
+
+While we did implement the codegen for the \texttt{task} and
+\texttt{taskwait} constructs, it's worth noting that even for these pragmas
+the implementation is incomplete. Currently, any clauses modifying the behavior
+of the pragmas is ignored. Probably the most common semantics this will change
+is for the semantcis of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
 Surprisingly, this had little effect on the correctness of the entire Barcelona
-OpenMP task suite. Indeed, it is likely that fixing this problem would marginally increase
-performance by reducing the number of memory
-copies for variables declared \texttt{private} in OpenMP code.  The fact
-that behaviour wasn't changed significantly is worth revisiting, and we do so
-in Section~\ref{Sec:Discussion}.
+OpenMP task suite. Indeed, it is likely that fixing this problem would
+marginally increase performance by reducing the number of memory copies for
+variables declared \texttt{private} in OpenMP code.  The fact that behaviour
+wasn't changed significantly is worth revisiting, and we do so in
+Section~\ref{Sec:Discussion}.
 
@@
 
-Thanks to Kei Davis for his excellent editing.
-
 Sandia National Laboratories is a multimission laboratory managed and operated 

### 2017-10-04 paper.tex
Addressed reviewers

@@
 
-\author{Kei Davis}
-\affiliation{\institution{Los Alamos National Laboratory}}
-\email{kei@lanl.gov}
-
 \author{Pat McCormick}
@@
 and compilers can potentially help reduce this burden. This is exactly
-analogous the problem LLVM solves: compiler optimizations are hard, so sharing
-them is valuable. Tapir is a demonstration that the LLVM approach to
+analogous to the problem LLVM solves: compiler optimizations are hard, so
+sharing them is valuable. Tapir is a demonstration that the LLVM approach to
 compilation of serial code can be effectively extended to help reason about
@@
 \code{sync} instructions. The \code{detach} statement is syntactically similar to
-a conditional branch in that it is a terminator instruction with with two blocks.
+a conditional branch in that it is a terminator instruction with two blocks.
 However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
@@
 \textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
-is analagous to a \code{fork} in most fork-join models.
+is analogous to a \code{fork} in most fork-join models.
 
@@
 by \code{detach} instructions within the current function to complete. The \code{sync}
-instruction is analagous to a \code{join} in most fork-join frameworks.
+instruction is analogous to a \code{join} in most fork-join frameworks.
 
@@
 of the pragmas is ignored. Probably the most common semantics this will change
-is for the semantcis of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
+is for the semantics of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
 Surprisingly, this had little effect on the correctness of the entire Barcelona
@@
 using Tapir. Generally, in addition the the \texttt{task} 
-and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP
-%pragmas in order to run parallel code.
-A standard pattern for building
-task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
-the necessary hardware threads, followed immediately by a \texttt{single}
-pragma to have only one thread continue on the specified statement. Then the
-other threads will get work from spawned tasks instead of implicitly executing
-the same code in a data parallel manner. For the purpose of this work we took
-the shortcut of replacing these pragmas with no-ops. This worked well for all
-examples except for one in which after the \texttt{parallel} and
-\texttt{single} pragmas the top level task was called with an unnecessary
-\texttt{task} pragma.  Because the wait was implicit and unhandled by our
-implementation, the simple fix was to remove the \texttt{task} pragma. This was
-the only change required to the source code. Our temporary no-op shortcut, of
-course, doesn't follow the OpenMP specification, and should be addressed by
-future work. Section~\ref{Sec:Future} discusses how how the 
+and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP.
+A standard pattern for building task-parallel OpenMP programs is to insert a
+\texttt{parallel} pragma to start the necessary hardware threads, followed
+immediately by a \texttt{single} pragma to have only one thread continue on the
+specified statement. Then the other threads will get work from spawned tasks
+instead of implicitly executing the same code in a data parallel manner. For
+the purpose of this work we took the shortcut of replacing these pragmas with
+no-ops. This worked well for all examples except for one in which after the
+\texttt{parallel} and \texttt{single} pragmas the top level task was called
+with an unnecessary \texttt{task} pragma.  Because the wait was implicit and
+unhandled by our implementation, the simple fix was to remove the \texttt{task}
+pragma. This was the only change required to the source code. Our temporary
+no-op shortcut, of course, doesn't follow the OpenMP specification, and should
+be addressed by future work. Section~\ref{Sec:Future} discusses how how the
 implementation can be improved.
@@
 shown in \figref{tapir-example} is not optimized. With optimizations turned
-on Tapir will replace a \texttt{detach} of a function call followed
+on, Tapir will replace a \texttt{detach} of a function call followed
 immediately by \texttt{sync} with a simple function call, resulting in faster
@@
 As mentioned in the Section~\ref{Sec:Implementation}, the gaps in the implementation
-changed program behavior in a couple cases. In the case of \texttt{FFT}, we were
+changed program behavior in a couple of cases. In the case of \texttt{FFT}, we were
 forced to remove an unnecessary \texttt{task} pragma, as the current implementation
@@
 
-In this section we speculate \emph{why} the performance varies in
-the ways it does. While we have yet to identify every discrepancy in
-performance, we hope to get some ideas for why the performance of our
-implementation is generally better and where each of the implementations is
-spending the bulk of the execution time. Our primary tool for this task is performance 
-counters. It's worth noting that some information is difficult to infer from 
-performance counters, such as sources of contention, reasons for memory locality,
-etc. Still, they do give us some insight into performance bottlenecks.
-
-Before we begin understanding why, it's worth noting again that the scope of the 
-work that this paper is responsible for is only the front end. All performance
-gains attributable to the existing Tapir back end and Cilk runtime represent prior work.
-We can break reasons for performance discrepancies into a few categories: 
+In this section we investigate \emph{why} performance varies in the ways it
+does. While we have yet to identify every discrepancy in performance, we hope
+to get some ideas for why the performance of our implementation is generally
+better and where each of the implementations is spending the bulk of the
+execution time. Our primary tool for this task is profiling. It's worth noting
+that lots of information is difficult to infer from profiling, such as sources
+of contention, reasons for memory locality, etc.
+
+It is worth noting again that the scope of the work that this paper is
+responsible for is only the front end. All performance gains attributable to
+the existing Tapir back end and Cilk runtime represent prior work.  We can
+break reasons for performance discrepancies into a few categories:
 \begin{itemize}
@@
 While we would like to distinguish between these, it is difficult to do so with
-only performance counter information. A more thorough investigation would
-require deep knowledge of each of the front ends and runtimes. While the
-importance of optimizations of parallel code is clear, we hypothesize that most
-of the disparity present in these results is more a function of runtime
-implementation differences. As justification for this hypothesis, consider
-the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
-and the code leaves little room for optimization (algorithmic rewrites
-aside), so it is effectively a benchmark of how fast a
-runtime can execute fork-join parallelism.
+only profiling information. A more thorough investigation would require deep
+knowledge of each of the front ends and runtimes. While the importance of
+optimizations of parallel code is clear, we hypothesize that most of the
+disparity present in these results is more a function of runtime implementation
+differences. As justification for this hypothesis, consider the
+\texttt{Fibonacci} benchmark. There are no memory accesses to the heap, and the
+code leaves little room for optimization (algorithmic rewrites aside), so it is
+effectively a benchmark of how fast a runtime can execute fork-join
+parallelism.
 
@@
 This supports our hypothesis that most of the slowdown incurred is overhead in
-the runtime. For Tapir the overhead is relatively small given the
-size of the work chunks, for \texttt{clang} and \texttt{icc} it is
-significantly larger, and for \texttt{gcc} it is overwhelming.
+the runtime. For Tapir the overhead is relatively small given the size of the
+work chunks, for \texttt{clang} and \texttt{icc} it is significantly larger,
+and for \texttt{gcc} it is overwhelming. It's worth noting that \texttt{clang} 
+and \texttt{icc} have effectively the same runtime, so the fact that their 
+overheads are similar is unsurprising.
 
@@
 example, if both OpenMP and Cilk targeted Tapir instructions, one could have a
-large application where some parts were written using Cilk, while other used
+large application where some parts were written using Cilk, while others used
 OpenMP, and the result would not suffer the incompatibility problems often
@@
 
-The surprising fact that program behavior wasn't changed significantly is
-worth discussing here. We surmise that it is likely due to the
-semantics of variables borrowed from the Cilk codegen and their relation to the OpenMP
-specification of semantics. In particular, \emph{OpenMP doesn't specify a
-memory model for shared variables}. This results in the Cilk style of only
-writing to a shared variable at the end of a parallel section as a perfectly
-valid implementation of OpenMP's specification. In contrast, existing OpenMP
+The surprising fact that program behavior wasn't changed significantly is worth
+discussing. We surmise that it is likely due to the semantics of variables
+borrowed from the Cilk codegen and their relation to the OpenMP specification
+of semantics. In particular, \emph{OpenMP specifies a very relaxed memory model
+for shared variables}. This results in the Cilk implementation of only writing
+to a shared variable at the end of a parallel section being a perfectly valid
+implementation of OpenMP's specification. In contrast, existing OpenMP
 implementations often write to shared variables on every write. This is a
-potential for performance discrepancy because of differing implementation-defined
-behavior and we will return to it when discussing performance on the
-Barcelona OpenMP task suite. There is of course the fact that technically, by
-copying the value of variables back to the surrounding context even for
-\texttt{private} variables, one isn't following the specification. We surmise
-that in practice this hasn't had an effect because
-\texttt{private} variables are generally used for performance rather than
-their semantic properties.
+potential for performance discrepancy because of differing
+implementation-defined behavior and we will return to it when discussing
+performance on the Barcelona OpenMP task suite. There is of course the fact
+that technically, by copying the value of variables back to the surrounding
+context even for \texttt{private} variables, one isn't following the
+specification. We surmise that in practice this hasn't had an effect because
+\texttt{private} variables are generally used for performance rather than their
+semantic properties.
 
@@
 in composing different programming models at run-time, but doesn't enable the
-static analyses that enabled by Tapir.
+static analyses that are enabled by Tapir.
 
@@
 
+Many thanks to Kei Davis for his excellent editing.
+
 Sandia National Laboratories is a multimission laboratory managed and operated 
@@
 This work was supported by the Director, Office of Advanced Scientific
-Computing Research, Office of Science, of the United States Department
-of Energy, under the guidance of Dr. Sonia Sachs. Los Alamos National
-Laboratory is operated by Los Alamos National Security LLC for the U.S.
-Department of Energy under contract DE-AC52-06NA25396.
-
-This research was supported by US National Science Foundation under CRII
-ACI 1565338, CNS 1527076, and CNS 1217948. Los Alamos National
-Laboratory is operated by Los Alamos National Security LLC for the US
-Department of Energy under contract DE-AC52-06NA25396.
-
-Work supported by the Advanced Simulation and Computing program of the
-U.S. Department of Energy's NNSA\@.  Los Alamos National Laboratory is
+Computing Research, Office of Science, of the United States Department of
+Energy, under the guidance of Dr. Sonia Sachs. This research was supported by
+US National Science Foundation under CRII ACI 1565338, CNS 1527076, and CNS
+1217948.  Work supported by the Advanced Simulation and Computing program of
+the U.S. Department of Energy's NNSA\@. Los Alamos National Laboratory is
 managed and operated by Los Alamos National Security, LLC (LANS), under
-contract number DE-AC52-06NA25396 for the Department of Energy's
-National Nuclear Security Administration (NNSA).
+contract number DE-AC52-06NA25396 for the Department of Energy's National
+Nuclear Security Administration (NNSA).
 

### 2017-10-04 plot.py

@@ -26,23 +26,23 @@ for p in pdat:
 
 width=0.75
 
-'''
-for p in ['floorplan']: # pdat:
+colors=['#2D8632', '#236467', '#AA6D39', '#AA3C39']
+
+for p in ['alignment', 'fft', 'fib', 'nqueens', 'sort', 'sparselu', 'strassen']:
     fig, ax = plt.subplots()
     means = [np.mean(pdat[p][t]) for t in types]
     means = [m for m in means if not np.isnan(m)]
     stds = [np.std(pdat[p][t]) for t in types]
     stds = [m for m in stds if not np.isnan(m)]
     ind = np.arange(len(means))
-    ax.bar(ind, means, width, yerr=stds)
+    ax.bar(ind, means, width, yerr=stds, color=colors)
     ax.set_ylabel("Time (Seconds, lower is better)")
     ax.set_xticks(ind)
     ax.set_xticklabels(types)
     ax.set_title(pdat[p]['title'])
     plt.savefig('figs/%s.pdf' %(p))
-'''
 
-for p in ['health']: 
+for p in ['health', 'floorplan']: 
     fig = plt.figure()
     axb = fig.add_subplot(111)
     ax = fig.add_subplot(211)
@@ -53,10 +53,15 @@ for p in ['health']:
     stds = [np.std(pdat[p][t]) for t in types]
     stds = [m for m in stds if not np.isnan(m)]
     ind = np.arange(len(means))
-    ax.bar(ind, means, width, yerr=stds)
-    ax2.bar(ind, means, width, yerr=stds)
-    ax.set_ylim(420, 480)
-    ax2.set_ylim(0, 20)
+    ax.bar(ind, means, width, yerr=stds, color=colors)
+    ax2.bar(ind, means, width, yerr=stds, color=colors)
+    if p == 'health':
+        ax.set_ylim(420, 480)
+        ax2.set_ylim(0, 20)
+    if p == 'floorplan':
+        ax.set_ylim(170, 190)
+        ax2.set_ylim(0,3)
+
 
 # hide the spines between ax and ax2
     ax.spines['bottom'].set_visible(False)

### 2017-10-05 paper.tex
Added acm prelude
Removed minted dependecy

@@
 \usepackage{multicol}
-\usepackage{minted}
 \graphicspath{{./figs/}}
@@
 %\settopmatter{printacmref=false}
-\setcopyright{rightsretained}
+%\setcopyright{rightsretained}
 %\setcopyright{usgov}
-%\setcopyright{usgovmixed}
+\setcopyright{usgovmixed}
 %\setcopyright{cagov}
@@
 % DOI
-\acmDOI{10.475/123_4}
+\acmDOI{10.1145/3148173.3148186}
 
 % ISBN
-\acmISBN{123-4567-24-567/08/06}
+\acmISBN{978-1-4503-5565-0/17/11}
 
@@
 To evaluate the use of Tapir we compared performance on the Barcelona OpenMP Task 
-Suite to several existing OpenMP implementations. The suite is a set 
-of tests intended to test performance of OpenMP implementations
-using both irregular and regular tasking.\footnote{An unbalanced tree search benchmark was
-very recently added to the suite. Unfortunately we were unable to succesfully run 
-it on any implementation, even for medium input sizes. We have therefore left 
-it out of our evaluation.  We will continue to work on this benchmark and will 
-include the results if the paper is accepted to the workshop.} The following list provides abridged descriptions of 
-each benchmark taken verbatim from the test suite distribution.
+Suite to several existing OpenMP implementations. The suite is a set of tests
+intended to test performance of OpenMP implementations using both irregular and
+regular tasking. The following list provides abridged descriptions of each
+benchmark taken verbatim from the test suite distribution \cite{bots}.
 

### 2017-10-05 paper.tex
Small tweaks to my name (that darn 'c'), small edits and made sure we referenced ECP in the ACKS.

--- 2017-10-04 paper.tex
@@
 \usetikzlibrary{decorations.pathreplacing,shapes,arrows,positioning,calc}
+
+\newcommand{\McCormick} {M\raise .40ex\hbox{c}Cormick}
 
@@
 
-\author{Pat McCormick}
+\author{Patrick \McCormick}
 \affiliation{\institution{Los Alamos National Laboratory}}
@@
 Compilers implement the majority of their analyses and optimizations on
-intermediate representations \textbf{(IRs)} which allow such transformations
+intermediate representations \textbf{(IRs)} that allow such transformations
 to be written once with relative ease and apply to a variety of source
@@
 of design and implementaton details to be considered. We argue that the benefits vastly outweigh 
-the costs and we hope to see additional future work explore additional possibilities.
+the costs and we hope to see future work explore additional possibilities.
 
@@
 
-Sandia National Laboratories is a multimission laboratory managed and operated 
-by National Technology and Engineering Solutions of Sandia, LLC., a wholly 
-owned subsidiary of Honeywell International, Inc., for the U.S. Department of 
-Energy's National Nuclear Security Administration under contract DE-NA-0003525.
-
 This work was supported by the Director, Office of Advanced Scientific
-Computing Research, Office of Science, of the United States Department of
-Energy, under the guidance of Dr. Sonia Sachs. This research was supported by
-US National Science Foundation under CRII ACI 1565338, CNS 1527076, and CNS
-1217948.  Work supported by the Advanced Simulation and Computing program of
-the U.S. Department of Energy's NNSA\@. Los Alamos National Laboratory is
-managed and operated by Los Alamos National Security, LLC (LANS), under
-contract number DE-AC52-06NA25396 for the Department of Energy's National
-Nuclear Security Administration (NNSA).
+Computing Research, Office of Science, of the United States Department
+of Energy, under the guidance of Dr. Sonia Sachs. This research was
+supported by US National Science Foundation under CRII ACI 1565338,
+CNS 1527076, and CNS 1217948 and by the Exascale Computing Project
+(17-SC-20-SC), a collaborative effort of the U.S. Department of Energy
+Office of Science and the National Nuclear Security Administration.
+Los Alamos National Laboratory is managed and operated by Los Alamos
+National Security, LLC (LANS), under contract number DE-AC52-06NA25396
+for the Department of Energy's National Nuclear Security
+Administration (NNSA). Sandia National Laboratories is a multimission
+laboratory managed and operated by National Technology and Engineering
+Solutions of Sandia, LLC., a wholly owned subsidiary of Honeywell
+International, Inc., for the U.S. Department of Energy's National
+Nuclear Security Administration under contract DE-NA-0003525.
 

### 2017-10-05 paper.tex
Another small tweak.

@@
 approaches is best. Approaches that are easier at first, such as metadata
-or intrinsic approaches, are similarly easy to extend to become unwieldy. There is
+or intrinsic approaches, are similarly easy to extend but can become unwieldy. There is
 a reason that it's difficult to add instructions to LLVM: any added IR
@@
 A related approach to composing programming models is to keep the same code
-generation approach, but replace calls into built in runtimes with calls into a
+generation approach, but replace calls into built-in runtimes with calls into a
 shared runtime. This is the approach taken by Lithe, where Pan et

### 2017-10-06 paper.tex
Fixed citation for Barcelona benchmarks.
edits to memory model discussion

@@
 regular tasking. The following list provides abridged descriptions of each
-benchmark taken verbatim from the test suite distribution \cite{bots}.
+benchmark taken verbatim from the test suite distribution \cite{barcelona}.
 
@@
 borrowed from the Cilk codegen and their relation to the OpenMP specification
-of semantics. In particular, \emph{OpenMP specifies a very relaxed memory model
-for shared variables}. This results in the Cilk implementation of only writing
-to a shared variable at the end of a parallel section being a perfectly valid
+of semantics. In particular, \emph{OpenMP specifies a relaxed memory model
+for shared variables}, in which writes by one thread need only be visible at
+to other threads at explicit or implicit flushes. 
+This results in the Cilk implementation of only writing
+to a shared variable at the end of a parallel section being a valid
 implementation of OpenMP's specification. In contrast, existing OpenMP
-implementations often write to shared variables on every write. This is a
+implementations sometimes write shared variables to memory on every write access. This is a
 potential for performance discrepancy because of differing 
-implementation-defined behavior and we will return to it when discussing
-performance on the Barcelona OpenMP task suite. There is of course the fact
+implementation behavior.
+%Stephen:This topic is not discuessed later, but was touched on before -- moved?
+%and we will return to it when discussing
+%performance on the Barcelona OpenMP task suite. 
+There is of course the fact
 that technically, by copying the value of variables back to the surrounding
@@
 specification. We surmise that in practice this hasn't had an effect because
-\texttt{private} variables are generally used for performance rather than their
-semantic properties.
+\texttt{private} variables are often used for input parameters in the 
+progams or for performance rather than their semantic properties.
+Also, \texttt{firstprivate} is the default data-sharing attribute for tasks.
 
@@
 for the Department of Energy's National Nuclear Security
-Administration (NNSA). Sandia National Laboratories is a multimission
-laboratory managed and operated by National Technology and Engineering
-Solutions of Sandia, LLC., a wholly owned subsidiary of Honeywell
-International, Inc., for the U.S. Department of Energy's National
-Nuclear Security Administration under contract DE-NA-0003525.
+Administration (NNSA).
+
+Sandia National Laboratories is a multimission laboratory managed and
+operated by National Technology and Engineering Solutions of Sandia,
+LLC., a wholly owned subsidiary of Honeywell International, Inc., for
+the U.S. Department of Energy's National Nuclear Security
+Administration under contract DE-NA-0003525.
 

### 2017-10-06 paper.tex
add middle initial
minor rewording of that memory model section

@@
 
-\author{Stephen Olivier}
+\author{Stephen L. Olivier}
 \affiliation{\institution{Center for Computing Research \\ Sandia National Laboratories}}
@@
 specification. We surmise that in practice this hasn't had an effect because
-\texttt{private} variables are often used for input parameters in the 
-progams or for performance rather than their semantic properties.
+\texttt{private} variables are often used for inputs to calculations that are 
+not accessed after the tasks complete, 
+or for performance rather than their semantic properties.
 Also, \texttt{firstprivate} is the default data-sharing attribute for tasks.

### 2017-10-06 paper.tex
fix spelling typo

@@
 represents a small step in that direction and there is still a herculean amount
-of design and implementaton details to be considered. We argue that the benefits vastly outweigh 
+of design and implementation details to be considered. We argue that the benefits vastly outweigh 
 the costs and we hope to see future work explore additional possibilities.

### 2017-10-06 paper.tex
Fixed up paragraph in section 7.1

@@
 generally outperforming or matching the best of existing implementations. For
-benchmarks where the overhead-to-work ratio is high is where the Tapir (and 
-Cilk) implementation shows significant benefit. For example, for the \texttt{Fibonacci}
-benchmark, our implementation runs significantly faster than any of the
-others. While ICC failed for that size input, we did test with smaller inputs
-and found it was similarly outperformed by the Tapir implementation.  \texttt{NQueens}
-and \texttt{FFT} show similar behavior but to lesser degrees. In contrast,
-benchmarks that have a lower overhead to work ratio unsurprisingly differ less
-in performance. For example,
-
-%%% something missing here!
+benchmarks where the overhead-to-work ratio is high is where the Tapir + Cilk
+implementation shows significant benefit. For example, for the
+\texttt{Fibonacci} benchmark, our implementation runs significantly faster than
+any of the others. \texttt{NQueens} and \texttt{FFT} show similar behavior but
+to lesser degrees. In contrast, benchmarks that have a lower overhead to work
+ratio unsurprisingly differ less in performance.
 

### 2017-10-08 paper.tex
fix overlapping text in header
Added uts footnote

@@
 
+\renewcommand{\shortauthors}{G. Stelle et al.}
+
 \begin{abstract}
@@
 For this paper our prototype implementation uses the Cilk runtime as a back end,
-and we run our prototype implementation on the Barcelona OpenMP task suite~\cite{barcelona}. 
-We discuss the kinds of optimizations this enables and how
-the work can be extended to include other parallel constructs and semantics
+and we run our prototype implementation on the Barcelona OpenMP task
+suite~\cite{barcelona}.  We discuss the kinds of optimizations this enables and
+how the work can be extended to include other parallel constructs and semantics
 in OpenMP. We also discuss the possibility of extending this work to other
-programming models, which would help to alleviate some of the problems with fragmentation
-previously discussed.
+programming models, which would help to alleviate some of the problems with
+fragmentation previously discussed.
 
@@
 intended to test performance of OpenMP implementations using both irregular and
-regular tasking. The following list provides abridged descriptions of each
-benchmark taken verbatim from the test suite distribution \cite{barcelona}.
+regular tasking.\footnote{Since original publication, an unbalanced tree search
+benchmark has been added to the benchmark suite. Unfortunately, due to time
+constraints, we weren't able get results for this additional benchmark.} The
+following list provides abridged descriptions of each benchmark taken verbatim
+from the test suite distribution \cite{barcelona}.
 
@@
 
+Another way we can discriminate the cause of performance discrepency is by disabling 
+Tapir optimizations. This allows us to directly measure the impact that Tapir
+optimizations are having on runtime. For \texttt{fibonacci}, for example,
+disabling optimizations results in a roughly ~2x performance loss. This is likely
+
 \section{Discussion} \label{Sec:Discussion}
@@
 implementation of OpenMP's specification. In contrast, existing OpenMP
-implementations sometimes write shared variables to memory on every write access. This is a
-potential for performance discrepancy because of differing 
+implementations sometimes write shared variables to memory on every write
+access. This is a potential for performance discrepancy because of differing
 implementation behavior.
-%Stephen:This topic is not discuessed later, but was touched on before -- moved?
-%and we will return to it when discussing
-%performance on the Barcelona OpenMP task suite. 
-There is of course the fact
-that technically, by copying the value of variables back to the surrounding
-context even for \texttt{private} variables, one isn't following the
-specification. We surmise that in practice this hasn't had an effect because
-\texttt{private} variables are often used for inputs to calculations that are 
-not accessed after the tasks complete, 
-or for performance rather than their semantic properties.
-Also, \texttt{firstprivate} is the default data-sharing attribute for tasks.
+%Stephen:This topic is not discuessed later, but was touched on before --
+%moved?  and we will return to it when discussing performance on the Barcelona
+%OpenMP task suite. 
+There is of course the fact that technically, by copying the value of variables
+back to the surrounding context even for \texttt{private} variables, one isn't
+following the specification. We surmise that in practice this hasn't had an
+effect because \texttt{private} variables are often used for inputs to
+calculations that are not accessed after the tasks complete, or for performance
+rather than their semantic properties.  Also, \texttt{firstprivate} is the
+default data-sharing attribute for tasks.
 
@@
 
-One of the important questions in this work is whether the Tapir
-instructions can be extended to cover the full OpenMP semantics.
+One of the important questions in this work is whether the Tapir instructions
+can be used to implement the full OpenMP semantics.
 %As we have shown, even the full semantics of OpenMP tasks weren't covered.
-Runtime
-calls aside, implementing the full semantics of OpenMP pragmas would be a
-significant undertaking.
+Runtime calls aside, implementing the full semantics of OpenMP pragmas would be
+a significant undertaking.
 
@@
 \texttt{critical} pragmas from the \texttt{Floorplan} benchmark. The
-\texttt{atomic} pragma would likely not be too difficult to translate
-from existing OpenMP codegen. One would need to simply maintain and then
-modify the contents of the original stack pointer to the shared variable. The
+\texttt{atomic} pragma would likely not be too difficult to translate from
+existing OpenMP codegen. One would need to simply maintain and then modify the
+contents of the original stack pointer to the shared variable. The
 \texttt{critical} section would likely be a little more difficult. Still, it

### 2017-10-08 paper.tex
Fixed double words

@@
 mpi}.  From libraries, to language extensions, to standalone languages, to
-combinations of the these, the fragmentation of parallel programming models is a
+combinations of these, the fragmentation of parallel programming models is a
 problem that is only getting worse. This means that when writing tooling,
@@
 There are other issues that had to be overcome to run full OpenMP programs
-using Tapir. Generally, in addition the the \texttt{task} 
-and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP.
-A standard pattern for building task-parallel OpenMP programs is to insert a
+using Tapir. Generally, in addition the \texttt{task} and \texttt{taskwait}
+pragmas, a program must contain pragmas to initialize OpenMP.  A standard
+pattern for building task-parallel OpenMP programs is to insert a
 \texttt{parallel} pragma to start the necessary hardware threads, followed
@@
 no-op shortcut, of course, doesn't follow the OpenMP specification, and should
-be addressed by future work. Section~\ref{Sec:Future} discusses how how the
+be addressed by future work. Section~\ref{Sec:Future} discusses how the
 implementation can be improved.

### 2017-10-08 paper.tex
Fixed up discussion of disabling tapir opts

@@
 only profiling information. A more thorough investigation would require deep
-knowledge of each of the front ends and runtimes. While the importance of
-optimizations of parallel code is clear, we hypothesize that most of the
-disparity present in these results is more a function of runtime implementation
-differences. As justification for this hypothesis, consider the
+knowledge of each of the front ends and runtimes. We hypothesize that much of
+the disparity present in these results is more a function of runtime
+implementation differences. As justification for this hypothesis, consider the
 \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, and the
@@
 
-There are two cases we examine using the profiler. The first is the
-case when there is a large difference in run times between implementations. For 
-these our hypothesis is that the slower implementations
-are for some reason spending more time in the runtime. As mentioned, 
-it is difficult to understand exactly why an implementation is spending more 
-time in the runtime without further investigation, but it is a useful 
-sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
-Indeed, we get roughly the following breakdowns of work-to-overhead ratio
-according to the Linux \texttt{perf} tool.
+There are two cases we examine using the profiler. The first is the case when
+there is a large difference in run times between implementations. For these our
+hypothesis is that the slower implementations are for some reason spending more
+time in the runtime. As mentioned, it is difficult to understand exactly why an
+implementation is spending more time in the runtime without further
+investigation, but it is a useful sanity check nonetheless. For this case, we
+turn to the \texttt{Fib} benchmark.  Indeed, we get roughly the following
+breakdowns of work-to-overhead ratio according to the Linux \texttt{perf} tool.
 \begin{itemize}
@@
 \end{itemize}
-This supports our hypothesis that most of the slowdown incurred is overhead in
+This supports our hypothesis that much of the slowdown incurred is overhead in
 the runtime. For Tapir the overhead is relatively small given the size of the
@@
 optimizations are having on runtime. For \texttt{fibonacci}, for example,
-disabling optimizations results in a roughly ~2x performance loss. This is likely
+disabling optimizations results in a roughly ~2x performance loss. This is
+likely due to Tapir inlining the second task call and moving some memory
+operations into registers. This is strong evidence that Tapir optimizations are
+an important factor in the performance of compiled task-parallel code. Again,
+note that this compounds with runtime performance discrepency to result in the
+measured performance reported in Section~\ref{Sec:Results}.
 

### 2017-10-08 paper.tex
Add MIT NSF grant
Pretty pictures

@@
 \graphicspath{{./figs/}}
+
+\usepackage[x11names]{xcolor}
+\usepackage{fccode}
+\usepackage{cilkbookstyle}
+
+% FCCode macro
+\newcommand{\fccode}[1]{\terminal{\fontsize{7.5}{9}\selectfont #1}}
+\newcommand{\fc}[2]{\PY{#1}{#2}}
 
@@
 \newcommand{\McCormick} {M\raise .40ex\hbox{c}Cormick}
+\newenvironment{tab}[2][\linewidth]
+{\begin{tabular*}{#1}[t]{@{\extracolsep{\fill}}>{\hspace{4pt}}#2}}%
+{\end{tabular*}}
 
 \tikzset{
-  basic block/.style={
-    rectangle,
+  pblock/.style={
+    circle,
     rounded corners,
@@
     % font={\ttfamily\fontsize{8.5}{10.2}\selectfont},
-    font={\ttfamily\fontsize{7.5}{9}\selectfont},
+    font={\ttfamily\fontsize{16}{18}\selectfont},
     text=terminalcolor,
@@
 
-The contributions of this work include
+The contributions of this work include:
 \begin{itemize}
@@
 relative ease of both experimentation and modifications, and extensibility~\cite{lattner2004llvm}.
-Historically, LLVM has only had a sequential semantics: there is no explicit notion
+Historically, LLVM has only had sequential semantics: there is no explicit notion
 of concurrent or parallel processes. This has resulted in existing compilers
@@
 \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l@{}l}
-
 \subfiglabel{a}\\
-\begin{lstlisting}
-void search(int low, int high) {
-  if (low == high) search_base(low);
-  else {
-    #pragma omp task
-    search(low, (low+high)/2);
-    #pragma omp task
-    search((low+high)/2 + 1, high);
-    #pragma omp taskwait
-  } 
-}
-\end{lstlisting}
+\ccodefig{figs/search}
 \\
 \subfiglabel{b}\\
-\begin{lstlisting}
-void search(int low, int high) {
-  if (low == high) search_base(low);
-  else {
-    int mid = (low+high)/2;
-    #pragma omp task
-    search(low, mid);
-    #pragma omp task
-    search(mid + 1, high);
-    #pragma omp taskwait
-  } 
-}
-\end{lstlisting}
+\ccodefig{figs/search2}
 \vspace{0.1ex}
@@
 \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l@{}l}
-\begin{lstlisting}
-void task_1(int* clos) {
-  search(*clos[0], (*clos[0] + *clos[1])/2);
-}
-
-void task_2(int* clos) {
-  search(*clos[0], (*clos[0] + *clos[1])/2);
-}
-
-void search(int low, int high) {
-  if (low == high) search_base(low);
-  else {
-    int* closure_1[] = {&low, &high};
-    omp_run_task(task_1, closure_1);
-    int* closure_2[] = {&low, &high};
-    omp_run_task(task_2, closure_2);
-    omp_taskwait();
-  } 
-}
-\end{lstlisting}
+\ccodefig{figs/search3}
 \vspace{0.1ex}
@@
 \begin{minipage}[T]{0.45\linewidth}
-      \begin{lstlisting}
-int f(int x) {
-  int y = 10;
-  if (x == 0) {
-    y = 12;
-  }
-  return y;
-}
-      \end{lstlisting}
+\ccodefig{figs/llvm}
     \end{minipage}
@@
     \begin{minipage}[T]{0.45\linewidth}
-\begin{lstlisting}[language=llvm]
-define i32 @f(i32 %x) {
-entry:
-  %y = alloca i32, align 4
-  store i32 10, i32* %y, align 4
-  %cmp = icmp eq i32 %x, 0
-  br i1 %cmp, label %if.then, label %if.end
-
-if.then:
-  store i32 12, i32* %y, align 4
-  br label %if.end
-
-if.end:
-  %1 = load i32, i32* %y, align 4
-  ret i32 %1
-}
-\end{lstlisting}
+\llcodefig{figs/llvm}
     \end{minipage}\\
@@
 
-\begin{figure*}
-\begin{center}
-\begin{tabular}{| c c c | c c c | c c c |}
+\begin{figure*}[t]
+  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lll}
+    \subfiglabel{a} & \subfiglabel{b} & \subfiglabel{c} \\
+    \begin{minipage}[T]{0.33\linewidth}
+      \begin{flushleft}
+        \ccodefig{figs/fibcode}
+      \end{flushleft}
+    \end{minipage}
+    &
+    \begin{minipage}[T]{0.33\linewidth}
+    \centering
+        \begin{tikzpicture}[slave]
+          \node [pblock] (A)
+          {A};
+          \node [pblock, below= of A, xshift=-.7cm, yshift=.15cm] (B)
+          {B};
+          \node [pblock, below= of A, xshift=.7cm, yshift=.15cm] (C)
+          {C};
+          \node [pblock, below= of C, xshift=-.7cm,
+          yshift=.15cm] (D)
+          {D};
+          \path [cfedge]
+          (A)  edge node {} (B)
+          (A)  edge node {} (C)
+          (B)  edge node {} (D)
+          (C)  edge node {} (D);
+        \end{tikzpicture}
+    \end{minipage}
+    &
+    \begin{minipage}[T]{0.33\linewidth}
+    \centering
+    { \large
+    \begin{tabular}{ |l||c|c|c|  }
       \hline
- & 
-\begin{lstlisting}
-int fib(int n) 
-{                    // A
-  if (n < 2)
-    return n;
-  else
-  {
-    int x, y;
-    #pragma omp task
-      x = fib(n-1);  // B
-    #pragma omp task
-      y = fib(n-2);  // C
-    #pragma omp taskwait
-    return (x+y);    // D
-  }
-}
-\end{lstlisting}
-& & &
-    \centering{
-   \raisebox{-0.6in}{\includegraphics[scale=0.35]{fib-task-graph.pdf}}
-    }
-& & &
-    \centering{
-     \raisebox{-0.25in}{\includegraphics[scale=0.35]{fib-schedule.pdf}}
-    }
- &  \\
+      \textbf{Time} & $t_0$ & $t_1$ & $t_2$\\
+      \hline
+      \textbf{Thread 0}  & A    & B & D\\
+      \hline
+      \textbf{Thread 1}  &      & C &  \\
       \hline
     \end{tabular}
-\end{center}
+    }
+    \end{minipage}\\
+\addlinespace[0.5ex]
+% ~
+%     \\
+    \bottomrule
+  \end{tabular*}
   \caption{Code, task graph, and schedule of a simple brute force recursive
@@
   \label{fig:fib-graph-to-schedule}
+\vspace{-.4cm}
 \end{figure*}
@@
 \begin{minipage}[T]{0.45\linewidth}
-\begin{lstlisting}
-  #pragma omp task
-    x = fib(n-1);  
-  #pragma omp task
-    y = fib(n-2);  
-  #pragma omp taskwait
-\end{lstlisting}
+\ccodefig{figs/fibomp}
 \end{minipage}
@@
 \begin{minipage}[T]{0.45\linewidth}
-\begin{lstlisting}[language=llvm]
-if.end:                                 
-  detach label %det.achd, label %det.cont
-
-det.achd:                                 
-  %2 = load i32, i32* %n.addr, align 4
-  %sub = sub nsw i32 %2, 1
-  %call = call i32 @fib(i32 %sub)
-  store i32 %call, i32* %x, align 4
-  reattach label %det.cont
-
-det.cont:                                   
-  detach label %det.achd1, label %det.cont4
-
-det.achd1:                                
-  %3 = load i32, i32* %n.addr, align 4
-  %sub2 = sub nsw i32 %3, 2
-  %call3 = call i32 @fib(i32 %sub2)
-  store i32 %call3, i32* %y, align 4
-  reattach label %det.cont4
-
-det.cont4:                                  
-  sync label %sync.continue
-\end{lstlisting}
+\llcodefig{figs/fibtapir}
 \end{minipage}\\
@@
 for the Department of Energy's National Nuclear Security
-Administration (NNSA).
+Administration (NNSA). Additionally, this research was supported in
+part by NSF Grants 1314547 and 1533644 and an MIT SuperUROP.
 

### 2017-10-08 figs/fibcode.c

@@ -1,0 +1,15 @@
+int fib(int n)
+{                    // A
+  if (n < 2)
+    return n;
+  else
+  {
+    int x, y;
+    #pragma omp task
+      x = fib(n-1);  // B
+    #pragma omp task
+      y = fib(n-2);  // C
+    #pragma omp taskwait
+    return (x+y);    // D
+  }
+}

### 2017-10-08 figs/llvm.c

@@ -1,0 +1,8 @@
+int f(int x) {
+  int y = 10;
+  if (x == 0) {
+    y = 12;
+  }
+  return y;
+}
+

### 2017-10-08 figs/search.c

@@ -1,0 +1,10 @@
+void search(int low, int high) {
+  if (low == high) search_base(low);
+  else {
+    #pragma omp task
+    search(low, (low+high)/2);
+    #pragma omp task
+    search((low+high)/2 + 1, high);
+    #pragma omp taskwait
+  }
+}

### 2017-10-08 figs/search2.c

@@ -1,0 +1,11 @@
+void search(int low, int high) {
+  if (low == high) search_base(low);
+  else {
+    int mid = (low+high)/2;
+    #pragma omp task
+    search(low, mid);
+    #pragma omp task
+    search(mid + 1, high);
+    #pragma omp taskwait
+  }
+}

### 2017-10-08 tapir.py

@@ -1,0 +1,112 @@
+import re
+
+from pygments.lexer import RegexLexer, include, bygroups, using, words, \
+    DelegatingLexer
+from pygments.lexers.c_cpp import CppLexer, CLexer
+from pygments.lexers.d import DLexer
+from pygments.token import Text, Name, Number, String, Comment, Punctuation, \
+    Other, Keyword, Operator
+
+class CustomLexer(RegexLexer):
+    """
+    For Tapir/LLVM assembly code.
+    """
+    name = 'Tapir'
+    aliases = ['tapir']
+    filenames = ['*.ll']
+    mimetypes = ['text/x-llvm']
+
+    #: optional Comment or Whitespace
+    string = r'"[^"]*?"'
+    identifier = r'([-a-zA-Z$._][\w\-$.]*|' + string + ')'
+
+    tokens = {
+        'root': [
+            include('whitespace'),
+
+            # Before keywords, because keywords are valid label names :(...
+            (identifier + '\s*:', Name.Label),
+
+            include('keyword'),
+
+            (r'%' + identifier, Name.Variable),
+            (r'@' + identifier, Name.Variable.Global),
+            (r'%\d+', Name.Variable.Anonymous),
+            (r'@\d+', Name.Variable.Global),
+            (r'#\d+', Name.Variable.Global),
+            (r'!' + identifier, Name.Variable),
+            (r'!\d+', Name.Variable.Anonymous),
+            (r'c?' + string, String),
+
+            (r'0[xX][a-fA-F0-9]+', Number),
+            (r'-?\d+(?:[.]\d+)?(?:[eE][-+]?\d+(?:[.]\d+)?)?', Number),
+
+            (r'[=<>{}\[\]()*.,!]|x\b', Punctuation)
+        ],
+        'whitespace': [
+            (r'(\n|\s)+', Text),
+            (r';.*?\n', Comment)
+        ],
+        'keyword': [
+            # Regular keywords
+            (words((
+                'begin', 'end', 'true', 'false', 'declare', 'define', 'global',
+                'constant', 'private', 'linker_private', 'internal',
+                'available_externally', 'linkonce', 'linkonce_odr', 'weak',
+                'weak_odr', 'appending', 'dllimport', 'dllexport', 'common',
[…]

### 2017-10-08 paper.tex
Title changes?
Kei is done

--- 2017-10-08 paper.tex
@@
 interoperate either at the level of compile-time analysis or at run-time
-execution. In addition, many of these programming models are implemented as
+execution. Moreover, many of these programming models are implemented as
 runtime libraries with only primitive compiler support. As a consequence, most are
@@
 fragmentation among models presents a significant challenge for compiler
-writers, which must choose a single model, or multiple intermediate forms, for
+writers who must choose a single model, or multiple intermediate forms, for
 analysis and optimization. An ideal solution would be a 
@@
 intermediate representations \textbf{(IRs)} that allow such transformations
-to be written once with relative ease and apply to a variety of source
+to be written once with relative ease and applied to a variety of source
 front-end languages such as C or C++.  Historically, the IR of mainstream
@@
 benchmark performance and test accuracy they wrote a front end which translated
-programs written with Cilk, a parallel programming model for C/C++, into the
+programs written with Cilk, a parallel programming model for C/C++, into their
 Tapir IR. They also implemented a back end to \textbf{lower} Tapir programs to
@@
 
-The remainder of the paper is organized as follows.  In Section~\ref{Sec:Background} 
-we give background and motivation for the presented work. Following that, we describe OpenMP
-tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
-discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
-Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
-evaluation setup, including which implementations we compare to and how. We discuss
-the results of evaluation in Section~\ref{Sec:Results}, including possible
-explanations for discrepancies. We discuss the implications of the results and the
-significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
-we cover some of the many ways the work could be extended in
-Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.
+The remainder of the paper is organized as follows.  In
+Section~\ref{Sec:Background} we give background and motivation for the
+presented work. Following that, we describe Tapir in Section~\ref{Sec:Tapir},
+and OpenMP tasking in Section~\ref{Sec:OpenMP}.  We then discuss the
+implementation, and how we compile OpenMP task constructs to Tapir IR in
+Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe
+the evaluation setup, including which implementations we compare to and
+how. We discuss the results of evaluation in Section~\ref{Sec:Results},
+including possible explanations for discrepancies. We discuss the implications
+of the results and the significance of this work in a large context
+in Section~\ref{Sec:Discussion}. Finally, we cover some of the many ways the work
+could be extended in Section~\ref{Sec:Future}, and conclude in
+Section~\ref{Sec:Conclusion}.
 
@@
 of reasoning about the corresponding semantics in a uniform fashion. The
-addition of parallel constructs were recently explored by Schardl et
-al.~\cite{tapir}. By adding three first-class instructions to LLVM, Tapir
+addition of parallel constructs was recently explored by Schardl et
+al.
+%~\cite{tapir}.
+By adding three first-class instructions to LLVM, Tapir
 enables analyses and optimizations of parallel constructs. We briefly describe
@@
 % description see \cite{tapir}.
+% KD removed it again.  Also, citatations are meta-syntactic.
 
@@
 problem that is only getting worse. This means that when writing tooling,
-optimizations, or analyses, one much choose which model to target as well as
+optimizations, or analyses, one must choose which models to target as well as
 understand points of contention and interoperability between them. A common
@@
 the possibility of multiple programming models coexisting peacefully. We will
-return to this consideration later in Section~\ref{Sec:Discussion} when
+return to this consideration later in Section~\ref{Sec:Future} when
 discussing future work.
@@
 \section{Tapir} \label{Sec:Tapir}
-Tapir is an extension to LLVM proposed by Schardl, Moses, and Leiserson
+Tapir is an extension to LLVM
+%proposed by Schardl, Moses, and Leiserson
 seeking to resolve issues in optimizing parallel code. Prior to
-the introduction of Tapir compilers would represent parallel programs
+the introduction of Tapir, compilers would represent parallel programs
 by directly translating their syntax to opaque runtime calls. This allowed
@@
 \caption[Simplified compilation of the unoptimized search code from \figref{search}.]
-{Simplified compilation of the unoptimized search code from~\figref{search}.  
-\subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own 
+{Simplified compilation of the unoptimized search code from~\figref{search}.\subfigcap{a}
+          %~The function \code{search}.
+The parallel tasks are moved into their own 
 functions and accompanying closures. These functions are then passed to OpenMP runtime 
-calls. This system obfuscates the program, making it infeasible for an optimizer to 
+calls. This obfuscates the program, making it infeasible for an optimizer to 
 reason about what is happening.}
@@
 analysis and optimizations to work on parallel programs without significant modifications.
-To properly describe Tapir requires a brief introduction to key aspects of LLVM.
+To properly describe Tapir requires a brief introduction to key aspects of LLVM\@.
 LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
@@
 
-Recursive task creations and synchronizations using the constructs result in 
+Recursive task creation and synchronization using the constructs result in 
 an implicit directed acyclic graph (DAG) that allows both reasoning about and 
@@
 shows some example code, a view of the task DAG, and a simplified execution 
-schedule mapping the tasks to a team of two threads.  In the example, the Nth 
+schedule mapping the tasks to a team of two threads.  In the example, the $n$th 
 Fibonacci number is calculated by recursively generating tasks to calculate 
-the (N-1)th and (N-2)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
+the ($n-1$)th and ($n-2$)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
 the child tasks have completed before their answers are combined to yield the 
@@
 marginally increase performance by reducing the number of memory copies for
-variables declared \texttt{private} in OpenMP code.  The fact that behaviour
+variables declared \texttt{private} in OpenMP code.  The fact that behavior
 wasn't changed significantly is worth revisiting, and we do so in
@@
 variables such as \texttt{OMP\_NUM\_THREADS} are ignored and replaced by
-\texttt{CILK\_NWORKERS}.  We discuss some of these issues below, and return to
+\texttt{CILK\_NWORKERS}.  We discuss some of these issues following and return to
 address others in the discussion of future work.
@@
 detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
-Finally, the continuation for the second call immediately calls sync. This is
+Finally, the continuation for the second call immediately calls \texttt{sync}. This is
 a nice example of code that will be optimized by Tapir. The LLVM code
@@
 \begin{itemize}
-\item \texttt{Alignment}: aligns all protein sequences from  an  input
+\item \texttt{Alignment} aligns all protein sequences from  an  input
 file  against  every  other  sequence  using  the Myers and Miller 
@@
 pair of them.
-\item \texttt{FFT}: computes the one-dimensional Fast Fourier Transform
+\item \texttt{FFT} computes the one-dimensional Fast Fourier Transform
 of a vector of $n$ complex values using the Cooley-Tukey 
@@
 divisions multiple tasks are generated.
-\item \texttt{Fibonacci}: computes the $n$th Fibonacci number using a  recursive
-parallelization. While  not  representative  of  an efficient  fibonacci
+\item \texttt{Fibonacci} computes the $n$th Fibonacci number using a  recursive
+parallelization. While  not  representative  of  an efficient  Fibonacci
 computation  it  is  still  useful  because  it  is a simple test case of a
 deep tree composed of very fine grain tasks.  
-\item \texttt{Floorplan}: kernel computes the optimal floorplan distribution
+\item \texttt{Floorplan} kernel computes the optimal floorplan distribution
 of a number of cells. The algorithm gets an input file with  cells'
@@
 introduced in the code to maintain the parent state alive.
-\item \texttt{Health}: simulates the Columbian Health Care System. It
+\item \texttt{Health} simulates the Columbian Health Care System. It
 uses multilevel lists where each element in the structure  represents  a
@@
 synchronization occurs. 
-\item \texttt{NQueens}: computes  all  solutions  of  the n-queens
+\item \texttt{NQueens} computes  all  solutions  of  the n-queens
 problem, whose objective is to find a placement for $n$ queens on an $n \;
@@
 of the solution.
-\item \texttt{Sort}: sorts a random permutation of $n$ 32-bit numbers with  a
+\item \texttt{Sort} sorts a random permutation of $n$ 32-bit numbers with  a
 fast  parallel  sorting  variation of  the  ordinary mergesort.  First, it
@@
 elements).
-\item \texttt{SparseLU}: computes an LU matrix factorization over
+\item \texttt{SparseLU} computes an LU matrix factorization over
 sparse matrices. A first level matrix is composed by pointers to  small
@@
 matrix that is not empty.
-\item \texttt{Strassen}: algorithm  uses  hierarchical  decomposition of a
+\item \texttt{Strassen} uses  hierarchical  decomposition of a
 matrix for multiplication of large dense matrices. Decomposition is done by
@@
 We set a time-out of 10 minutes, as at least one implementation was always
-finishing within 10 seconds, anything running more than 60 times slower becomes
+finishing within 10 seconds so anything running more than 60 times slower becomes
 irrelevant.  Because of time constraints correctness checks were not performed on
-every run so it is possible some non-determinism was missed.
+every run so it is possible that some non-determinism was missed.
 
@@
 implementation, and how its performance compared to the existing
-implementations listed in Section~\ref{Sec:Evaluation}. See
-Figures~\ref{fig:results} and \ref{fig:results2} for a simple visualization of
-the performance. Each graph represents a single benchmark from the previously
+implementations listed in Section~\ref{Sec:Evaluation}. 
+Figures~\ref{fig:results} and \ref{fig:results2} give simple visualizations of
+performance. Each graph represents a single benchmark from the previously
 enumerated Barcelona benchmarks, and each bar represents one of the
@@
 
-As mentioned earlier on \texttt{Floorplan} our implementation returned correct
+As mentioned earlier regarding \texttt{Floorplan} our implementation returned correct
 results nondeterministically because of not having an implementation of the OpenMP
 \texttt{critical} and \texttt{atomic} pragmas. While ICC and Clang finished in
-roughly 1-2 seconds, the Tapir implementation was finishing in 0.02 seconds
-and computing the correct result roughly half of the time. This raises interesting
+roughly 1-2 seconds, the Tapir implementation was finished in 0.02 seconds
+and computed the correct result roughly half of the time. This raises interesting
 questions on the cost/benefit relation of the missing OpenMP pragmas.
@@
 better and where each of the implementations is spending the bulk of the
-execution time. Our primary tool for this task is profiling. It's worth noting
-that lots of information is difficult to infer from profiling, such as sources
+execution time. Our primary tool for this task is profiling, acknowledging
+that many kinds information are difficult to infer from profiling, such as sources
 of contention, reasons for memory locality, etc.
 
-It is worth noting again that the scope of the work that this paper is
-responsible for is only the front end. All performance gains attributable to
+It is worth noting again that the scope of the work that this paper reports
+is only the front end. All performance gains attributable to
 the existing Tapir back end and Cilk runtime represent prior work.  We can
-break reasons for performance discrepancies into a few categories:
+separate reasons for performance discrepancies into a few categories:
 \begin{itemize}
@@
 investigation, but it is a useful sanity check nonetheless. For this case, we
-turn to the \texttt{Fib} benchmark.  Indeed, we get roughly the following
+turn to the \texttt{Fibonacci} benchmark.  We get roughly the following
 breakdowns of work-to-overhead ratio according to the Linux \texttt{perf} tool.
 \begin{itemize}
-\item \texttt{tapir: ~30\% runtime overhead, ~50\% work, ~20\% other}
-\item \texttt{clang: ~85\% runtime overhead, ~5\% work, ~10\% other}
-\item \texttt{icc: ~85\% runtime overhead, ~5\% work, ~10\% other}
-\item \texttt{gcc: ~100\% runtime overhead, ~0\% work, ~0\% other}
+\item \texttt{tapir}: ~30\% runtime overhead, ~50\% work, ~20\% other
+\item \texttt{clang}: ~85\% runtime overhead, ~5\% work, ~10\% other
+\item \texttt{icc}: ~85\% runtime overhead, ~5\% work, ~10\% other
+\item \texttt{gcc}: ~100\% runtime overhead, ~0\% work, ~0\% other
 \end{itemize}
@@
 work chunks, for \texttt{clang} and \texttt{icc} it is significantly larger,
-and for \texttt{gcc} it is overwhelming. It's worth noting that \texttt{clang} 
+and for \texttt{gcc} it is overwhelming. \texttt{clang} 
 and \texttt{icc} have effectively the same runtime, so the fact that their 
@@
 Another way we can discriminate the cause of performance discrepency is by disabling 
-Tapir optimizations. This allows us to directly measure the impact that Tapir
+Tapir optimizations. This allows us to directly measure the effect that Tapir
 optimizations are having on runtime. For \texttt{fibonacci}, for example,
@@
 an important factor in the performance of compiled task-parallel code. Again,
-note that this compounds with runtime performance discrepency to result in the
+note that this compounds with runtime performance differences to give the
 measured performance reported in Section~\ref{Sec:Results}.
@@
 for Cilk and surmised that it could be an effective IR for a large subset of
-  OpenMP~\cite{tapir}. This work has taken the first step in that direction.
+  OpenMP. This work has taken the first step in that direction.
 
 One interesting aspect of this work is that it compiles a language extension
-intended for one runtime (OpenMP) to use another (the Cilk runtime). While in
+intended for one runtime (OpenMP) to use another (Cilk). While in
 many ways this is a weakness of this work, it does open the possibility of
@@
 of semantics. In particular, \emph{OpenMP specifies a relaxed memory model
-for shared variables}, in which writes by one thread need only be visible at
+for shared variables} in which writes by one thread need only be visible at
 to other threads at explicit or implicit flushes. 
@@
 \texttt{atomic} pragma would likely not be too difficult to translate from
-existing OpenMP codegen. One would need to simply maintain and then modify the
+existing OpenMP codegen: one would need to simply maintain and then modify the
 contents of the original stack pointer to the shared variable. The
@@
 compiler cannot reason about code using those features, and many of the
-advantages we've discussed become moot, e.g., backend compatibility, generic 
-optimization, etc. Features like the OpenMP \texttt{atomic} pragma don't 
+advantages we've discussed, e.g., backend compatibility, generic 
+optimization, etc., become moot, Features like the OpenMP \texttt{atomic} pragma don't 
 require any changes to control flow, and therefore are likely to work well 
-with Tapir analyses and optimizations. On the other hand features like OpenMP's
+with Tapir analyses and optimizations. On the other hand, features like OpenMP's
 \texttt{critical} section \emph{do} require changes to control flow that seem
@@
 simple, concise set of parallelism constructs, something that metadata and
-intrinsics approaches can easily avoid.
+intrinsics approaches can promptly defeat.
 

### 2017-10-08 paper.tex
WTF?  almost every change I made resulted in a conflict

@@
 
-\title{OpenMPIR}
-\subtitle{Implementing OpenMP tasks with Tapir}
+\title{OpenMP+Tapir: A Case Study of Using Tapir as a Common Parallel Intermediate Representation}
 
@@
 Tapir is a compelling solution to this problem because it allows existing
-analysis and optimizations to work on parallel programs without significant modifications.
-To properly describe Tapir requires a brief introduction to key aspects of LLVM\@.
-LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
-sequences of instructions. Blocks always end with terminator instructions that
-define the edges of the control flow graph. For example, \figref{CFG} shows a simple
+analysis and optimizations to work on parallel programs without significant
+modifications.  To properly describe Tapir requires a brief introduction to
+key aspects of LLVM\@.  LLVM represents expression/evaluation order through a
+control-flow graph (CFG) of blocks that contain sequences of
+instructions. Blocks always end with terminator instructions that define the
+edges of the control flow graph. For example, \figref{CFG} shows a simple
 function and its corresponding LLVM code.
@@
 \item \texttt{NQueens} computes  all  solutions  of  the n-queens
-problem, whose objective is to find a placement for $n$ queens on an $n \;
-\times \; n$ chessboard such that none of the queens attack any other. It uses
+problem, whose objective is to find a placement for $n$ queens on an $n
+\times n$ chessboard such that none of the queens attack any other. It uses
 a backtracking search algorithm with pruning. A task is created for each step
@@
 \section*{Acknowledgments}
-
 Many thanks to Kei Davis for his excellent editing.

### 2018-04-15 plot.py
Added presentation

@@ -2,6 +2,8 @@ import numpy as np
 import matplotlib.pyplot as plt
 import glob
 
+
+import csv
 pdat = {}
 pdat['alignment'] = {'title' : "Alignment"}
 pdat['fft'] = {'title' : "FFT"}
@@ -13,7 +15,7 @@ pdat['sort'] = {'title' : "Sort"}
 pdat['sparselu'] = {'title' : "SparseLU"}
 pdat['strassen'] = {'title' : "Strassen"}
 
-types = ['tapir', 'icc', 'clang', 'gcc']
+types = ['tapir', 'icc', 'clang', 'gcc', 'tapir-omp']
 
 for p in pdat:
     for t in types:
@@ -24,13 +26,19 @@ for p in pdat:
                          for f in fs]
         print(np.mean(pdat[p][t]))
 
-width=0.75
+with open("means.csv", 'w') as f:
+    wr = csv.writer(f)
+    wr.writerow(['compiler'] + list(pdat.keys()))
+    for t in types:
+        wr.writerow([t] + [np.mean(pdat[p][t]) for p in pdat])
 
+width=0.75
 colors=['#2D8632', '#236467', '#AA6D39', '#AA3C39']
 
 for p in ['alignment', 'fft', 'fib', 'nqueens', 'sort', 'sparselu', 'strassen']:
     fig, ax = plt.subplots()
     means = [np.mean(pdat[p][t]) for t in types]
+    valid_types = [t for (t,m) in zip(types, means) if not np.isnan(m)]
     means = [m for m in means if not np.isnan(m)]
     stds = [np.std(pdat[p][t]) for t in types]
     stds = [m for m in stds if not np.isnan(m)]
@@ -38,9 +46,9 @@ for p in ['alignment', 'fft', 'fib', 'nqueens', 'sort', 'sparselu', 'strassen']:
     ax.bar(ind, means, width, yerr=stds, color=colors)
     ax.set_ylabel("Time (Seconds, lower is better)")
     ax.set_xticks(ind)
-    ax.set_xticklabels(types)
+    ax.set_xticklabels(valid_types)
     ax.set_title(pdat[p]['title'])
-    plt.savefig('figs/%s.pdf' %(p))
+    plt.savefig('figs/%s.png' %(p))
 
 for p in ['health', 'floorplan']: 
     fig = plt.figure()
@@ -49,6 +57,7 @@ for p in ['health', 'floorplan']:
     ax2 = fig.add_subplot(212, sharex=ax)
 
[…]
